### 2015-02-01 manuscript.Rmd
rename as manuscript.Rmd

---
title: "An Introduction to Rocker: Docker Containers for R"
author:
  - name: Carl Boettiger
    affiliation: UC Santa Cruz
    address:
    - line 1
    - line 2
    email:  cboettig@ropensci.org
  - name: Dirk Eddlebuettel
    affiliation: Debian, ...
    address:
    - line 1
    - line 2
    email:  author2@work
abstract: >
  An abstract of less than 150 words.
preamble: >
  % Any extra latex you need in the preamble
output: rticles::rjournal_article
---

## Introduction

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

## An overview of Docker

## Rocker organization and workflow

Docker Hub, automated builds, linked repositories

## Rocker base images

Image dependency graph

## The r-base image

The official, verified Docker image for R.

example use.

Link to further details in Rocker wiki

## The r-devel image

example use

## The rstudio image

example use

screenshot

Custom configurations

Link to further details in Rocker wiki

## Additional images

## Summary

This file is only a basic article template. For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](http://journal.r-project.org/latex/RJauthorguide.pdf).

\bibliography{RJreferences}

### 2015-02-01 manuscript.Rmd
header fixes

@@
     email:  cboettig@ropensci.org
-  - name: Dirk Eddlebuettel
-    affiliation: Debian, ...
+  - name: Dirk Eddelbuettel
+    affiliation: Debian and R Projects
     address:
@@
     - line 2
-    email:  author2@work
+    email:  edd@debian.org
 abstract: >

### 2016-11-20 manuscript.Rmd
outline notes

--- 2015-02-01 manuscript.Rmd
@@
   - name: Carl Boettiger
-    affiliation: UC Santa Cruz
+    affiliation: UC Berkeley 
     address:
@@
 
+- Community-optimized images
+
+- Portability & cloud computing
+
+- Reproducibility
+
+- Convenience
+
+  - Installation of certain packages can be challenging and non-intuitive, particularly when those packages require specific external dependencies. 
+  - Many users are reluctant to upgrade their suite of installed packages.
+Using the Rocker images, users can try out packages and run code without having to upgrade or otherwise alter their own system's R configuration, which may break existing work
+
+- Instruction
+
 ## An overview of Docker
+
+\citep{Boettiger2015}
 
@@
 
+UBSAN, Clang
+
 example use
 
-## The rstudio image
+## rocker-versioned stack
 

### 2016-12-14 manuscript.Rmd
WIP on intro

--- 2015-02-01 manuscript.Rmd
@@
   - name: Carl Boettiger
-    affiliation: UC Santa Cruz
+    affiliation: UC Berkeley 
     address:
-    - line 1
-    - line 2
-    email:  cboettig@ropensci.org
+    - ESPM Department, University of California,
+    - 130 Mulford Hall Berkeley, CA 94720-3114, USA
+    email:  cboettig@berkeley.edu
   - name: Dirk Eddlebuettel
@@
     - line 2
-    email:  author2@work
+    email:  edd@debian.org
 abstract: >
-  An abstract of less than 150 words.
+  We describe the Rocker project, which provides a widely used suite of Docker images
+  with customized R environments for particular tasks. We discuss how this suite is
+  organized and how these tools can increase portability, scaling, reproducibility,
+  and convenience of R users and developers.
 preamble: >
@@
 
-Introductory section which may include references in parentheses
-\citep{R}, or cite a reference such as \citet{R} in the text.
+The Rocker project was launched in October 2014 as a collaboration between CB & DE to provide high-quality
+Docker images containing the R environment \citep{edd2014}. Since that time, the project has seen both 
+considerable uptake in the community and substantial development and evolution. Here we seek to document
+the project's objectives and uses.
 
-## An overview of Docker
+What is Docker?
+
+Docker is a popular open source tool to create, distribute, deploy, and run software applications using
+*containers*.  Containers provide a virtual environment requiring everything an application needs to run:
+code, runtime, system tools, system runtime.  Docker containers are lightweight (sharing the operating system kernel,
+starting instantly, using a layered filesystem to minimize disk footprint and download time), built on open standards
+that run on all major platforms (Linux, Mac, Windows) and provide an added layer of security by running an
+application in an isolated environment \citep{what-docker}.
+
+What is a Rocker image?
+
+There are several different but complementary reasons for creating the containerized R environments provided by
+the Rocker project, including convenience, community input, portability and scaling, reproducibility, and instruction. 
+We describe these in turn before presenting the commands required to deploy these environments in different use cases.
+
+### Portability & cloud computing
+
+One of the most popular uses of the Rocker containers is to provide a fast and reliable mechanism to deploy an R envionment on a remote server, such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's Jetstream servers, or private or institutional server hardware.
+Being able to test code in a predictable, pre-configured R environment on a local machine and then run the same code in an identical environment on a remote server (e.g.
+for access to greater RAM, more processors, or merely to free up the local machine from a long-running computation) is essential for low-friction scaling of analysis.
+Without such contanerization, getting code to run appropriately in a remote environment can be a major undertaking, requiring both time and knowledge many would-be users may not have.
+
+The following code installs Docker on almost any Linux-based server and launches a Rocker container providing the RStudio-server environment over a web interface.
+
+```bash
+wget -qO- https://get.docker.com/ | sh
+sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+```
+
+After running these two lines, the user may log into the server merely by pasting its IP address or DNS name into a browser and entering their password for user `rstudio`.
+This provides the user with a familiar, interactive environment running on a remote machine while requiring a minimum of expertise (with many cloud providers, these two lines of code can be run in a startup script without ever logging into the server over ssh).
+Some caveats and additional configuration options are discussed in the example section below.
+The user can select a Rocker container which contains the required software pre-installed, such as `rocker/tidyverse`, which provides the popular `tidyverse` suite of R packages \citep(tidyverse), or even use their own containers which can extend the Rocker containers, as discussed below.
+This workflow saves considerable time relative to installing each package from scratch, which can take over an hour to reconstruct the environment provided by some rocker containers.
+As many commericial cloud providers charge by the hour, a fast and simple mechanism to deploy the desired environment in the cloud is crucial to making this approach practical for more exploratory analyses.
+
+Unlike virtual machine images, multiple such Docker containers can be run on the same server in the same way (though unique TCP ports must be assigned to each).
+More importantly, it is much easier to test software locally using the identical Docker container environment before deploying on a remote server than it is to run a virtual machine such as an Amazon Machine Image (AMI) locally, or to reproduce the runtime environment identically on local and server systems.
+
+Of course Rocker images may be used for quickly deploying R code on remote servers without the use of the RStudio interface, running R in terminal mode over an `ssh` connection or simply in batch mode.
+
+Another illustration of this portability is with regard to Continuous Integration servers such as Travis <https://travis.com> or Circle-CI <https://circle-ci.com>.
+While support for the R language on Travis has greatly improved recently (\citep{r-travis}), very customized environments can still be difficult to run (particularly for testing scripts not laid out in the R package structure).
+
+### Convenience
+
+Installation of certain R packages can be challenging and non-intuitive, particularly when those packages require specific external dependencies.
+By running R through a Rocker image, users can deploy an R environment where their desired packages and libraries have already been installed and configured.
+As we discuss below, this configuration benefits from the input of a large community of users, helping to ensure better and more reliable performance.
+For instance, one example of such a configuration is in building and testing R packages using compiler sanatizers \citep{edd_sanitizers}, which requires R itself be re-compiled with specialized settings not avaialable or familiar to most R users.
+Using the Rocker images `rocker/r-devel-san` and `rocker/r-devel-ubsan-clang` provides a fast and simple way for users to test
+
+Many users are reluctant to upgrade their suite of installed packages, which may break their existing code or even their R environment if the installation goes poorly.
+However, upgrading packages and/or the R environment is often necessary to run analysis of a colleague or access more recent methods.
+Using Rocker, users can run the latest or specific earlier versions of R and R packages in an isolated environment without impacting their local system.
+
+### Extensible and Community-optimized images
+
+Rocker images benefit from widespread use, testing and contributions from many members of the R community.
+
+### Reproducibility
+
+\citep{Boettiger2015}
+
+### Instruction
 
@@
 
-Docker Hub, automated builds, linked repositories
+The Rocker Project consists of a suite of images built automatically by and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.
+Source Dockerfiles, supporting scripts and documentation are hosted on GitHub.
+The issues tracker and pull requests are used to provide community input, disucssions and contributions to these images.
+The rocker project wiki <https://github.com/rocker-org/rocker/wiki> provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
+
+- Table of main repos, links to Docker-Hub & GitHub addresses
+
+- list/table of topics on the wiki
 
@@
 
-The official, verified Docker image for R.
+The r-base image is the official, verified Docker image for R.
+Written and maintained by the Rocker developers on behalf of the Docker project, this image is built and hosted directly by Docker's own developers, who also approve any changes or updates to the image.
 
@@
 
-## The r-devel image
+## The r-devel images
 
-example use
+These images provide both the current and under-development versions of R, including images with particular compiler option configurations (UBSAN, clang).
+Package authors submitting R packages to the Central R Archive Network (CRAN) are expected to test their packages on both current and development versions of R prior to submission.
 
-## The rstudio image
+## rocker-versioned stack
 
@@
 
+## Metrics of use
+
 ## Summary

### 2016-12-14 manuscript.Rmd

@@
     email:  cboettig@berkeley.edu
-  - name: Dirk Eddlebuettel
-    affiliation: Debian, ...
+  - name: Dirk Eddelbuettel
+    affiliation: Debian and R Projects
     address:
@@
 
-```bash
+```
 wget -qO- https://get.docker.com/ | sh

### 2017-06-17 manuscript.Rmd
WIP

@@
 
-What is Docker?
+### What is Docker?
 
@@
 
-What is a Rocker image?
+### Images in the Rocker Project
+
+The Rocker Project aims to provide a small core if Docker images that serve both as convenient base images for a wide range of possible applications while still providing a 'batteries included' approach of images that can be used out of the box.
+The challenges of balancing diverse needs driven by very different use cases (discussed in more detail below) against the overarching goals of creating images that are not sufficiently light-weight, easy to use and easy to maintain is a difficult art.
+The implementation in both individual Rocker images and image stacks can never perfect that balance for everyone, but today reflects the considerable community input and testing over past few years.
+
+<!--What is a Dockerfile? How do they stack? -->
+
+<!-- some overview on using vs extending images (i.e. writing derivative Dockerfiles). -->
+
+The Rocker Project can be largely divided into two stacks which address different needs.
+The first stack is based on `debian:testing`, the second, more recently introduced stack is based only on `debian` stable releases.
+These stacks have different aims and thus provide different images, as shown in Tables 1 & 2.
+
+The Debian platform provides a small base image, the well known `apt` package management system and rich ecosystem of software libraries, making it the base image of choice for Docker images, including many of the "official" images maintained by Docker's own development team.
+The Debian platform is also perhaps the best supported Linux platform within the R community, including an active `r-sig-debian` listserve.
+Unfortunately, the relatively long period between stable Debian releases (roughly two years recently) means that software in the `debian:stable` releases can lag significantly behind current releases.
+This issue is the primary motivation behind the `debian:testing` stack.
+
+The Debian release cycle follows three stages: `unstable`, `testing`, and `stable`.
+
+The `debian:testing` stack aims to makes the most efficient use of upstream builds -- the precompiled `.deb` binaries provided by the Debian repositories.
+It is both quicker and easier to install software from binaries, since the package manager (`apt`) manages the necessary dependencies and it bypasses the time-consuming process of compiling from source.
+Basing this stack on `debian:testing` means that much more recent versions of commonly used libraries and compilers are available as binaries than would be found in a Debian stable release.
+In order to provide access to the most recent avialble binaries, this stack uses apt-pinning [@apt-pinning] to allow the `apt` package manager to also install binaries from `debian:unstable`, which represents the most recent, bleeding edge of packages built for Debian when necessary.
+For instance, the `r-base` image provided by Rocker installs the most recent version of R as a binary from Debian `unstable`.
+Similarly, recent versions of many popular R packages can also be installed through the package manager, e.g.
+`apt-get install r-cran-XML`.
+This can be particularly helpful for packages with external system dependencies (such as `libxml2-dev` in this example) which cannot be installed from the R console.
+
+As the names `testing` and `unstable` imply, this approach is not without a downside.
+The particular version of any given package can change as packages move from `unstable` into `testing` and newer versions are sent to `unstable` during the normal course of Debian development.
+This can occassionally break an previously working installation command in a Dockerfile until the maintainer redirects the package manager to install a the package from the `unstable` sources that could previously be installed from `testing`, or vice versa (using the `-t` option in `apt`, see Box 1 on apt-pinning.)
+
+The `debian:testing` stack thus offers the following advantages:
+
+1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
+2. These Dockerfiles always install the most recent available software
+
+and these down-sides:
+
+1. These Dockerfiles require occassional minor maintenance to ensure successful builds
+2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
+3. The use of apt pinning may be unfamiliar to some users.
+
+### debian-stable stack
+
+This stack emphasizes stability and reproducibility of the Docker build.
 
@@
 Some caveats and additional configuration options are discussed in the example section below.
-The user can select a Rocker container which contains the required software pre-installed, such as `rocker/tidyverse`, which provides the popular `tidyverse` suite of R packages \citep(tidyverse), or even use their own containers which can extend the Rocker containers, as discussed below.
+The user can select a Rocker container which contains the required software pre-installed, such as `rocker/tidyverse`, which provides the popular `tidyverse` suite of R packages \citep{tidyverse}, or even use their own containers which can extend the Rocker containers, as discussed below.
 This workflow saves considerable time relative to installing each package from scratch, which can take over an hour to reconstruct the environment provided by some rocker containers.
@@
 
-Unlike virtual machine images, multiple such Docker containers can be run on the same server in the same way (though unique TCP ports must be assigned to each).
+<!--
+FIXME describe other modes of running containers, e.g. R terminal
+
+-->
+
+Unlike virtual machine images, multiple such Docker containers can be run on the same server in the same way (though unique TCP ports must be assigned to each if they are accessed through the RStudio web interface).
 More importantly, it is much easier to test software locally using the identical Docker container environment before deploying on a remote server than it is to run a virtual machine such as an Amazon Machine Image (AMI) locally, or to reproduce the runtime environment identically on local and server systems.
 
-Of course Rocker images may be used for quickly deploying R code on remote servers without the use of the RStudio interface, running R in terminal mode over an `ssh` connection or simply in batch mode.
+Rocker images may be used for quickly deploying R code on remote servers without the use of the RStudio interface, running R in terminal mode over an `ssh` connection.
+As with any interactive docker instance, users should specify the interactive (`-i`) and terminal (`-t`, here combined with interactive as `-ti`) flags, and specify the desired executable environment (e.g.
+`R`, though other common options may be `Rscript` or `bash`):
+
+```
+docker run --rm -ti rocker/tidyverse R
+```
+
+This example shows the use of the `--rm` flag to indicate that the container should be removed when the interactive session is finished.
 

### 2017-06-18 manuscript.Rmd
WIP

@@
 
+## The debian:testing base
+
 The `debian:testing` stack aims to makes the most efficient use of upstream builds -- the precompiled `.deb` binaries provided by the Debian repositories.
@@
 2. These Dockerfiles always install the most recent available software
+3. These Dockerfiles can almost always build relatively quickly
 
@@
 3. The use of apt pinning may be unfamiliar to some users.
+4. `testing` and `unstable` do not recieve the same attention to security updates.
 
-### debian-stable stack
+### debian-stable base
 
 This stack emphasizes stability and reproducibility of the Docker build.
+The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.
+A user specifies the version desired using an image tag, e.g.
+`rocker/r-ver:3.3.1` will refer to an image with R version 3.3.1 installed.
+Omitting the tag is equivalent to using the tag `latest`, which, as the name implies, will always point to an image using the current R release.
+Thus, users wanting to create downstream Dockerfiles which are based on the current release at time but will continue to reconstruct the same environment in the future after newer R versions are released should explicilty include the corresponding version tag, e.g.
+`rocker/r-ver:3.4.0` at the time of writing, and not the `latest` tag.
+Users can also run the current development version of R using the tag `devel`, which is built nightly from R-devel sources.
+
+To facilitate installation of only contemporaneous versions of R packages on these images, the default CRAN mirror from which to install R packages is fixed to a snapshot of CRAN corresponding to the last date for which that version of R was the latest release.
+These snapshots are provided by the MRAN archive created by Revolution Analytics (now aquired by Microsoft), which archives daily snapshots of all of CRAN, from which a user can install packages with the usual `install.packages()` function  [@MRAN].
+Users can always override this default by passing any current CRAN repository explicitly.
+Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule, Bioconductor packages be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.
+
+These version tags are propegated throughout this stack: for instance, `rocker/tidyverse:devel` will provide the currently released versions of the R packages in the `tidyverse` [@tidyverse] installed on the nightly build of R-devel.
+Developers building packages on this stack are encouraged to tag their images accordingly as well.
+Table 3 indicates which versions of R are currently available in the stack, going back to `3.1.0`.
+While older versions may be added to the stack at a later date, we note that the MRAN snapshots began in starting 2014-09-17 and thus go back only to R `3.1` era.
+
+In this stack, the desired version of R is always built directly from source.
+Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack.
+Dependencies needed to compile R that are not required at runtime are removed once R is installed, keeping the base images light-weight for faster download times.
+While most system dependencies required by common R packages can still be installed from the `apt` repositories, occassionally a more recent version must be compiled from source (e.g.
+the Gibbs Sampling library, JAGS, and the geospatial toolkit GDAL, must both be compiled from source on images based on the Debian 8.0 release, codenamed `debian:jessie`).
+In this stack, users should never install R packages using `apt`, since this will install a second, older version of R from the Debian repositories and a dated version of the R package.
+These
+
+### Images
+
+The image `rocker/r-ver` is the functional equivalent of `r-base`, though slightly pared down.
+Currently `r-ver` has tags for all minor R versions back to `3.1.0` (as far back as MRAN snapshots of CRAN are avialable); see [all tags](https://hub.docker.com/r/rocker/r-ver/tags).
+Users can attempt to build other versions from the `r-ver` Dockerfile by specifying `--build-arg R_VERSION=<VERSION>`, though this is unlikely to work with very old R versions.
+Likewise, all non-current images are automatically pinned to a MRAN snapshot of CRAN from the last day that version was current.
+Users can set a custom snapshot date for the `r-ver` images by using `--build-arg BUILD_DATE=<DATE>` when building that Dockerfile.
+
+The image `rocker/tidyverse` is the functional equivalent of `rocker/hadleyverse`, minus the LaTeX part which is moved into a downstream `rocker/verse` in order to keep the `tidyverse` image more compact.
+Now that [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html) is a thing, the term "hadleyverse" is on the way out.
+
+The `rocker/rstudio` image builds with RStudio `v1.0.44` by default.
+This can be customized by specifying the desired version in `--build-arg RSTUDIO_VERSION=<VERSION>` if building locally from its Dockerfile.
+*Note that:* `rocker/rstudio:latest` is the only image name in this stack that already existed.
+For consistency, `rocker/rstudio:latest` now builds on `r-ver:latest` instead of `r-base:latest`.
+This change may break downstream images, which should either build explicitly on the `debian:testing` stack by using the `testing` tag (`rocker/rstudio:testing`) or be patched to support this change.
+
+image            |  Dockerfiles  
+---------------- | -------------
+[r-ver](https://hub.docker.com/r/rocker/r-ver/tags)         | [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-...-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/)  [![](https://img.shields.io/badge/Dockerfile-v3.1.0-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/3.1.0/Dockerfile)
+[rstudio](https://hub.docker.com/r/rocker/rstudio/tags)     |  [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/rstudio/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker/blob/master/rstudio/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker/blob/master/rstudio/3.3.1/Dockerfile)
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse/tags) | [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/3.3.2/Dockerfile)   [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/3.3.1/Dockerfile)
+[verse](https://hub.docker.com/r/rocker/verse/tags)         |  [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/3.3.1/Dockerfile)
+
+## Maintenance and Updates
+
+These images are actively maintained.
+This means that while an effort is made to preserve the general function of these images over time, both these Dockerfiles and the resulting images are subject to some change over time.
+In particular:
+
+- Images are regularly re-built on Docker Hub whenever their base image changes, starting with changes to `debian` Docker image.
+This is the rough equivalent of running `apt-get upgrade` on `debian`, since all `apt-get` commands are re-run and will pull in the most current sources.
+This allows the images to receive security updates to any packages installed from the `debian` repositories, but will not in general change the versions of any software and is very unlikely to break anything.
+
+- The Dockerfiles themselves are subject to change, to improve performance, ease of use, readability, or other concerns raised in the issues.
+These changes should also not alter the general behavior of R or R packages on the image.
+These changes can be seen in the git history.
+The [rocker-versioned](https://github.com/rocker-org/rocker-versioned) repo will use its own semantic version tagging to indicate changes to this repository, with snapshots from these tags archived on Zenodo.
+
+## License ##
+
+The Dockerfiles in this repository are licensed under the GPL 2 or later.
 

### 2017-06-22 manuscript.Rmd
WIP

--- 2017-06-17 manuscript.Rmd
@@
 preamble: >
-  % Any extra latex you need in the preamble
-output: rticles::rjournal_article
+   \usepackage{longtable}
+   \usepackage{svg}
+output: 
+  rticles::rjournal_article:
+        latex_engine: lualatex
 ---
@@
 application in an isolated environment \citep{what-docker}.
+
+## Rocker organization and workflow
+
+The Rocker Project consists of a suite of images built automatically by and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.
+Source Dockerfiles, supporting scripts and documentation are hosted on GitHub.
+The issues tracker and pull requests are used to provide community input, disucssions and contributions to these images.
+The rocker project wiki <https://github.com/rocker-org/rocker/wiki> provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
+
+- GitHub, Docker Hub (Docker Cloud)
+
+ Automated builds on Docker Hub help ensure the integrity of the build process and ensure that the image build corresponds exactly to Dockerfile on GitHub.
 
@@
 3. The use of apt pinning may be unfamiliar to some users.
+4. `testing` and `unstable` do not recieve the same attention to security updates.
 
-### debian-stable stack
+### debian-stable base
 
 This stack emphasizes stability and reproducibility of the Docker build.
+The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.
+A user specifies the version desired using an image tag, e.g.
+`rocker/r-ver:3.3.1` will refer to an image with R version 3.3.1 installed.
+Omitting the tag is equivalent to using the tag `latest`, which, as the name implies, will always point to an image using the current R release.
+Thus, users wanting to create downstream Dockerfiles which are based on the current release at time but will continue to reconstruct the same environment in the future after newer R versions are released should explicilty include the corresponding version tag, e.g.
+`rocker/r-ver:3.4.0` at the time of writing, and not the `latest` tag.
+Users can also run the current development version of R using the tag `devel`, which is built nightly from R-devel sources from `subversion`.
+
+**MRAN archives**: To facilitate installation of only contemporaneous versions of R packages on these images, the default CRAN mirror from which to install R packages is fixed to a snapshot of CRAN corresponding to the last date for which that version of R was the latest release.
+These snapshots are provided by the MRAN archive created by Revolution Analytics (now aquired by Microsoft), which archives daily snapshots of all of CRAN, from which a user can install packages with the usual `install.packages()` function  [@MRAN].
+Users can always override this default by passing any current CRAN repository explicitly.
+Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule, Bioconductor packages be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.
+
+**Version tags**:  These version tags are propegated throughout this stack: for instance, `rocker/tidyverse:devel` will provide the currently released versions of the R packages in the `tidyverse` [@tidyverse] installed on the nightly build of R-devel.
+Developers building packages on this stack are encouraged to tag their images accordingly as well.
+Table 3 indicates which versions of R are currently available in the stack, going back to `3.1.0`.
+While older versions may be added to the stack at a later date, we note that the MRAN snapshots began in starting 2014-09-17 and thus go back only to R `3.1` era.
+Each tag must be built from a separate Dockerfile, enabling minor differences in the build instructions to accomidate changing dependencies.
+Dockerfiles for past versions (e.g.
+< `3.4.0` currently) are intended to remain static over the long term, while the tag for the current version, `latest`, and `devel` may be tweaked to accomidate new features or dependencies.
+
+In this stack, the desired version of R is always built directly from source rather than the `apt` repositories.
+Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack.
+Version tags `3.3.3` and older are based on the Debian 8.0 release, codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much newer versions of common system dependencies and compilers.
+Dependencies needed to compile R that are not required at runtime are removed once R is installed, keeping the base images light-weight for faster download times.
+While most system dependencies required by common R packages can still be installed from the `apt` repositories, occassionally a more recent version must be compiled from source (e.g.
+the Gibbs Sampling library, JAGS, and the geospatial toolkit GDAL, must both be compiled from source on `debian:jessie` images).
+In this stack, users should never install R packages, using `apt`, since this will install a second, older version of R from the Debian repositories and a dated version of the R package (since any `r-cran-pkgname` package in the Debian repositories will depend on `r-base` in `apt` as well).
+
+**Build schedule**: All images are built automatically from their corresponding Dockerfiles (found in the GitHub repositories `rocker-org/rocker-versioned` and `rocker-org/geospatial`).
+A `cron` job sends nightly build triggers to Docker Hub to rebuild the `latest` and `devel` tagged images throughout the stack.
+To decrease load on the hub, build triggers for the numeric version tags are sent monthly.
+Allthough the Dockerfiles for older R versions installs an almost-identical software environment every time, the monthly rebuilding of these images on Docker Hub ensures they continue to recieve Debian security updates from upstream, and proves the build recipe still executes successfully.
+Note that rebuilding images with software from external repositories never produces a bitwise identical image, and thus the image identifier hash will change at each build.
+
+image            | description                               | size   | downloads 
+---------------- | ----------------------------------------- | ------ | ------- 
+[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 2,000
+[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000
+[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
+[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 1,000
+
+Table: The `rocker-versioned` stack of images
+
+image            | description                               | size   | downloads 
+---------------- | ----------------------------------------- | ------ | ------- 
+[r-base](https://hub.docker.com/r/_/r-base)        | Version-stable base R & src build tools  | 254 MB | 510,000   
+[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
+[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanatizers | 1.1 GB  | 1,000  
+[r-devel-ubsan-clan](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanatizers, clang c compiler (instead of gcc) | 1.1 GB | 525
+[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
+[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000
+
+Table: the `debian:testing` image stack
+
+tag     | apt repos | MRAN date    | Build frequency  | images with tag 
+--------|-----------|--------------|------------------|-------------------------------           
+devel   | `stretch` | current date | nigthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.4.0   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.3.3   | `jessie`  | 2017-04-21   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial` 
+3.3.2   | `jessie`  | 2017-03-06   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.3.1   | `jessie`  | 2016-10-31   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.3.0   | `jessie`  | 2016-06-21   | monthly          | `r-ver`
+3.2.0   | `jessie`  | 2015-06-18   | monthly          | `r-ver`
+3.1.0   | `jessie`  | 2014-09-17   | monthly          | `r-ver`
+
+Table: Available tags in the `rocker-versioned` stack.
+
+Customizing builds: `--build-arg` options in `r-ver`: `R_VERSION`, `BUILD_DATE`. in `rstudio`: `RSTUDIO_VERSION`, `PANDOC_TEMPLATES_VERSION`.  
+Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.  These can be altered or extended in downstream images.
+
+## Maintenance and Updates
+
+These images are actively maintained.
+This means that while an effort is made to preserve the general function of these images over time, both these Dockerfiles and the resulting images are subject to some change over time.
+In particular:
+
+- Images are regularly re-built on Docker Hub whenever their base image changes, starting with changes to `debian` Docker image.
+This is the rough equivalent of running `apt-get upgrade` on `debian`, since all `apt-get` commands are re-run and will pull in the most current sources.
+This allows the images to receive security updates to any packages installed from the `debian` repositories, but will not in general change the versions of any software and is very unlikely to break anything.
+
+- The Dockerfiles themselves are subject to change, to improve performance, ease of use, readability, or other concerns raised in the issues.
+These changes should also not alter the general behavior of R or R packages on the image.
+These changes can be seen in the git history.
+The [rocker-versioned](https://github.com/rocker-org/rocker-versioned) repo will use its own semantic version tagging to indicate changes to this repository, with snapshots from these tags archived on Zenodo.
+
+## License ##
+
+The Dockerfiles in this repository are licensed under the GPL 2 or later.
+
+# Use cases
 
@@
 
-## Rocker organization and workflow
-
-The Rocker Project consists of a suite of images built automatically by and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.
-Source Dockerfiles, supporting scripts and documentation are hosted on GitHub.
-The issues tracker and pull requests are used to provide community input, disucssions and contributions to these images.
-The rocker project wiki <https://github.com/rocker-org/rocker/wiki> provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
-
 - Table of main repos, links to Docker-Hub & GitHub addresses
@@
 
-## The r-base image
-
-The r-base image is the official, verified Docker image for R.
-Written and maintained by the Rocker developers on behalf of the Docker project, this image is built and hosted directly by Docker's own developers, who also approve any changes or updates to the image.
-
-example use.
-
-Link to further details in Rocker wiki
-
-## The r-devel images
-
-These images provide both the current and under-development versions of R, including images with particular compiler option configurations (UBSAN, clang).
-Package authors submitting R packages to the Central R Archive Network (CRAN) are expected to test their packages on both current and development versions of R prior to submission.
-
-## rocker-versioned stack
-
-example use
-
-screenshot
-
-Custom configurations
-
-Link to further details in Rocker wiki
-
-## Additional images
-
-## Metrics of use
-
 ## Summary
@@
 This file is only a basic article template.
-For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](http://journal.r-project.org/latex/RJauthorguide.pdf).
+For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
 

### 2017-06-23 manuscript.Rmd
WIP

@@
    \usepackage{svg}
-output: 
-  rticles::rjournal_article:
-        latex_engine: lualatex
+output: rticles::rjournal_article
 ---
@@
 The Rocker Project consists of a suite of images built automatically by and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.
-Source Dockerfiles, supporting scripts and documentation are hosted on GitHub.
+Source Dockerfiles, supporting scripts and documentation are hosted on GitHub under the organization `rocker-org`, <https://github.com/rocker-org>.
 The issues tracker and pull requests are used to provide community input, disucssions and contributions to these images.
-The rocker project wiki <https://github.com/rocker-org/rocker/wiki> provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
-
-- GitHub, Docker Hub (Docker Cloud)
-
- Automated builds on Docker Hub help ensure the integrity of the build process and ensure that the image build corresponds exactly to Dockerfile on GitHub.
+The rocker project wiki provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
 
@@
 
-The Rocker Project aims to provide a small core if Docker images that serve both as convenient base images for a wide range of possible applications while still providing a 'batteries included' approach of images that can be used out of the box.
+The Rocker Project aims to provide a small core if Docker images that serve as convenient 'base' images on which other users can build custom R environments by writing their own Dockerfiles, while also providing a 'batteries included' approach of images that can be used out of the box.
 The challenges of balancing diverse needs driven by very different use cases (discussed in more detail below) against the overarching goals of creating images that are not sufficiently light-weight, easy to use and easy to maintain is a difficult art.
@@
 
-<!--What is a Dockerfile? How do they stack? -->
-
-<!-- some overview on using vs extending images (i.e. writing derivative Dockerfiles). -->
-
-The Rocker Project can be largely divided into two stacks which address different needs.
+All Rocker images are based on the Debian Linux Distribution.
+The Debian platform provides a small base image, the well known `apt` package management system and rich ecosystem of software libraries, making it the base image of choice for Docker images, including many of the "official" images maintained by Docker's own development team.
+The Debian platform is also perhaps the best supported Linux platform within the R community, including an active `r-sig-debian` listserve.
+Unfortunately, the relatively long period between stable Debian releases (roughly two years recently) means that software in the `debian:stable` releases can lag significantly behind current releases of popular software, including R.
+More recent versions of packages can be found in the pre-release distribution, `debian:testing`, while the very latest binary builds can be found on `debian:unstable`.
+The Rocker Project can be largely divided into two stacks which address different needs, defined by which Debian distribution they are based on.
 The first stack is based on `debian:testing`, the second, more recently introduced stack is based only on `debian` stable releases.
@@
 
-The Debian platform provides a small base image, the well known `apt` package management system and rich ecosystem of software libraries, making it the base image of choice for Docker images, including many of the "official" images maintained by Docker's own development team.
-The Debian platform is also perhaps the best supported Linux platform within the R community, including an active `r-sig-debian` listserve.
-Unfortunately, the relatively long period between stable Debian releases (roughly two years recently) means that software in the `debian:stable` releases can lag significantly behind current releases.
-This issue is the primary motivation behind the `debian:testing` stack.
-
-The Debian release cycle follows three stages: `unstable`, `testing`, and `stable`.
-
-## The debian:testing base
+## The `debian:testing`-based images
 
@@
 
-The `debian:testing` stack thus offers the following advantages:
+Relative to `stable`, the `debian:testing` stack thus offers the following advantages:
 
@@
 
-1. These Dockerfiles require occassional minor maintenance to ensure successful builds
+1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
-3. The use of apt pinning may be unfamiliar to some users.
-4. `testing` and `unstable` do not recieve the same attention to security updates.
+3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
+4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
 
-### debian-stable base
+### Images overview
+
+The `debian:testing`-based stack currently includes seven images actively maintained by the Rocker development team (Table 1).
+`r-base` builds on `debian:testing`, and the other six in the stack each build directly from `r-base`.
+The `r-base` image is unique in that it is designated as the official image for the R language by the Docker organization itself.
+This image is reviewed and built by employees of Docker Inc from the Dockerfile maintained by the Rocker team.
+Consequently, users should refer to this image in Docker commands without an organization napespace, e.g.
+`docker pull r-base` to access the official image.
+All other images in the Rocker project are not individually reviewed and built by Docker Inc and must be refered to require the organization namespace on Docker Hub, e.g.
+`docker pull rocker/r-devel`.
+
+Several of the images in this stack are oriented towards the R development community.
+`r-devel`, `drd`, `r-devel-san` & `r-devel-ubsan-clang` all add a copy of the development version of R side-by-side the current release of R provided by `r-base`.
+On these images, the development version is aliased to `RD` to distinguish from the current release, `R`.
+As the names suggest, each provide slightly different configurations.
+Of particular interest are the images providing development R built with support for C/C++ address sanitizers, which are notoriously difficult to configure [@edd_sanitizers].
+
+As these images focus on developers and/or as base images for custom uses, this stack does include many specific R packages.
+Additional dependencies and packages can easily be installed from `apt`.
+R packages not available in the `apt` repositories can be installed directly from CRAN using the `littler` scripts, see *Extending Rocker images* section below.
+
+This stack also include the images `shiny` and `rstudio:testing` that provide Shiny server and RStudio server IDE from RStudio Inc, built on the `r-base` image.
+RStudio and Shiny are registered trademarks of RStudio Inc, and their use and the distibution of their software in binary form on Docker Hub has been granted to the Rocker project by explicit permission from RStudio.
+Users should review RStudio's trademark use policy (<http://www.rstudio.com/about/trademark/>) and address inquiries about further distribution or other questions to <permissions@rstudio.com>.
+The Rocker project also provides images with RStudio server and Shiny server in the stable versioned stack.
+
+**Build schedule**: The official `r-base` image is rebuilt by Docker following any updates to the official `debian` images (roughly every few weeks).
+The rest of the stack uses build triggers that rebuild the images either whenever `r-base` is updated or the Dockerfile sources are updated on the corresponding GitHub repository.
+The only exception in this stack is the `drd` image, which is rebuilt each week by cron trigger.
+
+image            | description                               | size   | downloads 
+---------------- | ----------------------------------------- | ------ | ------- 
+[r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 510,000   
+[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
+[drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built almost daily  | 571 MB | 4,000
+[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanatizers | 1.1 GB  | 1,000  
+[r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
+[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
+[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000
+
+Table: the `debian:testing` image stack
+
+<!-- Should we mention r-apt separately? -->
+
+<!-- Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?  -->
+
+### The `debian:stable`-based stack
 
 This stack emphasizes stability and reproducibility of the Docker build.
+This stack was introduced much more recently, starting in November of 2016, in response to considerable user input and requests.
 The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.
@@
 Users can always override this default by passing any current CRAN repository explicitly.
-Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule, Bioconductor packages be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.
+Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule.
+Thus, Bioconductor packages can be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.
 
@@
 
-In this stack, the desired version of R is always built directly from source rather than the `apt` repositories.
+**Installation**: In this stack, the desired version of R is always built directly from source rather than the `apt` repositories.
 Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack.
@@
 To decrease load on the hub, build triggers for the numeric version tags are sent monthly.
-Allthough the Dockerfiles for older R versions installs an almost-identical software environment every time, the monthly rebuilding of these images on Docker Hub ensures they continue to recieve Debian security updates from upstream, and proves the build recipe still executes successfully.
+Although the Dockerfiles for older R versions installs an almost-identical software environment every time, the monthly rebuilding of these images on Docker Hub ensures they continue to recieve Debian security updates from upstream, and proves the build recipe still executes successfully.
 Note that rebuilding images with software from external repositories never produces a bitwise identical image, and thus the image identifier hash will change at each build.
+
+### Images overview
+
+In this stack, each image builds on the previous image, rather than all other images building directly on the base image, as in the `testing` stack.
+Table 2 lists the names and descriptions of the five images in this stack, along with image size and approximate download counts from Docker Hub.
+Sizes reflect (compressed) cumulative size: a user who has already downloaded the most recent version of `r-ver` and then pulls a copy of `rstudio` image will only need to download the additional 115 MB in the `rstudio` layers and not the full 334 MB listed.
+This linear design limits flexibility (no option for `tidyverse` without `rstudio`) but simplifies use and maintance.
+While no single environment will be optimal for everyone, both the packages selected in this stack and the stack ordering reflect considerable community input and tuning.
+
+The `rstudio` image includes a lightweight, easy-to-use and docker-friendly `init` system, s6 [@s6] for running persistant services, including the RStudio server.
+This system provides a convenient way for downstream Dockerfile developers to add additional persistent services (such as an ssh server) to a single container, or additional start-up or shutdown scripts that should be run when an instance starts up or shuts down.
+The `rstudio` image uses such a start-up script to configure user settings such as login password and permissions through environmental variables at run time, see *Using Rocker Images* below.
+
+The `tidyverse` image contains all required and suggested dependencies of the commonly used `tidyverse` and `devtools` R packages, including external database libraries (e.g.
+Maria-DB and Postgresql).
+Users should consult the package Dockerfiles or `installed.packages()` list directly for a complete list of installed packages.
+The `verse` library adds commonly used dependencies, notably a large but not comprehensive LaTeX environment and java development libraries.
+Previously the Rocker project provided the image `hadleyverse` which was since divided into `tidyverse` and `verse` through community input.
 
@@
 [rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
-[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000 (+46,000 as `hadleyverse`)
 [verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
@@
 Table: The `rocker-versioned` stack of images
-
-image            | description                               | size   | downloads 
----------------- | ----------------------------------------- | ------ | ------- 
-[r-base](https://hub.docker.com/r/_/r-base)        | Version-stable base R & src build tools  | 254 MB | 510,000   
-[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
-[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanatizers | 1.1 GB  | 1,000  
-[r-devel-ubsan-clan](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanatizers, clang c compiler (instead of gcc) | 1.1 GB | 525
-[rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
-[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000
-
-Table: the `debian:testing` image stack
 
@@
 
-Customizing builds: `--build-arg` options in `r-ver`: `R_VERSION`, `BUILD_DATE`. in `rstudio`: `RSTUDIO_VERSION`, `PANDOC_TEMPLATES_VERSION`.  
-Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.  These can be altered or extended in downstream images.
+Several images in the `rocker-versioned` stack can be customized on build when built locally (rather than pulling prebuilt images from Docker Hub) by using the `--build-arg` option of `docker build`.
+In the `r-ver` image, users can set `R_VERSION`, `BUILD_DATE` (MRAN default snapshot).
+In the `rstudio` image users can set `RSTUDIO_VERSION` (otherwise defaults to the most recent version), and the `PANDOC_TEMPLATES_VERSION` .
+
+<!--
+This stack also makes use of Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.
+These can be altered or extended in downstream images.
+-->
+
+<!--
 
@@
 
-# Use cases
+-->
+
+## Using Rocker images
+
+While more detailed documentation on the use of Rocker images can be found on the Rocker wiki, we provide a conceptual overview here which illustrates common patterns and capabilities.
+
+**Interactive shell**:
+
+**RStudio instance**:
+
+**Linking local volumes**: Docker
+
+Users of the Docker's GUI interface, Kitematic who use the `rstudio`-derived images will automatically have their local kitematic directory linked to the `/home/rstudio/kitematic` directory on the Docker container.
+
+**Root vs user instances**: By default, all rocker containers run as root.
+This is consistent with standard Docker practice, allowing both interactive users and downstream Dockerfiles to install software directly without having to switch to root.
+The main snag of this approach comes if a user links a local volume and modifies files on that volume, which will result in the local file being owned by `root` and not the default user.
+To avoid this, specify the non-root user at runtime (username `docker` on the `r-base` stack, or `rstudio` on images deriving from rstudio) when running an interactive shell.
+For users accessing R through RStudio, this is not necessary.
+The docker container will run as root, but a user logs in through the RStudio server web interface as the non-root user "rstudio", and thus any changes made to linked volumes will not alter file permissions on the home directory.
+For more details on user permissions, see the Rocker wiki.
+
+## Extending Rocker images
+
+Often Rocker images are not deployed directly, but used as a base image onto which users add software for their own needs in a custom Dockerfile.
+We offer a few recommendations here to streamline writing these extensions and promote best practices.
+
+- littler scripts
+
+ To streamline the installation of R packages in downstream Dockerfiles, the `r-base` image provides `littler` package [@littler] and associated `install2.r` script, enabling users to install additional packages with a command such as:
+
+    install2.r pkg1 pgk2 pkg3 ...
+
+- Dockerfile best practices
+- automated builds, build triggers
+- version-stable builds
+
+## Use cases
 
@@
 
-One of the most popular uses of the Rocker containers is to provide a fast and reliable mechanism to deploy an R envionment on a remote server, such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's Jetstream servers, or private or institutional server hardware.
+One common use case for Rocker containers is to provide a fast and reliable mechanism to deploy an R envionment on a remote server, such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's Jetstream servers, or private or institutional server hardware.
 Being able to test code in a predictable, pre-configured R environment on a local machine and then run the same code in an identical environment on a remote server (e.g.
@@
 
-```
     docker run --rm -ti rocker/tidyverse R
-```
 
@@
 
-## Rocker base images
-
-Image dependency graph
-
-## Summary
-
-This file is only a basic article template.
-For full details of _The R Journal_ style and information on how to prepare your article for submission, see the [Instructions for Authors](https://journal.r-project.org/share/author-guide.pdf).
-
 \bibliography{RJreferences}

### 2017-06-23 manuscript.Rmd
hard wrap for better diffs

@@
 
-The Rocker project was launched in October 2014 as a collaboration between CB & DE to provide high-quality
-Docker images containing the R environment \citep{edd2014}. Since that time, the project has seen both 
-considerable uptake in the community and substantial development and evolution. Here we seek to document
-the project's objectives and uses.
+The Rocker project was launched in October 2014 as a collaboration
+between CB & DE to provide high-quality Docker images containing the
+R environment \citep{edd2014}. Since that time, the project has seen
+both considerable uptake in the community and substantial development
+and evolution. Here we seek to document the project's objectives and uses.
 
@@
 
-Docker is a popular open source tool to create, distribute, deploy, and run software applications using
-*containers*.  Containers provide a virtual environment requiring everything an application needs to run:
-code, runtime, system tools, system runtime.  Docker containers are lightweight (sharing the operating system kernel,
-starting instantly, using a layered filesystem to minimize disk footprint and download time), built on open standards
-that run on all major platforms (Linux, Mac, Windows) and provide an added layer of security by running an
-application in an isolated environment \citep{what-docker}.
+Docker is a popular open source tool to create, distribute, deploy,
+and run software applications using *containers*.  Containers provide a
+virtual environment requiring everything an application needs to run:
+code, runtime, system tools, system runtime.  Docker containers are
+lightweight (sharing the operating system kernel, starting instantly,
+using a layered filesystem to minimize disk footprint and download time),
+built on open standards that run on all major platforms (Linux, Mac,
+Windows) and provide an added layer of security by running an application
+in an isolated environment \citep{what-docker}.
 
@@
 
-The Rocker Project consists of a suite of images built automatically by and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.
-Source Dockerfiles, supporting scripts and documentation are hosted on GitHub under the organization `rocker-org`, <https://github.com/rocker-org>.
-The issues tracker and pull requests are used to provide community input, disucssions and contributions to these images.
-The rocker project wiki provides a place to synthesize community-contributed documentation, use-cases, and other knowledge about using the Rocker images.
+The Rocker Project consists of a suite of images built automatically by
+and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
+Dockerfiles, supporting scripts and documentation are hosted on GitHub
+under the organization `rocker-org`, <https://github.com/rocker-org>.
+The issues tracker and pull requests are used to provide community input,
+disucssions and contributions to these images. The rocker project wiki
+provides a place to synthesize community-contributed documentation,
+use-cases, and other knowledge about using the Rocker images.
 
@@
 
-The Rocker Project aims to provide a small core if Docker images that serve as convenient 'base' images on which other users can build custom R environments by writing their own Dockerfiles, while also providing a 'batteries included' approach of images that can be used out of the box.
-The challenges of balancing diverse needs driven by very different use cases (discussed in more detail below) against the overarching goals of creating images that are not sufficiently light-weight, easy to use and easy to maintain is a difficult art.
-The implementation in both individual Rocker images and image stacks can never perfect that balance for everyone, but today reflects the considerable community input and testing over past few years.
+The Rocker Project aims to provide a small core if Docker images that
+serve as convenient 'base' images on which other users can build custom
+R environments by writing their own Dockerfiles, while also providing a
+'batteries included' approach of images that can be used out of the box.
+The challenges of balancing diverse needs driven by very different use
+cases (discussed in more detail below) against the overarching goals
+of creating images that are not sufficiently light-weight, easy to use
+and easy to maintain is a difficult art.  The implementation in both
+individual Rocker images and image stacks can never perfect that balance
+for everyone, but today reflects the considerable community input and
+testing over past few years.
 
 All Rocker images are based on the Debian Linux Distribution.
-The Debian platform provides a small base image, the well known `apt` package management system and rich ecosystem of software libraries, making it the base image of choice for Docker images, including many of the "official" images maintained by Docker's own development team.
-The Debian platform is also perhaps the best supported Linux platform within the R community, including an active `r-sig-debian` listserve.
-Unfortunately, the relatively long period between stable Debian releases (roughly two years recently) means that software in the `debian:stable` releases can lag significantly behind current releases of popular software, including R.
-More recent versions of packages can be found in the pre-release distribution, `debian:testing`, while the very latest binary builds can be found on `debian:unstable`.
-The Rocker Project can be largely divided into two stacks which address different needs, defined by which Debian distribution they are based on.
-The first stack is based on `debian:testing`, the second, more recently introduced stack is based only on `debian` stable releases.
-These stacks have different aims and thus provide different images, as shown in Tables 1 & 2.
+The Debian platform provides a small base image, the well known `apt`
+package management system and rich ecosystem of software libraries,
+making it the base image of choice for Docker images, including many
+of the "official" images maintained by Docker's own development team.
+The Debian platform is also perhaps the best supported Linux platform
+within the R community, including an active `r-sig-debian` listserve.
+Unfortunately, the relatively long period between stable Debian releases
+(roughly two years recently) means that software in the `debian:stable`
+releases can lag significantly behind current releases of popular
+software, including R.  More recent versions of packages can be found in
+the pre-release distribution, `debian:testing`, while the very latest
+binary builds can be found on `debian:unstable`.  The Rocker Project
+can be largely divided into two stacks which address different needs,
+defined by which Debian distribution they are based on.  The first stack
+is based on `debian:testing`, the second, more recently introduced stack
+is based only on `debian` stable releases.  These stacks have different
+aims and thus provide different images, as shown in Tables 1 & 2.
 
@@
 
-The `debian:testing` stack aims to makes the most efficient use of upstream builds -- the precompiled `.deb` binaries provided by the Debian repositories.
-It is both quicker and easier to install software from binaries, since the package manager (`apt`) manages the necessary dependencies and it bypasses the time-consuming process of compiling from source.
-Basing this stack on `debian:testing` means that much more recent versions of commonly used libraries and compilers are available as binaries than would be found in a Debian stable release.
-In order to provide access to the most recent avialble binaries, this stack uses apt-pinning [@apt-pinning] to allow the `apt` package manager to also install binaries from `debian:unstable`, which represents the most recent, bleeding edge of packages built for Debian when necessary.
-For instance, the `r-base` image provided by Rocker installs the most recent version of R as a binary from Debian `unstable`.
-Similarly, recent versions of many popular R packages can also be installed through the package manager, e.g.
-`apt-get install r-cran-XML`.
-This can be particularly helpful for packages with external system dependencies (such as `libxml2-dev` in this example) which cannot be installed from the R console.
+The `debian:testing` stack aims to makes the most efficient use of
+upstream builds -- the precompiled `.deb` binaries provided by the
+Debian repositories.  It is both quicker and easier to install software
+from binaries, since the package manager (`apt`) manages the necessary
+dependencies and it bypasses the time-consuming process of compiling
+from source.  Basing this stack on `debian:testing` means that much more
+recent versions of commonly used libraries and compilers are available
+as binaries than would be found in a Debian stable release.  In order
+to provide access to the most recent avialble binaries, this stack uses
+apt-pinning [@apt-pinning] to allow the `apt` package manager to also
+install binaries from `debian:unstable`, which represents the most recent,
+bleeding edge of packages built for Debian when necessary.  For instance,
+the `r-base` image provided by Rocker installs the most recent version of
+R as a binary from Debian `unstable`.  Similarly, recent versions of many
+popular R packages can also be installed through the package manager,
+e.g. `apt-get install r-cran-XML`.  This can be particularly helpful
+for packages with external system dependencies (such as `libxml2-dev`
+in this example) which cannot be installed from the R console.
 
-As the names `testing` and `unstable` imply, this approach is not without a downside.
-The particular version of any given package can change as packages move from `unstable` into `testing` and newer versions are sent to `unstable` during the normal course of Debian development.
-This can occassionally break an previously working installation command in a Dockerfile until the maintainer redirects the package manager to install a the package from the `unstable` sources that could previously be installed from `testing`, or vice versa (using the `-t` option in `apt`, see Box 1 on apt-pinning.)
+As the names `testing` and `unstable` imply, this approach is not
+without a downside.  The particular version of any given package can
+change as packages move from `unstable` into `testing` and newer versions
+are sent to `unstable` during the normal course of Debian development.
+This can occassionally break an previously working installation command
+in a Dockerfile until the maintainer redirects the package manager to
+install a the package from the `unstable` sources that could previously
+be installed from `testing`, or vice versa (using the `-t` option in
+`apt`, see Box 1 on apt-pinning.)
 
-Relative to `stable`, the `debian:testing` stack thus offers the following advantages:
+Relative to `stable`, the `testing` stack thus offers the following advantages:
 
@@
 
-The `debian:testing`-based stack currently includes seven images actively maintained by the Rocker development team (Table 1).
-`r-base` builds on `debian:testing`, and the other six in the stack each build directly from `r-base`.
-The `r-base` image is unique in that it is designated as the official image for the R language by the Docker organization itself.
-This image is reviewed and built by employees of Docker Inc from the Dockerfile maintained by the Rocker team.
-Consequently, users should refer to this image in Docker commands without an organization napespace, e.g.
-`docker pull r-base` to access the official image.
-All other images in the Rocker project are not individually reviewed and built by Docker Inc and must be refered to require the organization namespace on Docker Hub, e.g.
-`docker pull rocker/r-devel`.
+The `debian:testing`-based stack currently includes seven images actively
+maintained by the Rocker development team (Table 1).  `r-base` builds on
+`debian:testing`, and the other six in the stack each build directly from
+`r-base`.  The `r-base` image is unique in that it is designated as the
+official image for the R language by the Docker organization itself.
+This image is reviewed and built by employees of Docker Inc from the
+Dockerfile maintained by the Rocker team.  Consequently, users should
+refer to this image in Docker commands without an organization napespace,
+e.g. `docker pull r-base` to access the official image.  All other images
+in the Rocker project are not individually reviewed and built by Docker
+Inc and must be refered to require the organization namespace on Docker
+Hub, e.g. `docker pull rocker/r-devel`.
 
-Several of the images in this stack are oriented towards the R development community.
-`r-devel`, `drd`, `r-devel-san` & `r-devel-ubsan-clang` all add a copy of the development version of R side-by-side the current release of R provided by `r-base`.
-On these images, the development version is aliased to `RD` to distinguish from the current release, `R`.
+Several of the images in this stack are oriented towards the R development
+community. `r-devel`, `drd`, `r-devel-san` & `r-devel-ubsan-clang` all
+add a copy of the development version of R side-by-side the current
+release of R provided by `r-base`.  On these images, the development
+version is aliased to `RD` to distinguish from the current release, `R`.
 As the names suggest, each provide slightly different configurations.
-Of particular interest are the images providing development R built with support for C/C++ address sanitizers, which are notoriously difficult to configure [@edd_sanitizers].
+Of particular interest are the images providing development R built with
+support for C/C++ address sanitizers, which are notoriously difficult
+to configure [@edd_sanitizers].
 
-As these images focus on developers and/or as base images for custom uses, this stack does include many specific R packages.
-Additional dependencies and packages can easily be installed from `apt`.
-R packages not available in the `apt` repositories can be installed directly from CRAN using the `littler` scripts, see *Extending Rocker images* section below.
+As these images focus on developers and/or as base images for custom uses,
+this stack does include many specific R packages. Additional dependencies
+and packages can easily be installed from `apt`.  R packages not available
+in the `apt` repositories can be installed directly from CRAN using the
+`littler` scripts, see *Extending Rocker images* section below.
 
-This stack also include the images `shiny` and `rstudio:testing` that provide Shiny server and RStudio server IDE from RStudio Inc, built on the `r-base` image.
-RStudio and Shiny are registered trademarks of RStudio Inc, and their use and the distibution of their software in binary form on Docker Hub has been granted to the Rocker project by explicit permission from RStudio.
-Users should review RStudio's trademark use policy (<http://www.rstudio.com/about/trademark/>) and address inquiries about further distribution or other questions to <permissions@rstudio.com>.
-The Rocker project also provides images with RStudio server and Shiny server in the stable versioned stack.
+This stack also include the images `shiny` and `rstudio:testing` that
+provide Shiny server and RStudio server IDE from RStudio Inc, built on the
+`r-base` image.  RStudio and Shiny are registered trademarks of RStudio
+Inc, and their use and the distibution of their software in binary form on
+Docker Hub has been granted to the Rocker project by explicit permission
+from RStudio. Users should review RStudio's trademark use policy
+(<http://www.rstudio.com/about/trademark/>) and address inquiries about
+further distribution or other questions to <permissions@rstudio.com>.
+The Rocker project also provides images with RStudio server and Shiny
+server in the stable versioned stack.
 
-**Build schedule**: The official `r-base` image is rebuilt by Docker following any updates to the official `debian` images (roughly every few weeks).
-The rest of the stack uses build triggers that rebuild the images either whenever `r-base` is updated or the Dockerfile sources are updated on the corresponding GitHub repository.
-The only exception in this stack is the `drd` image, which is rebuilt each week by cron trigger.
+**Build schedule**: The official `r-base` image is rebuilt by Docker
+following any updates to the official `debian` images (roughly every
+few weeks).  The rest of the stack uses build triggers that rebuild the
+images either whenever `r-base` is updated or the Dockerfile sources are
+updated on the corresponding GitHub repository.  The only exception in
+this stack is the `drd` image, which is rebuilt each week by cron trigger.
 
@@
 This stack emphasizes stability and reproducibility of the Docker build.
-This stack was introduced much more recently, starting in November of 2016, in response to considerable user input and requests.
-The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.
-A user specifies the version desired using an image tag, e.g.
-`rocker/r-ver:3.3.1` will refer to an image with R version 3.3.1 installed.
-Omitting the tag is equivalent to using the tag `latest`, which, as the name implies, will always point to an image using the current R release.
-Thus, users wanting to create downstream Dockerfiles which are based on the current release at time but will continue to reconstruct the same environment in the future after newer R versions are released should explicilty include the corresponding version tag, e.g.
-`rocker/r-ver:3.4.0` at the time of writing, and not the `latest` tag.
-Users can also run the current development version of R using the tag `devel`, which is built nightly from R-devel sources from `subversion`.
+This stack was introduced much more recently, starting in November
+of 2016, in response to considerable user input and requests. The key
+feature of this stack is the ability to run older versions of R, along
+with the contemporaneous versions of R packages.  A user specifies the
+version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
+refer to an image with R version 3.3.1 installed.  Omitting the tag
+is equivalent to using the tag `latest`, which, as the name implies,
+will always point to an image using the current R release.  Thus, users
+wanting to create downstream Dockerfiles which are based on the current
+release at time but will continue to reconstruct the same environment in
+the future after newer R versions are released should explicilty include
+the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
+of writing, and not the `latest` tag.  Users can also run the current
+development version of R using the tag `devel`, which is built nightly
+from R-devel sources from `subversion`.
 
-**MRAN archives**: To facilitate installation of only contemporaneous versions of R packages on these images, the default CRAN mirror from which to install R packages is fixed to a snapshot of CRAN corresponding to the last date for which that version of R was the latest release.
-These snapshots are provided by the MRAN archive created by Revolution Analytics (now aquired by Microsoft), which archives daily snapshots of all of CRAN, from which a user can install packages with the usual `install.packages()` function  [@MRAN].
-Users can always override this default by passing any current CRAN repository explicitly.
-Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule.
-Thus, Bioconductor packages can be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.
+**MRAN archives**: To facilitate installation of only contemporaneous
+versions of R packages on these images, the default CRAN mirror from which
+to install R packages is fixed to a snapshot of CRAN corresponding to
+the last date for which that version of R was the latest release. These
+snapshots are provided by the MRAN archive created by Revolution
+Analytics (now aquired by Microsoft), which archives daily snapshots
+of all of CRAN, from which a user can install packages with the usual
+`install.packages()` function  [@MRAN]. Users can always override this
+default by passing any current CRAN repository explicitly.  Unlike CRAN,
+Bioconductor only updates its repositories through bi-annual releases
+aligned to R's Spring release schedule. Thus, Bioconductor packages
+can be installed in the usual way using `bioclite`, which automatically
+selects the Bioconductor release corresponding to the version of R in use.
 
-**Version tags**:  These version tags are propegated throughout this stack: for instance, `rocker/tidyverse:devel` will provide the currently released versions of the R packages in the `tidyverse` [@tidyverse] installed on the nightly build of R-devel.
-Developers building packages on this stack are encouraged to tag their images accordingly as well.
-Table 3 indicates which versions of R are currently available in the stack, going back to `3.1.0`.
-While older versions may be added to the stack at a later date, we note that the MRAN snapshots began in starting 2014-09-17 and thus go back only to R `3.1` era.
-Each tag must be built from a separate Dockerfile, enabling minor differences in the build instructions to accomidate changing dependencies.
-Dockerfiles for past versions (e.g.
-< `3.4.0` currently) are intended to remain static over the long term, while the tag for the current version, `latest`, and `devel` may be tweaked to accomidate new features or dependencies.
+**Version tags**:  These version tags are propegated throughout this
+stack: for instance, `rocker/tidyverse:devel` will provide the currently
+released versions of the R packages in the `tidyverse` [@tidyverse]
+installed on the nightly build of R-devel.  Developers building packages
+on this stack are encouraged to tag their images accordingly as well.
+Table 3 indicates which versions of R are currently available in the
+stack, going back to `3.1.0`.  While older versions may be added to
+the stack at a later date, we note that the MRAN snapshots began in
+starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
+be built from a separate Dockerfile, enabling minor differences in the
+build instructions to accomidate changing dependencies.  Dockerfiles for
+past versions (e.g. < `3.4.0` currently) are intended to remain static
+over the long term, while the tag for the current version, `latest`,
+and `devel` may be tweaked to accomidate new features or dependencies.
 
-**Installation**: In this stack, the desired version of R is always built directly from source rather than the `apt` repositories.
-Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack.
-Version tags `3.3.3` and older are based on the Debian 8.0 release, codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much newer versions of common system dependencies and compilers.
-Dependencies needed to compile R that are not required at runtime are removed once R is installed, keeping the base images light-weight for faster download times.
-While most system dependencies required by common R packages can still be installed from the `apt` repositories, occassionally a more recent version must be compiled from source (e.g.
-the Gibbs Sampling library, JAGS, and the geospatial toolkit GDAL, must both be compiled from source on `debian:jessie` images).
-In this stack, users should never install R packages, using `apt`, since this will install a second, older version of R from the Debian repositories and a dated version of the R package (since any `r-cran-pkgname` package in the Debian repositories will depend on `r-base` in `apt` as well).
+**Installation**: In this stack, the desired version of R is always built
+directly from source rather than the `apt` repositories.  Compilers and
+dependencies are still installed from the stable `apt` repositories,
+and thus lag behind the more recent versions found in the `testing`
+stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
+codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on
+Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
+newer versions of common system dependencies and compilers.  Dependencies
+needed to compile R that are not required at runtime are removed once R is
+installed, keeping the base images light-weight for faster download times.
+While most system dependencies required by common R packages can still
+be installed from the `apt` repositories, occassionally a more recent
+version must be compiled from source (e.g. the Gibbs Sampling library,
+JAGS, and the geospatial toolkit GDAL, must both be compiled from source
+on `debian:jessie` images).  In this stack, users should never install R
+packages, using `apt`, since this will install a second, older version
+of R from the Debian repositories and a dated version of the R package
+(since any `r-cran-pkgname` package in the Debian repositories will
+depend on `r-base` in `apt` as well).
 
-**Build schedule**: All images are built automatically from their corresponding Dockerfiles (found in the GitHub repositories `rocker-org/rocker-versioned` and `rocker-org/geospatial`).
-A `cron` job sends nightly build triggers to Docker Hub to rebuild the `latest` and `devel` tagged images throughout the stack.
-To decrease load on the hub, build triggers for the numeric version tags are sent monthly.
-Although the Dockerfiles for older R versions installs an almost-identical software environment every time, the monthly rebuilding of these images on Docker Hub ensures they continue to recieve Debian security updates from upstream, and proves the build recipe still executes successfully.
-Note that rebuilding images with software from external repositories never produces a bitwise identical image, and thus the image identifier hash will change at each build.
+**Build schedule**: All images are built automatically from
+their corresponding Dockerfiles (found in the GitHub repositories
+`rocker-org/rocker-versioned` and `rocker-org/geospatial`). A `cron` job
+sends nightly build triggers to Docker Hub to rebuild the `latest` and
+`devel` tagged images throughout the stack.  To decrease load on the hub,
+build triggers for the numeric version tags are sent monthly. Although
+the Dockerfiles for older R versions installs an almost-identical
+software environment every time, the monthly rebuilding of these images
+on Docker Hub ensures they continue to recieve Debian security updates
+from upstream, and proves the build recipe still executes successfully.
+Note that rebuilding images with software from external repositories
+never produces a bitwise identical image, and thus the image identifier
+hash will change at each build.
 
@@
 
-In this stack, each image builds on the previous image, rather than all other images building directly on the base image, as in the `testing` stack.
-Table 2 lists the names and descriptions of the five images in this stack, along with image size and approximate download counts from Docker Hub.
-Sizes reflect (compressed) cumulative size: a user who has already downloaded the most recent version of `r-ver` and then pulls a copy of `rstudio` image will only need to download the additional 115 MB in the `rstudio` layers and not the full 334 MB listed.
-This linear design limits flexibility (no option for `tidyverse` without `rstudio`) but simplifies use and maintance.
-While no single environment will be optimal for everyone, both the packages selected in this stack and the stack ordering reflect considerable community input and tuning.
+In this stack, each image builds on the previous image, rather than all
+other images building directly on the base image, as in the `testing`
+stack. Table 2 lists the names and descriptions of the five images in
+this stack, along with image size and approximate download counts from
+Docker Hub.  Sizes reflect (compressed) cumulative size: a user who has
+already downloaded the most recent version of `r-ver` and then pulls a
+copy of `rstudio` image will only need to download the additional 115
+MB in the `rstudio` layers and not the full 334 MB listed.  This linear
+design limits flexibility (no option for `tidyverse` without `rstudio`)
+but simplifies use and maintance.  While no single environment will be
+optimal for everyone, both the packages selected in this stack and the
+stack ordering reflect considerable community input and tuning.
 
-The `rstudio` image includes a lightweight, easy-to-use and docker-friendly `init` system, s6 [@s6] for running persistant services, including the RStudio server.
-This system provides a convenient way for downstream Dockerfile developers to add additional persistent services (such as an ssh server) to a single container, or additional start-up or shutdown scripts that should be run when an instance starts up or shuts down.
-The `rstudio` image uses such a start-up script to configure user settings such as login password and permissions through environmental variables at run time, see *Using Rocker Images* below.
+The `rstudio` image includes a lightweight, easy-to-use and
+docker-friendly `init` system, s6 [@s6] for running persistant services,
+including the RStudio server.  This system provides a convenient way for
+downstream Dockerfile developers to add additional persistent services
+(such as an ssh server) to a single container, or additional start-up
+or shutdown scripts that should be run when an instance starts up or
+shuts down.  The `rstudio` image uses such a start-up script to configure
+user settings such as login password and permissions through environmental
+variables at run time, see *Using Rocker Images* below.
 
-The `tidyverse` image contains all required and suggested dependencies of the commonly used `tidyverse` and `devtools` R packages, including external database libraries (e.g.
-Maria-DB and Postgresql).
-Users should consult the package Dockerfiles or `installed.packages()` list directly for a complete list of installed packages.
-The `verse` library adds commonly used dependencies, notably a large but not comprehensive LaTeX environment and java development libraries.
-Previously the Rocker project provided the image `hadleyverse` which was since divided into `tidyverse` and `verse` through community input.
+The `tidyverse` image contains all required and suggested dependencies
+of the commonly used `tidyverse` and `devtools` R packages, including
+external database libraries (e.g. Maria-DB and Postgresql).  Users should
+consult the package Dockerfiles or `installed.packages()` list directly
+for a complete list of installed packages.  The `verse` library adds
+commonly used dependencies, notably a large but not comprehensive LaTeX
+environment and java development libraries.  Previously the Rocker
+project provided the image `hadleyverse` which was since divided into
+`tidyverse` and `verse` through community input.
 
@@
 
-Several images in the `rocker-versioned` stack can be customized on build when built locally (rather than pulling prebuilt images from Docker Hub) by using the `--build-arg` option of `docker build`.
-In the `r-ver` image, users can set `R_VERSION`, `BUILD_DATE` (MRAN default snapshot).
-In the `rstudio` image users can set `RSTUDIO_VERSION` (otherwise defaults to the most recent version), and the `PANDOC_TEMPLATES_VERSION` .
+Several images in the `rocker-versioned` stack can be customized on build
+when built locally (rather than pulling prebuilt images from Docker Hub)
+by using the `--build-arg` option of `docker build`.   In the `r-ver`
+image, users can set `R_VERSION`, `BUILD_DATE` (MRAN default snapshot).
+In the `rstudio` image users can set `RSTUDIO_VERSION` (otherwise defaults
+to the most recent version), and the `PANDOC_TEMPLATES_VERSION` .
 
@@
 
-While more detailed documentation on the use of Rocker images can be found on the Rocker wiki, we provide a conceptual overview here which illustrates common patterns and capabilities.
+While more detailed documentation on the use of Rocker images can be
+found on the Rocker wiki, we provide a conceptual overview here which
+illustrates common patterns and capabilities.
 
@@
 
-Users of the Docker's GUI interface, Kitematic who use the `rstudio`-derived images will automatically have their local kitematic directory linked to the `/home/rstudio/kitematic` directory on the Docker container.
+Users of the Docker's GUI interface, Kitematic who use the
+`rstudio`-derived images will automatically have their local kitematic
+directory linked to the `/home/rstudio/kitematic` directory on the
+Docker container.
 
-**Root vs user instances**: By default, all rocker containers run as root.
-This is consistent with standard Docker practice, allowing both interactive users and downstream Dockerfiles to install software directly without having to switch to root.
-The main snag of this approach comes if a user links a local volume and modifies files on that volume, which will result in the local file being owned by `root` and not the default user.
-To avoid this, specify the non-root user at runtime (username `docker` on the `r-base` stack, or `rstudio` on images deriving from rstudio) when running an interactive shell.
-For users accessing R through RStudio, this is not necessary.
-The docker container will run as root, but a user logs in through the RStudio server web interface as the non-root user "rstudio", and thus any changes made to linked volumes will not alter file permissions on the home directory.
-For more details on user permissions, see the Rocker wiki.
+**Root vs user instances**: By default, all rocker containers run as
+root. This is consistent with standard Docker practice, allowing both
+interactive users and downstream Dockerfiles to install software directly
+without having to switch to root.  The main snag of this approach comes if
+a user links a local volume and modifies files on that volume, which will
+result in the local file being owned by `root` and not the default user.
+To avoid this, specify the non-root user at runtime (username `docker`
+on the `r-base` stack, or `rstudio` on images deriving from rstudio)
+when running an interactive shell.  For users accessing R through RStudio,
+this is not necessary.  The docker container will run as root, but a user
+logs in through the RStudio server web interface as the non-root user
+"rstudio", and thus any changes made to linked volumes will not alter file
+permissions on the home directory.  For more details on user permissions,
+see the Rocker wiki.
 
@@
 
-Often Rocker images are not deployed directly, but used as a base image onto which users add software for their own needs in a custom Dockerfile.
-We offer a few recommendations here to streamline writing these extensions and promote best practices.
+Often Rocker images are not deployed directly, but used as a base image
+onto which users add software for their own needs in a custom Dockerfile.
+We offer a few recommendations here to streamline writing these extensions
+and promote best practices.
 
@@
 
- To streamline the installation of R packages in downstream Dockerfiles, the `r-base` image provides `littler` package [@littler] and associated `install2.r` script, enabling users to install additional packages with a command such as:
+ To streamline the installation of R packages in downstream Dockerfiles,
+ the `r-base` image provides `littler` package [@littler] and associated
+ `install2.r` script, enabling users to install additional packages with
+ a command such as:
 
@@
 
-There are several different but complementary reasons for creating the containerized R environments provided by
-the Rocker project, including convenience, community input, portability and scaling, reproducibility, and instruction. 
-We describe these in turn before presenting the commands required to deploy these environments in different use cases.
+There are several different but complementary reasons for creating the
+containerized R environments provided by the Rocker project, including
+convenience, community input, portability and scaling, reproducibility,
+and instruction.  We describe these in turn before presenting the commands
+required to deploy these environments in different use cases.
 
@@
 
-One common use case for Rocker containers is to provide a fast and reliable mechanism to deploy an R envionment on a remote server, such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's Jetstream servers, or private or institutional server hardware.
-Being able to test code in a predictable, pre-configured R environment on a local machine and then run the same code in an identical environment on a remote server (e.g.
-for access to greater RAM, more processors, or merely to free up the local machine from a long-running computation) is essential for low-friction scaling of analysis.
-Without such contanerization, getting code to run appropriately in a remote environment can be a major undertaking, requiring both time and knowledge many would-be users may not have.
+One common use case for Rocker containers is to provide a fast and
+reliable mechanism to deploy an R envionment on a remote server, such
+as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
+Jetstream servers, or private or institutional server hardware.  Being
+able to test code in a predictable, pre-configured R environment on a
+local machine and then run the same code in an identical environment on a
+remote server (e.g. for access to greater RAM, more processors, or merely
+to free up the local machine from a long-running computation) is essential
+for low-friction scaling of analysis.  Without such contanerization,
+getting code to run appropriately in a remote environment can be a major
+undertaking, requiring both time and knowledge many would-be users may
+not have.
 
-The following code installs Docker on almost any Linux-based server and launches a Rocker container providing the RStudio-server environment over a web interface.
+The following code installs Docker on almost any Linux-based server and
+launches a Rocker container providing the RStudio-server environment
+over a web interface.
 
-```
     wget -qO- https://get.docker.com/ | sh 
     sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
-```
 
-After running these two lines, the user may log into the server merely by pasting its IP address or DNS name into a browser and entering their password for user `rstudio`.
-This provides the user with a familiar, interactive environment running on a remote machine while requiring a minimum of expertise (with many cloud providers, these two lines of code can be run in a startup script without ever logging into the server over ssh).
-Some caveats and additional configuration options are discussed in the example section below.
-The user can select a Rocker container which contains the required software pre-installed, such as `rocker/tidyverse`, which provides the popular `tidyverse` suite of R packages \citep{tidyverse}, or even use their own containers which can extend the Rocker containers, as discussed below.
-This workflow saves considerable time relative to installing each package from scratch, which can take over an hour to reconstruct the environment provided by some rocker containers.
-As many commericial cloud providers charge by the hour, a fast and simple mechanism to deploy the desired environment in the cloud is crucial to making this approach practical for more exploratory analyses.
+After running these two lines, the user may log into the server merely
+by pasting its IP address or DNS name into a browser and entering their
+password for user `rstudio`.  This provides the user with a familiar,
+interactive environment running on a remote machine while requiring
+a minimum of expertise (with many cloud providers, these two lines
+of code can be run in a startup script without ever logging into the
+server over ssh).  Some caveats and additional configuration options are
+discussed in the example section below.  The user can select a Rocker
+container which contains the required software pre-installed, such as
+`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
+packages \citep{tidyverse}, or even use their own containers which can
+extend the Rocker containers, as discussed below.  This workflow saves
+considerable time relative to installing each package from scratch,
+which can take over an hour to reconstruct the environment provided by
+some rocker containers.  As many commericial cloud providers charge by
+the hour, a fast and simple mechanism to deploy the desired environment
+in the cloud is crucial to making this approach practical for more
+exploratory analyses.
 
@@
 
-Unlike virtual machine images, multiple such Docker containers can be run on the same server in the same way (though unique TCP ports must be assigned to each if they are accessed through the RStudio web interface).
-More importantly, it is much easier to test software locally using the identical Docker container environment before deploying on a remote server than it is to run a virtual machine such as an Amazon Machine Image (AMI) locally, or to reproduce the runtime environment identically on local and server systems.
+Unlike virtual machine images, multiple such Docker containers can
+be run on the same server in the same way (though unique TCP ports
+must be assigned to each if they are accessed through the RStudio web
+interface). More importantly, it is much easier to test software locally
+using the identical Docker container environment before deploying on
+a remote server than it is to run a virtual machine such as an Amazon
+Machine Image (AMI) locally, or to reproduce the runtime environment
+identically on local and server systems.
 
-Rocker images may be used for quickly deploying R code on remote servers without the use of the RStudio interface, running R in terminal mode over an `ssh` connection.
-As with any interactive docker instance, users should specify the interactive (`-i`) and terminal (`-t`, here combined with interactive as `-ti`) flags, and specify the desired executable environment (e.g.
-`R`, though other common options may be `Rscript` or `bash`):
+Rocker images may be used for quickly deploying R code on remote servers
+without the use of the RStudio interface, running R in terminal mode
+over an `ssh` connection. As with any interactive docker instance, users
+should specify the interactive (`-i`) and terminal (`-t`, here combined
+with interactive as `-ti`) flags, and specify the desired executable
+environment (e.g. `R`, though other common options may be `Rscript` or
+`bash`):
 
@@
 
-This example shows the use of the `--rm` flag to indicate that the container should be removed when the interactive session is finished.
+This example shows the use of the `--rm` flag to indicate that the
+container should be removed when the interactive session is finished.
 
-Another illustration of this portability is with regard to Continuous Integration servers such as Travis <https://travis.com> or Circle-CI <https://circle-ci.com>.
-While support for the R language on Travis has greatly improved recently (\citep{r-travis}), very customized environments can still be difficult to run (particularly for testing scripts not laid out in the R package structure).
+Another illustration of this portability is with regard to Continuous
+Integration servers such as Travis <https://travis.com> or Circle-CI
+<https://circle-ci.com>.  While support for the R language on Travis has
+greatly improved recently (\citep{r-travis}), very customized environments
+can still be difficult to run (particularly for testing scripts not laid
+out in the R package structure).
 
@@
 
-Installation of certain R packages can be challenging and non-intuitive, particularly when those packages require specific external dependencies.
-By running R through a Rocker image, users can deploy an R environment where their desired packages and libraries have already been installed and configured.
-As we discuss below, this configuration benefits from the input of a large community of users, helping to ensure better and more reliable performance.
-For instance, one example of such a configuration is in building and testing R packages using compiler sanatizers \citep{edd_sanitizers}, which requires R itself be re-compiled with specialized settings not avaialable or familiar to most R users.
-Using the Rocker images `rocker/r-devel-san` and `rocker/r-devel-ubsan-clang` provides a fast and simple way for users to test
+Installation of certain R packages can be challenging and non-intuitive,
+particularly when those packages require specific external
+dependencies. By running R through a Rocker image, users can deploy an
+R environment where their desired packages and libraries have already
+been installed and configured. As we discuss below, this configuration
+benefits from the input of a large community of users, helping to
+ensure better and more reliable performance. For instance, one example
+of such a configuration is in building and testing R packages using
+compiler sanatizers \citep{edd_sanitizers}, which requires R itself
+be re-compiled with specialized settings not avaialable or familiar
+to most R users.  Using the Rocker images `rocker/r-devel-san` and
+`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
+to test
 
-Many users are reluctant to upgrade their suite of installed packages, which may break their existing code or even their R environment if the installation goes poorly.
-However, upgrading packages and/or the R environment is often necessary to run analysis of a colleague or access more recent methods.
-Using Rocker, users can run the latest or specific earlier versions of R and R packages in an isolated environment without impacting their local system.
+Many users are reluctant to upgrade their suite of installed packages,
+which may break their existing code or even their R environment if the
+installation goes poorly.  However, upgrading packages and/or the R
+environment is often necessary to run analysis of a colleague or access
+more recent methods. Using Rocker, users can run the latest or specific
+earlier versions of R and R packages in an isolated environment without
+impacting their local system.
 

### 2017-07-10 manuscript.Rmd
first (incomplete) pass with typo edits
minimal style / wording / content changes

@@
   - name: Dirk Eddelbuettel
-    affiliation: Debian and R Projects
+    affiliation: Debian and R Projects; Ketchum Trading
     address: 
-    - line 1
-    - line 2
+    - Chicago, IL, USA
     email:  edd@debian.org
 abstract: >
-  We describe the Rocker project, which provides a widely used suite of Docker images
+  We describe the Rocker project, which provides a widely-used suite of Docker images
   with customized R environments for particular tasks. We discuss how this suite is
@@
 and run software applications using *containers*.  Containers provide a
-virtual environment requiring everything an application needs to run:
+virtual environment requiring all operating-system components an application needs to run:
 code, runtime, system tools, system runtime.  Docker containers are
-lightweight (sharing the operating system kernel, starting instantly,
-using a layered filesystem to minimize disk footprint and download time),
-built on open standards that run on all major platforms (Linux, Mac,
-Windows) and provide an added layer of security by running an application
+lightweight as they share the operating system kernel, starting instantly,
+using a layered filesystem which minimizes disk footprint and download time,
+and are built on open standards that run on all major platforms (Linux, Mac,
+Windows), and provide an added layer of security by running an application
 in an isolated environment \citep{what-docker}.
@@
 under the organization `rocker-org`, <https://github.com/rocker-org>.
-The issues tracker and pull requests are used to provide community input,
+The issues tracker and pull requests are used for community input,
 disucssions and contributions to these images. The rocker project wiki
@@
 
-The Rocker Project aims to provide a small core if Docker images that
+The Rocker Project aims to provide a small core of Docker images that
 serve as convenient 'base' images on which other users can build custom
@@
 cases (discussed in more detail below) against the overarching goals
-of creating images that are not sufficiently light-weight, easy to use
+of creating images that are still sufficiently light-weight, easy to use
 and easy to maintain is a difficult art.  The implementation in both
@@
 All Rocker images are based on the Debian Linux Distribution.
-The Debian platform provides a small base image, the well known `apt`
+The Debian platform provides a small base image, the well-known `apt`
 package management system and rich ecosystem of software libraries,
@@
 from binaries, since the package manager (`apt`) manages the necessary
-dependencies and it bypasses the time-consuming process of compiling
+(binary) dependencies and bypasses the time-consuming process of compiling
 from source.  Basing this stack on `debian:testing` means that much more
-recent versions of commonly used libraries and compilers are available
+recent versions of commonly-used libraries and compilers are available
 as binaries than would be found in a Debian stable release.  In order
@@
 popular R packages can also be installed through the package manager,
-e.g. `apt-get install r-cran-XML`.  This can be particularly helpful
+e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
 for packages with external system dependencies (such as `libxml2-dev`
-in this example) which cannot be installed from the R console.
+in this example) which cannot be installed from the R console as they 
+are system dependencies rather than R packages installed from within R.
 
 As the names `testing` and `unstable` imply, this approach is not
-without a downside.  The particular version of any given package can
-change as packages move from `unstable` into `testing` and newer versions
+without challenges.  The particular version of any given package can
+change as packages move from `unstable` into `testing`. New versions
 are sent to `unstable` during the normal course of Debian development.
@@
 4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
+<!-- DE: Are you sure about that point 4 ? -->
 

### 2017-07-10 manuscript.Rmd
WIP

--- 2017-06-23 manuscript.Rmd
@@
    \usepackage{longtable}
-   \usepackage{svg}
 output: rticles::rjournal_article
 ---
+
+<!-- Notes to Dirk: 
+     Some of this material may be more appropriate for documentation elsewhere, appearing here in only abbreviated form.  
+     To get started, I've just tried to err on the side of comprehensive, particularly since our documentation in places is sparse
+     or dated.  Feel free to flag any areas that should be moved in part or whole to documentation.  -->
 
@@
 
-## Using Rocker images
+## Design principles & use cases
+
+- Portable: from laptop to cloud
+- Sandboxed: experiment without touching local system
+- Optimized: community-tweaked configurations
+- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
+- Extensible: interactively and persistent community extensions
+- Versioned:
+
+research use, instructional use
+
+### Portability: From laptop to cloud
+
+One common use case for Rocker containers is to provide a fast and
+reliable mechanism to deploy a custom R envionment to a remote server, such
+as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
+Jetstream servers, or private or institutional server hardware.  Being
+able to test code in a predictable, pre-configured R environment on a
+local machine and then run the same code in an identical environment on a
+remote server (e.g. for access to greater RAM, more processors, or merely
+to free up the local machine from a long-running computation) is essential
+for low-friction scaling of analysis.  Without such contanerization,
+getting code to run appropriately in a remote environment can be a major
+undertaking, requiring both time and knowledge many would-be users may
+not have.
+
+The following code installs Docker on almost any Linux-based server and
+launches a Rocker container providing the RStudio-server environment
+over a web interface.
+
+    wget -qO- https://get.docker.com/ | sh 
+    sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+
+After running these two lines, the user may log into the server merely
+by pasting its IP address or DNS name into a browser and entering the
+appropriate password.  This provides the user with a familiar,
+interactive environment running on a remote machine while requiring
+a minimum of expertise.
+
+The user can select a Rocker
+container which contains the required software pre-installed, such as
+`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
+packages \citep{tidyverse}, or even use their own containers which can
+extend the Rocker containers, as discussed below.  This workflow saves
+considerable time relative to installing each package from scratch,
+which can take over an hour to reconstruct the environment provided by
+some rocker containers.  As many commericial cloud providers charge by
+the hour, a fast and simple mechanism to deploy the desired environment
+in the cloud is crucial to making this approach practical for more
+exploratory analyses.
+
+Unlike virtual machine images, multiple such Docker containers can
+be run on the same server in the same way (though unique TCP ports
+must be assigned to each if they are accessed through the RStudio web
+interface). More importantly, it is much easier to test software locally
+using the identical Docker container environment before deploying on
+a remote server than it is to run a virtual machine such as an Amazon
+Machine Image (AMI) locally, or to reproduce the runtime environment
+identically on local and server systems.
+
+### Interfaces
+
+Rocker images may be used for quickly deploying R code on remote servers
+without the use of the RStudio interface, running R in terminal mode
+over an `ssh` connection. As with any interactive docker instance, users
+should specify the interactive (`-i`) and terminal (`-t`, here combined
+with interactive as `-ti`) flags, and specify the desired executable
+environment (e.g. `R`, though other common options may be `Rscript` or
+`bash`):
+
+    docker run --rm -ti rocker/tidyverse R
+
+This example shows the use of the `--rm` flag to indicate that the
+container should be removed when the interactive session is finished.
 
@@
 
-## Extending Rocker images
-
-Often Rocker images are not deployed directly, but used as a base image
-onto which users add software for their own needs in a custom Dockerfile.
-We offer a few recommendations here to streamline writing these extensions
-and promote best practices.
-
-- littler scripts
-
- To streamline the installation of R packages in downstream Dockerfiles,
- the `r-base` image provides `littler` package [@littler] and associated
- `install2.r` script, enabling users to install additional packages with
- a command such as:
-
-    install2.r pkg1 pgk2 pkg3 ...
-
-- Dockerfile best practices 
-- automated builds, build triggers 
- - version-stable builds
-
-## Use cases
-
-There are several different but complementary reasons for creating the
-containerized R environments provided by the Rocker project, including
-convenience, community input, portability and scaling, reproducibility,
-and instruction.  We describe these in turn before presenting the commands
-required to deploy these environments in different use cases.
-
-### Portability & cloud computing
-
-One common use case for Rocker containers is to provide a fast and
-reliable mechanism to deploy an R envionment on a remote server, such
-as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers, or private or institutional server hardware.  Being
-able to test code in a predictable, pre-configured R environment on a
-local machine and then run the same code in an identical environment on a
-remote server (e.g. for access to greater RAM, more processors, or merely
-to free up the local machine from a long-running computation) is essential
-for low-friction scaling of analysis.  Without such contanerization,
-getting code to run appropriately in a remote environment can be a major
-undertaking, requiring both time and knowledge many would-be users may
-not have.
-
-The following code installs Docker on almost any Linux-based server and
-launches a Rocker container providing the RStudio-server environment
-over a web interface.
-
-    wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
-
-After running these two lines, the user may log into the server merely
-by pasting its IP address or DNS name into a browser and entering their
-password for user `rstudio`.  This provides the user with a familiar,
-interactive environment running on a remote machine while requiring
-a minimum of expertise (with many cloud providers, these two lines
-of code can be run in a startup script without ever logging into the
-server over ssh).  Some caveats and additional configuration options are
-discussed in the example section below.  The user can select a Rocker
-container which contains the required software pre-installed, such as
-`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
-packages \citep{tidyverse}, or even use their own containers which can
-extend the Rocker containers, as discussed below.  This workflow saves
-considerable time relative to installing each package from scratch,
-which can take over an hour to reconstruct the environment provided by
-some rocker containers.  As many commericial cloud providers charge by
-the hour, a fast and simple mechanism to deploy the desired environment
-in the cloud is crucial to making this approach practical for more
-exploratory analyses.
-
-<!--
-FIXME describe other modes of running containers, e.g. R terminal
-
--->
-
-Unlike virtual machine images, multiple such Docker containers can
-be run on the same server in the same way (though unique TCP ports
-must be assigned to each if they are accessed through the RStudio web
-interface). More importantly, it is much easier to test software locally
-using the identical Docker container environment before deploying on
-a remote server than it is to run a virtual machine such as an Amazon
-Machine Image (AMI) locally, or to reproduce the runtime environment
-identically on local and server systems.
-
-Rocker images may be used for quickly deploying R code on remote servers
-without the use of the RStudio interface, running R in terminal mode
-over an `ssh` connection. As with any interactive docker instance, users
-should specify the interactive (`-i`) and terminal (`-t`, here combined
-with interactive as `-ti`) flags, and specify the desired executable
-environment (e.g. `R`, though other common options may be `Rscript` or
-`bash`):
-
-    docker run --rm -ti rocker/tidyverse R
-
-This example shows the use of the `--rm` flag to indicate that the
-container should be removed when the interactive session is finished.
-
-Another illustration of this portability is with regard to Continuous
-Integration servers such as Travis <https://travis.com> or Circle-CI
-<https://circle-ci.com>.  While support for the R language on Travis has
-greatly improved recently (\citep{r-travis}), very customized environments
-can still be difficult to run (particularly for testing scripts not laid
-out in the R package structure).
-
 ### Convenience
@@
 
+## Extending Rocker images
+
+Often Rocker images are not deployed directly, but used as a base image
+onto which users add software for their own needs in a custom Dockerfile.
+We offer a few recommendations here to streamline writing these extensions
+and promote best practices.
+
+- littler scripts
+
+ To streamline the installation of R packages in downstream Dockerfiles,
+ the `r-base` image provides `littler` package [@littler] and associated
+ `install2.r` script, enabling users to install additional packages with
+ a command such as:
+
+    install2.r pkg1 pgk2 pkg3 ...
+
+- Dockerfile best practices 
+- automated builds, build triggers 
+ - version-stable builds
+
 ### Reproducibility

### 2017-07-10 manuscript.Rmd
second pass with typo edits
minimal style / wording / content changes

@@
 few weeks).  The rest of the stack uses build triggers that rebuild the
-images either whenever `r-base` is updated or the Dockerfile sources are
+images either whenever `r-base` is updated, or the Dockerfile sources are
 updated on the corresponding GitHub repository.  The only exception in
@@
 
-<!-- Should we mention r-apt separately? -->
+<!-- CB: Should we mention r-apt separately? -->
+<!-- DE: Or maybe not at all. It is somewhat specialised. -->
 
-<!-- Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?  -->
+<!-- CB: Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?
+-->
+<!-- DE: Not sure. -->
 
@@
 This stack emphasizes stability and reproducibility of the Docker build.
-This stack was introduced much more recently, starting in November
-of 2016, in response to considerable user input and requests. The key
-feature of this stack is the ability to run older versions of R, along
+This stack was introduced much more recently (November 2016) 
+in response to considerable user input and requests. The key
+feature of this stack is the ability to run older versions of R along
 with the contemporaneous versions of R packages.  A user specifies the
@@
 snapshots are provided by the MRAN archive created by Revolution
-Analytics (now aquired by Microsoft), which archives daily snapshots
-of all of CRAN, from which a user can install packages with the usual
+Analytics (now part of Microsoft). It archives daily snapshots
+of all of CRAN from which a user can install packages with the usual
 `install.packages()` function  [@MRAN]. Users can always override this
@@
 over the long term, while the tag for the current version, `latest`,
-and `devel` may be tweaked to accomidate new features or dependencies.
+and `devel` may be tweaked to accomodate new features or dependencies.
 
@@
 depend on `r-base` in `apt` as well).
+<!-- DE: Likely a non-issue as /usr/local is always search before /usr. -->
 
@@
 downstream Dockerfile developers to add additional persistent services
-(such as an ssh server) to a single container, or additional start-up
+(such as an `ssh` server) to a single container, or additional start-up
 or shutdown scripts that should be run when an instance starts up or
@@
 of the commonly used `tidyverse` and `devtools` R packages, including
-external database libraries (e.g. Maria-DB and Postgresql).  Users should
+external database libraries (e.g. MariaDB and PostgreSQL).  Users should
 consult the package Dockerfiles or `installed.packages()` list directly
@@
 commonly used dependencies, notably a large but not comprehensive LaTeX
-environment and java development libraries.  Previously the Rocker
+environment and Java development libraries.  Previously the Rocker
 project provided the image `hadleyverse` which was since divided into
@@
 
-Installation of certain R packages can be challenging and non-intuitive,
+Installation of certain R packages can be challenging,
 particularly when those packages require specific external
@@
 `rocker/r-devel-ubsan-clang` provides a fast and simple way for users
-to test
+to test.
 

### 2017-07-14 manuscript.Rmd
wip

@@
 
-All Rocker images are based on the Debian Linux Distribution.
+All Rocker images are based on the Debian Linux distribution.
 The Debian platform provides a small base image, the well-known `apt`
@@
 within the R community, including an active `r-sig-debian` listserve.
-Unfortunately, the relatively long period between stable Debian releases
+The relatively long period between stable Debian releases
 (roughly two years recently) means that software in the `debian:stable`
@@
 to provide access to the most recent avialble binaries, this stack uses
-apt-pinning [@apt-pinning] to allow the `apt` package manager to also
+apt-pinning \citep{apt-pinning} to allow the `apt` package manager to also
 install binaries from `debian:unstable`, which represents the most recent,
@@
 are system dependencies rather than R packages installed from within R.
+We should note, however, that only about 500 of the currently almost
+11000 CRAN packages are available as Debian packages.
 
@@
 be installed from `testing`, or vice versa (using the `-t` option in
-`apt`, see Box 1 on apt-pinning.)
+`apt`, see Box 1 on apt-pinning.)  That said, packages only migrate from 
+`unstable` to `testing` after a period of several days---and if the 
+migration and installation of the particular version is free of 
+interactions with other packages in their dependency graph. That way,
+`unstable` serves as validation lab which leaves `testing` reasonably 
+stable yet current.
 
@@
 1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
+<!-- DE 1.) would be true of stable too, no?  -->
 2. These Dockerfiles always install the most recent available software
@@
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
-3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
+3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt-pinning}.
 4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
@@
 
-The `debian:testing`-based stack currently includes seven images actively
-maintained by the Rocker development team (Table 1).  `r-base` builds on
-`debian:testing`, and the other six in the stack each build directly from
-`r-base`.  The `r-base` image is unique in that it is designated as the
-official image for the R language by the Docker organization itself.
-This image is reviewed and built by employees of Docker Inc from the
-Dockerfile maintained by the Rocker team.  Consequently, users should
-refer to this image in Docker commands without an organization napespace,
-e.g. `docker pull r-base` to access the official image.  All other images
-in the Rocker project are not individually reviewed and built by Docker
-Inc and must be refered to require the organization namespace on Docker
-Hub, e.g. `docker pull rocker/r-devel`.
+The `debian:testing`-based stack currently includes seven images
+actively maintained by the Rocker development team (Table 1).
+`r-base` builds on `debian:testing`, and the other six in the stack
+each build directly from `r-base`.  The `r-base` image is unique in
+that it is designated as the official image for the R language by the
+Docker organization itself.  This official image is reviewed and then
+built by employees of Docker Inc based on the Dockerfile maintained by
+the Rocker team.  Consequently, users should refer to this image in
+Docker commands without an organization napespace, e.g. `docker pull
+r-base` to access the official image.  All other images in the Rocker
+project are not individually reviewed and built by Docker Inc and must
+be refered to require the organization namespace on Docker Hub,
+e.g. `docker pull rocker/r-devel`.
 
-Several of the images in this stack are oriented towards the R development
-community. `r-devel`, `drd`, `r-devel-san` & `r-devel-ubsan-clang` all
-add a copy of the development version of R side-by-side the current
-release of R provided by `r-base`.  On these images, the development
-version is aliased to `RD` to distinguish from the current release, `R`.
-As the names suggest, each provide slightly different configurations.
-Of particular interest are the images providing development R built with
-support for C/C++ address sanitizers, which are notoriously difficult
-to configure [@edd_sanitizers].
+Several of the images in this stack are oriented towards the R
+development community. `r-devel`, `drd`, `r-devel-san` &
+`r-devel-ubsan-clang` all add a copy of the development version of R
+side-by-side the current release of R provided by `r-base`.  On these
+images, the development version is aliased to `RD` to distinguish from
+the current release, `R`.  As the names suggest, each provide slightly
+different configurations.  Of particular interest are the images
+providing development R built with support for C/C++ address and
+undefined-behavior sanitizers, which are somewhat more difficult to
+configure \citep{nedd_sanitizers}.
 
@@
 updated on the corresponding GitHub repository.  The only exception in
-this stack is the `drd` image, which is rebuilt each week by cron trigger.
+this stack is the `drd` image, which is rebuilt each week by a `cron` trigger.
 
@@
 
-Table: the `debian:testing` image stack
+Table: The `debian:testing` image stack
 
@@
 wanting to create downstream Dockerfiles which are based on the current
-release at time but will continue to reconstruct the same environment in
-the future after newer R versions are released should explicilty include
+release at time (but will continue to reconstruct the same environment in
+the future after newer R versions are released) should explicilty include
 the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
@@
 Bioconductor only updates its repositories through bi-annual releases
-aligned to R's Spring release schedule. Thus, Bioconductor packages
+aligned to R's spring release schedule. Thus, Bioconductor packages
 can be installed in the usual way using `bioclite`, which automatically
@@
 
-**Version tags**:  These version tags are propegated throughout this
-stack: for instance, `rocker/tidyverse:devel` will provide the currently
+**Version tags**:  The version tags are propegated throughout this
+stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
 released versions of the R packages in the `tidyverse` [@tidyverse]
@@
 the stack at a later date, we note that the MRAN snapshots began in
-starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
+starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
 be built from a separate Dockerfile, enabling minor differences in the
-build instructions to accomidate changing dependencies.  Dockerfiles for
+build instructions to accomodate changing dependencies.  Dockerfiles for
 past versions (e.g. < `3.4.0` currently) are intended to remain static
@@
 The `tidyverse` image contains all required and suggested dependencies
-of the commonly used `tidyverse` and `devtools` R packages, including
+of the commonly-used `tidyverse` and `devtools` R packages, including
 external database libraries (e.g. MariaDB and PostgreSQL).  Users should
@@
 for a complete list of installed packages.  The `verse` library adds
-commonly used dependencies, notably a large but not comprehensive LaTeX
+commonly-used dependencies, notably a large but not comprehensive LaTeX
 environment and Java development libraries.  Previously the Rocker
@@
 - Extensible: interactively and persistent community extensions
-- Versioned:
-
-research use, instructional use
+- Versioned: research use, instructional use
 
@@
 able to test code in a predictable, pre-configured R environment on a
-local machine and then run the same code in an identical environment on a
+local machine and to then run the same code in an identical environment on a
 remote server (e.g. for access to greater RAM, more processors, or merely
@@
 permissions on the home directory.  For more details on user permissions,
-see the Rocker wiki.
+see the Rocker wiki. [@rockerwiki]
 
@@
 of such a configuration is in building and testing R packages using
-compiler sanatizers \citep{edd_sanitizers}, which requires R itself
+compiler sanitizers \citep{edd_sanitizers}, which requires R itself
 be re-compiled with specialized settings not avaialable or familiar

### 2017-07-18 manuscript.Rmd
WIP

--- 2017-07-10 manuscript.Rmd
@@
 user settings such as login password and permissions through environmental
-variables at run time, see *Using Rocker Images* below.
+variables at run time.  <!-- link wiki howto page? -->
 
@@
 
-<!--
-This stack also makes use of Docker metadata labels: <http://schema-label.org> metadata for `license`, `vcs-url` (GitHub repository), and `vendor`.
-These can be altered or extended in downstream images.
--->
+This stack also makes use of Docker metadata labels defined by
+<http://schema-label.org>,
+indicating image `license` (GPL-2.0), `vcs-url` (GitHub repository), and `vendor` (Rocker Project).  These metadata can be altered or extended in downstream images.
 
@@
 
+<!-- 
+Use cases: instruction, development, 
+-->
+
 ## Design principles & use cases
 
-- Portable: from laptop to cloud
-- Sandboxed: experiment without touching local system
-- Optimized: community-tweaked configurations
-- Interfaces: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
-- Extensible: interactively and persistent community extensions
-- Versioned:
-
-research use, instructional use
+- **Portable**: from laptop to cloud
+- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
+- **Sandboxed**: experiment without touching local system
+- **Optimized**: community-tweaked configurations
+- **Extensible**: interactively and persistent community extensions
+- **Versioned**: reproducible environments
 
@@
 as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers, or private or institutional server hardware.  Being
+Jetstream servers [@jetstream], or private or institutional server hardware.  Being
 able to test code in a predictable, pre-configured R environment on a
@@
 
-The following code installs Docker on almost any Linux-based server and
+<!-- comment on use in instructional context -->
+This portablility is similarly valuable in an instructional context. Requiring students
+to install all necessary software on personal laptops can be particularly challenging
+for short workshops, where download and installation time and troubleshooting across
+heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
+By deploying a Rocker image or Rocker-derived image (see extensibility) on a cloud machine,
+an instructor can easily provide all students access to the pre-configured software environment
+using only the browser on their laptops.  Unlike a vendor-specific virtual machine image (e.g. Amazon Machine Images, AMIs),
+this approach does not preclude students being able to run the same software locally later on.
+After installing Docker for their machine they too can run the course-specific software environment
+easily on their own laptops and deploy it on other cloud providers.  This strategy has proven
+effective in our own experience in both workshops and semester-length courses. Similar Docker-based
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke [@Mine] and
+UC Berkeley [@data8].
+
+<!-- NOTE: Not sure if we should include such example commands, may be best left to the documentation? -->
+
+For instance, the following code installs Docker on almost any Linux-based server and
 launches a Rocker container providing the RStudio-server environment
@@
 
+Many academic and commerical cloud providers make it
+possible to execute such code snippets when an instance is launched, without
+ever needing to `ssh` into the machine.
 After running these two lines, the user may log into the server merely
@@
 
-The user can select a Rocker
-container which contains the required software pre-installed, such as
-`rocker/tidyverse`, which provides the popular `tidyverse` suite of R
-packages \citep{tidyverse}, or even use their own containers which can
-extend the Rocker containers, as discussed below.  This workflow saves
-considerable time relative to installing each package from scratch,
-which can take over an hour to reconstruct the environment provided by
-some rocker containers.  As many commericial cloud providers charge by
-the hour, a fast and simple mechanism to deploy the desired environment
-in the cloud is crucial to making this approach practical for more
-exploratory analyses.
-
-Unlike virtual machine images, multiple such Docker containers can
-be run on the same server in the same way (though unique TCP ports
-must be assigned to each if they are accessed through the RStudio web
-interface). More importantly, it is much easier to test software locally
-using the identical Docker container environment before deploying on
-a remote server than it is to run a virtual machine such as an Amazon
-Machine Image (AMI) locally, or to reproduce the runtime environment
-identically on local and server systems.
-
 ### Interfaces
 
-Rocker images may be used for quickly deploying R code on remote servers
-without the use of the RStudio interface, running R in terminal mode
-over an `ssh` connection. As with any interactive docker instance, users
+An important aspect of the Rocker project design is the ability
+for users to interact with the software on the container through either
+an interactive shell session (such as the R shell or a bash shell), or
+through a web browser accessing the RStudio\textsuperscript{\textregistered} Server
+integrated development environment (IDE).  Traditional remote and high-performance
+computing workflows for R users have usually required the use of `ssh` and 
+a terminal only interface, posing a challenge for interactive graphics 
+and a barrier to users unfamiliar with these tools and environments. Accessing
+an RStudio instance through the browser removes these barriers.  Rocker
+images include the RStudio-server software pre-installed and configured with
+the explicit permission of RStudio Inc.
+
+Users of the `rstudio` instances can access a root `bash` shell in an 
+instance running the container using
+
+```
+docker exec -ti <container-id> bash
+```
+
+which can be useful for administrative tasks such as installing system dependencies.
+All Rocker images can also be run as an interactive R, RScript or bash shell without running
+RStudio, which can be useful for batch jobs or for anyone who prefers that environment.
+
+<!--  Reserve these details for documentation, reference here? --> 
+As with any interactive docker instance, users
 should specify the interactive (`-i`) and terminal (`-t`, here combined
@@
 
-While more detailed documentation on the use of Rocker images can be
-found on the Rocker wiki, we provide a conceptual overview here which
-illustrates common patterns and capabilities.
-
-**Interactive shell**:
-
-**RStudio instance**:
-
-**Linking local volumes**: Docker
+**Linking local volumes**
 
@@
 
+<!-- Blah blah blah ... link to documentation instead(?) -->
 **Root vs user instances**: By default, all rocker containers run as
@@
 
-### Convenience
+### Sandboxed
 
-Installation of certain R packages can be challenging and non-intuitive,
-particularly when those packages require specific external
-dependencies. By running R through a Rocker image, users can deploy an
-R environment where their desired packages and libraries have already
-been installed and configured. As we discuss below, this configuration
-benefits from the input of a large community of users, helping to
-ensure better and more reliable performance. For instance, one example
-of such a configuration is in building and testing R packages using
-compiler sanatizers \citep{edd_sanitizers}, which requires R itself
-be re-compiled with specialized settings not avaialable or familiar
-to most R users.  Using the Rocker images `rocker/r-devel-san` and
-`rocker/r-devel-ubsan-clang` provides a fast and simple way for users
-to test
-
-Many users are reluctant to upgrade their suite of installed packages,
+Another feature of Rocker instances is the ability to provide a sandboxed 
+environment, isolated from software and potentially from other data on
+the machine.  Many users are reluctant to upgrade their suite of installed packages,
 which may break their existing code or even their R environment if the
@@
 environment is often necessary to run analysis of a colleague or access
-more recent methods. Using Rocker, users can run the latest or specific
-earlier versions of R and R packages in an isolated environment without
-impacting their local system.
+more recent methods.  Rocker instances offer an easy solution. For 
+instance, a user can run R code requiring the most recent versions of R
+and related packages inside a Rocker container without having to upgrade their
+local installations first.  Conversely, one could use Rocker to run code
+on an older R release with prior versions of R packages, again without
+having to make any alteration to one's local R install.  Another common
+use case is to access an instance of R with support for particular
+options such as using gcc or clang compiler sanatizers \citep{edd_sanitizers}.
+These requires R itself be built with specialized settings that may not be
+not available or familiar to many R users on their native system, but can
+be easily deployed by pulling the Rocker images `rocker/r-devel-san` or
+`rocker/r-devel-ubsan-clang`.
 
-### Extensible and Community-optimized images
+This sandboxing feature of Docker containers is also valuable in the 
+remote computing context, where it can allow system administrators to
+provide users with administrative (root) privilages to alter and install
+software within a container without altering the computational environment
+of other users on the remote machine.  Unlike traditional virtual machines,
+these containers do not impose a large footprint of reserved resources --
+a typical host can easily support 100s of containers [@docker-vs-vm].
 
-Rocker images benefit from widespread use, testing and contributions from many members of the R community.
+<!-- volume linking isolation -->
 
-## Extending Rocker images
+### Community Optimized
+
+Having a shared, transparent computational environment created by a publicly hosted, reproducible
+recipe facilitates community input into configuration details. R and many of its packages
+and related software can be configured with a wide range of options, compilers, different
+linear-algebra libraries and so forth.  While this flexibility reflects varying needs,
+many users rely on default settings which may be optimized more for simplicity of installation
+than performance.  The Rocker recipes reflect significant community input on these choices,
+as well as the considerable experience and expertise of the R Debian mantainer of 20 years
+in what configuration options to use.  This helps create a more finely tuned, optimized reference
+implementation of the R environment as well as a platform for comparing and discussing these 
+issues which are often overlooked elsewhere.  Issues and Pull Requests on the Rocker 
+repositories on GitHub attest to these discussions and improvements.
+Widespread use of the Rocker image helps promote both testing of these choices and 
+contributions further tweaking the configuration from many members of the R community.
+
+### Versioned
+
+[@Boettiger2015]
+
+### Extensible
+
+Rocker images are designed to provide
 
@@
 
-### Reproducibility
-
-\citep{Boettiger2015}
-
-### Instruction
-
-- Table of main repos, links to Docker-Hub & GitHub addresses
-
-- list/table of topics on the wiki
+## Conclusions
 

### 2017-07-19 manuscript.Rmd
WIP

@@
 
+<!--  The main point here is that these are generic features that could be goals 
+      of any effort to provide researchers with virtual computing environments.
+
+      While most virtual computing environments (AMI images, scientific cloud
+      platforms like cyverse, etc) provide some subset of these, they are most
+      effective in concert.  -->
+
 - **Portable**: from laptop to cloud
@@
 - **Sandboxed**: experiment without touching local system
+- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
 - **Optimized**: community-tweaked configurations
@@
 not have.
+
+Perhaps more importantly, Rocker containers are just as easy to run
+on locally on most modern laptops on Windows, MacOS, and Linux-based
+operating systems.
 
@@
 
+### Transparent
+
+Users can easily determine the software stack installed on any Rocker image
+by examining the associated Dockerfile recipe, which provides a consise, human-readable
+record of the installation.
+All Rocker images use automated builds through DockerHub, which also acts as the central,
+default repository distributing the images.  Using automated builds rather than uploading
+pre-built image binaries to the Docker Hub avoids the potential for the build
+not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
+and in the linked GitHub repository of the Rocker project (<https://github.com/rocker-org), 
+which provides a transparent versioned history of all changes made to these recipes,
+as well as documentation, a community wiki, and issues trackers for discussing proposed
+changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
+
 ### Community Optimized
@@
 
-Rocker images are designed to provide
+<!-- Not intended as a guide to writing Dockerfiles, see http://ropenscilabs.github.io/r-docker-tutorial/05-dockerfiles.html -->
 
-Often Rocker images are not deployed directly, but used as a base image
-onto which users add software for their own needs in a custom Dockerfile.
-We offer a few recommendations here to streamline writing these extensions
-and promote best practices.
+Users can easily add additional software to any running Rocker image using
+the standard R and Debian mechanisms. To make these changes persistent, Rocker 
+images can easily be extended by any user writing their own Dockerfiles based
+on an appropriate Rocker image.
 
-- littler scripts
+Rocker images provide a few utility functions to streamline this process, including the 
+`littler` [@littler] scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.
 
- To streamline the installation of R packages in downstream Dockerfiles,
- the `r-base` image provides `littler` package [@littler] and associated
- `install2.r` script, enabling users to install additional packages with
- a command such as:
+    RUN install2.r pkg1 pgk2 pkg3 ...
 
-    install2.r pkg1 pgk2 pkg3 ...
+Users building on the version-tagged images will by default use the MRAN snapshot mirror
+associated with the most recent date for which that image was current.  This ensures that
+a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
+available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  (This default
+can of course be overwritten by specifying a different CRAN mirror explicitly.)
 
-- Dockerfile best practices 
-- automated builds, build triggers 
- - version-stable builds
+Users writing their own Dockerfiles are encouraged to follow the same practices as the
+Rocker Project, such as the [Dockerfile Best Practices](), the use of automated builds,
+and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
+every time the relevant upstream DOckerfile is updated by using build triggers on Docker Hub.
+
+Note that users can also preserve changes to Rocker images that they have modified interactively using
+the `docker commit` mechanism, which creates a new binary image which can be pushed back to
+a personal account on the Docker Hub.  While this is sometimes convenient, we encourage users to
+consider writing Dockerfiles instead whenever possible, as this creates a more transparent
+and reproducible mechanism to accomplish the same thing.
+
+- Dockerfiles. community input.  [`hlapp/rpopgen`](https://github.com/NESCent/popgen-docker)
 

### 2017-07-19 manuscript.Rmd

@@
 1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
+<!-- DE 1.) would be true of stable too, no?   -->
+<!-- CB  those sources are sometimes too old; e.g. jags sources on jessie are incompatible with rjags, etc -->
 2. These Dockerfiles always install the most recent available software
@@
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
-3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility [@apt-pinning].
-4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team.
+3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt-pinning}.
+4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team. [@debian-security]
 <!-- DE: Are you sure about that point 4 ? -->
+<!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->
 
@@
 few weeks).  The rest of the stack uses build triggers that rebuild the
-images either whenever `r-base` is updated or the Dockerfile sources are
+images either whenever `r-base` is updated, or the Dockerfile sources are
 updated on the corresponding GitHub repository.  The only exception in
-this stack is the `drd` image, which is rebuilt each week by cron trigger.
+this stack is the `drd` image, which is rebuilt each week by a `cron` trigger.
 
@@
 
-Table: the `debian:testing` image stack
+Table: The `debian:testing` image stack
 
-<!-- Should we mention r-apt separately? -->
+<!-- CB: Should we mention r-apt separately? -->
+<!-- DE: Or maybe not at all. It is somewhat specialised. +1 -->
 
-<!-- Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?  -->
+<!-- CB: Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?
+-->
+<!-- DE: Not sure. -->
 
@@
 the stack at a later date, we note that the MRAN snapshots began in
-starting 2014-09-17 and thus go back only to R `3.1` era.  Each tag must
+starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
 be built from a separate Dockerfile, enabling minor differences in the
-build instructions to accomidate changing dependencies.  Dockerfiles for
+build instructions to accomodate changing dependencies.  Dockerfiles for
 past versions (e.g. < `3.4.0` currently) are intended to remain static
 over the long term, while the tag for the current version, `latest`,
-and `devel` may be tweaked to accomidate new features or dependencies.
+and `devel` may be tweaked to accomodate new features or dependencies.
 
@@
 depend on `r-base` in `apt` as well).
+<!-- DE: Likely a non-issue as /usr/local is always search before /usr. -->
+<!-- Right, apparently RStudio manages to flub this and ignores PATH order when looking for R.
+     Anyway, much of this should probably be moved to documentation elsewhere -->
 
@@
 The `tidyverse` image contains all required and suggested dependencies
-of the commonly used `tidyverse` and `devtools` R packages, including
-external database libraries (e.g. Maria-DB and Postgresql).  Users should
+of the commonly-used `tidyverse` and `devtools` R packages, including
+external database libraries (e.g. MariaDB and PostgreSQL).  Users should
 consult the package Dockerfiles or `installed.packages()` list directly
 for a complete list of installed packages.  The `verse` library adds
-commonly used dependencies, notably a large but not comprehensive LaTeX
-environment and java development libraries.  Previously the Rocker
+commonly-used dependencies, notably a large but not comprehensive LaTeX
+environment and Java development libraries.  Previously the Rocker
 project provided the image `hadleyverse` which was since divided into
@@
 
-<!-- volume linking isolation -->
+<!-- volume linking isolation? -->
 

### 2017-07-19 manuscript.Rmd
WIP: a bit more text on the "extensible" section

@@
 
+Any portable computational environment faces an inevitable tension between
+the "kitchen sink problem" at one extreme, and the "discovery problem" on
+the other.  A kitchen sink image seeks to accomidate too many use cases
+in a single image.  Such images are inevitably very large and thus slow
+or difficult to deploy, maintain and optimize.  At the other extreme, providing
+too many specialized images makes it more difficult for a user to discover 
+the one they need.  The Rocker project seeks to avoid both of these problems
+by providing images that an be easily extended by individuals and communities.
+
 Users can easily add additional software to any running Rocker image using
 the standard R and Debian mechanisms. To make these changes persistent, Rocker 
-images can easily be extended by any user writing their own Dockerfiles based
-on an appropriate Rocker image.
+images can be extended by any user writing their own Dockerfiles based
+on an appropriate Rocker image.  Many users maintain their own Dockerfiles
+which add software specific to their own needs on top of an existing Rocker image.
+This provides a convenient way to reap all the benefits listed above in a 
+customized environment.
 
+Sharing these Dockerfiles can also facilitate the emergence 
+of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
+image emerged from the input of a number of Rocker users all adding common
+geospatial libraries and packages on top of the existing Rocker images. 
+This coalescence helped create a more fine-tuned image with broad support 
+for a wide range of commonly used data formats and libraries.  Other community
+images are developed and maintained independently of the Rocker Project,
+such as the `popgen` image of population-genetics-oriented software developed 
+by the National Evolutionary Synthesis Center (NESCent).
+
+<!-- see docs instead -->
 Rocker images provide a few utility functions to streamline this process, including the 
@@
 Users writing their own Dockerfiles are encouraged to follow the same practices as the
-Rocker Project, such as the [Dockerfile Best Practices](), the use of automated builds,
+Rocker Project, such as the Dockerfile Best Practices [@best-practices], the use of automated builds,
 and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
@@
 
-- Dockerfiles. community input.  [`hlapp/rpopgen`](https://github.com/NESCent/popgen-docker)
-
 ## Conclusions
 
+Lessons for related efforts
+
+Future directions, Open Container Initiative, Singularity & Shifter (support for Docker on HPC)
+
 \bibliography{RJreferences}

### 2017-09-13 manuscript.Rmd
update citation format

@@
 to provide access to the most recent avialble binaries, this stack uses
-apt-pinning \citep{apt-pinning} to allow the `apt` package manager to also
+apt-pinning \citep{apt_pinning} to allow the `apt` package manager to also
 install binaries from `debian:unstable`, which represents the most recent,
@@
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
-3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt-pinning}.
-4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team. [@debian-security]
+3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
+4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team \citep{debian_security}.
 <!-- DE: Are you sure about that point 4 ? -->
@@
 undefined-behavior sanitizers, which are somewhat more difficult to
-configure \citep{nedd_sanitizers}.
+configure \citep{edd_sanitizers}.
 
@@
 of all of CRAN from which a user can install packages with the usual
-`install.packages()` function  [@MRAN]. Users can always override this
+`install.packages()` function  \citep{MRAN}. Users can always override this
 default by passing any current CRAN repository explicitly.  Unlike CRAN,
@@
 stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
-released versions of the R packages in the `tidyverse` [@tidyverse]
+released versions of the R packages in the `tidyverse` \citep{tidyverse}
 installed on the nightly build of R-devel.  Developers building packages
@@
 The `rstudio` image includes a lightweight, easy-to-use and
-docker-friendly `init` system, s6 [@s6] for running persistant services,
+docker-friendly `init` system, s6 \citep{s6} for running persistant services,
 including the RStudio server.  This system provides a convenient way for
@@
 as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers [@jetstream], or private or institutional server hardware.  Being
+Jetstream servers \citep{jetstream}, or private or institutional server hardware.  Being
 able to test code in a predictable, pre-configured R environment on a
@@
 effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke [@Mine] and
-UC Berkeley [@data8].
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+UC Berkeley \citep{data8}.
 
@@
 permissions on the home directory.  For more details on user permissions,
-see the Rocker wiki. [@rockerwiki]
+see the Rocker website, <https://rocker-project.org>.
 
@@
 these containers do not impose a large footprint of reserved resources --
-a typical host can easily support 100s of containers [@docker-vs-vm].
+a typical host can easily support 100s of containers \cite{what-docker}.
 
@@
 
-[@Boettiger2015]
+\citep{@Boettiger2015}
 
@@
 Rocker images provide a few utility functions to streamline this process, including the 
-`littler` [@littler] scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.
+`littler` \citep{littler} scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.
 
@@
 Users writing their own Dockerfiles are encouraged to follow the same practices as the
-Rocker Project, such as the Dockerfile Best Practices [@best-practices], the use of automated builds,
+Rocker Project, such as the Dockerfile Best Practices \citep{best_practices}, the use of automated builds,
 and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
-every time the relevant upstream DOckerfile is updated by using build triggers on Docker Hub.
+every time the relevant upstream Dockerfile is updated by using build triggers on Docker Hub.
 

### 2017-09-19 manuscript.Rmd
tighten up text, move more technical docs to website

@@
 
-<!-- CB: Should we mention r-apt separately? -->
-<!-- DE: Or maybe not at all. It is somewhat specialised. +1 -->
-
-<!-- CB: Do we describe version tags of r-base, which do provide older versions of R but are based on Dockerfiles which will no longer build since the requested R version is no longer in `unstable`?
--->
-<!-- DE: Not sure. -->
-
 ### The `debian:stable`-based stack
@@
 depend on `r-base` in `apt` as well).
-<!-- DE: Likely a non-issue as /usr/local is always search before /usr. -->
-<!-- Right, apparently RStudio manages to flub this and ignores PATH order when looking for R.
-     Anyway, much of this should probably be moved to documentation elsewhere -->
 
@@
 latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.4.1   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
 3.4.0   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
@@
       platforms like cyverse, etc) provide some subset of these, they are most
-      effective in concert.  -->
+      effective in concert.
 
@@
 
+-->
+
 ### Portability: From laptop to cloud
@@
 as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers \citep{jetstream}, or private or institutional server hardware.  Being
+Jetstream servers \citep{jetstream}, or private or institutional server hardware. 
+Lawrence Berkeley National Labs (LBNL) has made `Singularity`: software which can import
+and run Docker containers without root privileges, as most cluster administrators require [@singularity].  
+Another LBNL project, *Shifter*, allows Docker containers to be deployed on leadership-scale
+high performance computing platforms [@shifter].
+
+Importantly, Rocker containers are just as easy to run on locally on most modern laptops
+by first installing the appropraite Docker distribution for Windows, MacOS, and Linux-based
+operating systems. By sharing volumes with the local host, users can still manipulate
+files with familiar, native tools while performing computation through a reproducible,
+containerized environment [@Boettiger2015]. Being
 able to test code in a predictable, pre-configured R environment on a
@@
 to free up the local machine from a long-running computation) is essential
-for low-friction scaling of analysis.  Without such contanerization,
+for low-friction scaling of analysis.  Without such containerization,
 getting code to run appropriately in a remote environment can be a major
@@
 not have.
-
-Perhaps more importantly, Rocker containers are just as easy to run
-on locally on most modern laptops on Windows, MacOS, and Linux-based
-operating systems.
-
-<!-- comment on use in instructional context -->
-This portablility is similarly valuable in an instructional context. Requiring students
-to install all necessary software on personal laptops can be particularly challenging
-for short workshops, where download and installation time and troubleshooting across
-heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
-By deploying a Rocker image or Rocker-derived image (see extensibility) on a cloud machine,
-an instructor can easily provide all students access to the pre-configured software environment
-using only the browser on their laptops.  Unlike a vendor-specific virtual machine image (e.g. Amazon Machine Images, AMIs),
-this approach does not preclude students being able to run the same software locally later on.
-After installing Docker for their machine they too can run the course-specific software environment
-easily on their own laptops and deploy it on other cloud providers.  This strategy has proven
-effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
-UC Berkeley \citep{data8}.
-
-<!-- NOTE: Not sure if we should include such example commands, may be best left to the documentation? -->
 
@@
 possible to execute such code snippets when an instance is launched, without
-ever needing to `ssh` into the machine.
-After running these two lines, the user may log into the server merely
+ever needing to `ssh` into the machine. The user may log into the server merely
 by pasting its IP address or DNS name into a browser and entering the
@@
 a minimum of expertise.
+
+This portablility is also valuable in an instructional context. Requiring students
+to install all necessary software on personal laptops can be particularly challenging
+for short workshops, where download and installation time and troubleshooting across
+heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
+By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
+an instructor can easily provide all students access to the pre-configured software environment
+using only the browser on their laptops. This strategy has proven
+effective in our own experience in both workshops and semester-length courses. Similar Docker-based
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+UC Berkeley \citep{data8}.
 
@@
 
-**Linking local volumes**
-
-Users of the Docker's GUI interface, Kitematic who use the
-`rstudio`-derived images will automatically have their local kitematic
-directory linked to the `/home/rstudio/kitematic` directory on the
-Docker container.
-
-<!-- Blah blah blah ... link to documentation instead(?) -->
-**Root vs user instances**: By default, all rocker containers run as
-root. This is consistent with standard Docker practice, allowing both
-interactive users and downstream Dockerfiles to install software directly
-without having to switch to root.  The main snag of this approach comes if
-a user links a local volume and modifies files on that volume, which will
-result in the local file being owned by `root` and not the default user.
-To avoid this, specify the non-root user at runtime (username `docker`
-on the `r-base` stack, or `rstudio` on images deriving from rstudio)
-when running an interactive shell.  For users accessing R through RStudio,
-this is not necessary.  The docker container will run as root, but a user
-logs in through the RStudio server web interface as the non-root user
-"rstudio", and thus any changes made to linked volumes will not alter file
-permissions on the home directory.  For more details on user permissions,
-see the Rocker website, <https://rocker-project.org>.
+Details on sharing volumes, managing user permissions, and more can be found
+on the Rocker website, <https://rocker-project.org>.
 
@@
 a typical host can easily support 100s of containers \cite{what-docker}.
-
-<!-- volume linking isolation? -->
 
@@
 
-\citep{@Boettiger2015}
+The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
+and `geospatial`), provides images which are intended to build an identical
+software stack every time, regardless of the release of new libraries and packages.  This 
+can be important for users who need computational reproducibility more than having the latest
+release of any piece of software, since subsequent releases can alter the behavior of code,
+introducing errors or altering previous results.  Users should specify an R version tag in
+the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
+is explicitly requested, Docker will provide the image with the tag `:latest`, which will
+always have the latest available versions of the software (built nightly).
+
+Users building on the version-tagged images will by default use the MRAN snapshot mirror [@MRAN]
+associated with the most recent date for which that image was current.  This ensures that
+a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
+available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
+can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
+command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
+in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
+writing) will continue to advance on the Docker-hub image until the next R release.  Software
+installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
+and thus not change versions (though it will recieve security patches).  Packages installed from BioConductor
+using `bioclite()` utlitity will also install the version appropriate to the version of R found on the system
+(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
+from GitHub or other sources can request a specific git release tag or hash for a more reproducibile build,
+or adopt a more comprehensive solution such as packrat [@packrat]; however, @packrat should not be necessary
+for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
+A more general discussion of the use of Docker for computational reproducibility can be found in [@Boettiger2015].
 
 ### Extensible
-
-<!-- Not intended as a guide to writing Dockerfiles, see http://ropenscilabs.github.io/r-docker-tutorial/05-dockerfiles.html -->
 
@@
 the one they need.  The Rocker project seeks to avoid both of these problems
-by providing images that an be easily extended by individuals and communities.
+by providing a carefully curated suite of images that an be easily extended by
+individuals and communities.
 
+To make extensiosn transparent and persistent, Rocker images can be extended by
+any user by writing their own Dockerfiles based on an appropriate Rocker image.  
+The Dockerfiles in the Rocker stack should themselves provide a simple of example
+of this.  A user begins by selecting an appropriate base image for their needs: 
+if the RStudio interace is desired, a user might start with `FROM rocker/rstudio`;
+an image for testing a particular C code in an R package might use `FROM rocker/r-devel-san`,
+an image for reproducing a data analysis will probably select a stable version tag
+in addition to an appropriate base library, e.g.: `FROM rocker/tidyverse:3.4.1`
 Users can easily add additional software to any running Rocker image using
-the standard R and Debian mechanisms. To make these changes persistent, Rocker 
-images can be extended by any user writing their own Dockerfiles based
-on an appropriate Rocker image.  Many users maintain their own Dockerfiles
-which add software specific to their own needs on top of an existing Rocker image.
-This provides a convenient way to reap all the benefits listed above in a 
-customized environment.
+the standard R and Debian mechanisms. Details on how to extend Rocker images can
+be found at <https://rocker-project.org>.
 
@@
 
-<!-- see docs instead -->
-Rocker images provide a few utility functions to streamline this process, including the 
-`littler` \citep{littler} scripts which provide a concise syntax for installing packages in Dockerfiles, e.g.
-
-    RUN install2.r pkg1 pgk2 pkg3 ...
-
-Users building on the version-tagged images will by default use the MRAN snapshot mirror
-associated with the most recent date for which that image was current.  This ensures that
-a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
-available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  (This default
-can of course be overwritten by specifying a different CRAN mirror explicitly.)
-
-Users writing their own Dockerfiles are encouraged to follow the same practices as the
-Rocker Project, such as the Dockerfile Best Practices \citep{best_practices}, the use of automated builds,
-and when appropriate, versioned tags.  Users can ensure their automated builds are rebuilt
-every time the relevant upstream Dockerfile is updated by using build triggers on Docker Hub.
-
-Note that users can also preserve changes to Rocker images that they have modified interactively using
-the `docker commit` mechanism, which creates a new binary image which can be pushed back to
-a personal account on the Docker Hub.  While this is sometimes convenient, we encourage users to
-consider writing Dockerfiles instead whenever possible, as this creates a more transparent
-and reproducible mechanism to accomplish the same thing.
-
 ## Conclusions

### 2017-09-22 manuscript.Rmd
a complete draft!

@@
 The issues tracker and pull requests are used for community input,
-disucssions and contributions to these images. The rocker project wiki
+discussions and contributions to these images. The rocker project wiki
 provides a place to synthesize community-contributed documentation,
@@
 as binaries than would be found in a Debian stable release.  In order
-to provide access to the most recent avialble binaries, this stack uses
+to provide access to the most recent available binaries, this stack uses
 apt-pinning \citep{apt_pinning} to allow the `apt` package manager to also
@@
 are sent to `unstable` during the normal course of Debian development.
-This can occassionally break an previously working installation command
+This can occasionally break an previously working installation command
 in a Dockerfile until the maintainer redirects the package manager to
@@
 
-1. These Dockerfiles require occassional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
+1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
 3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
-4. `testing` and `unstable` do not recieve the attention to security updates from the Debian security team \citep{debian_security}.
+4. `testing` and `unstable` do not receive the attention to security updates from the Debian security team \citep{debian_security}.
 <!-- DE: Are you sure about that point 4 ? -->
@@
 Docker organization itself.  This official image is reviewed and then
-built by employees of Docker Inc based on the Dockerfile maintained by
+built by employees of Docker Inc based on a Dockerfile maintained by
 the Rocker team.  Consequently, users should refer to this image in
-Docker commands without an organization napespace, e.g. `docker pull
+Docker commands without an organization namespace, e.g. `docker pull
 r-base` to access the official image.  All other images in the Rocker
 project are not individually reviewed and built by Docker Inc and must
-be refered to require the organization namespace on Docker Hub,
+be referenced using the `rocker` namespace,
 e.g. `docker pull rocker/r-devel`.
@@
 `r-base` image.  RStudio and Shiny are registered trademarks of RStudio
-Inc, and their use and the distibution of their software in binary form on
+Inc, and their use and the distribution of their software in binary form on
 Docker Hub has been granted to the Rocker project by explicit permission
@@
 [drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built almost daily  | 571 MB | 4,000
-[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanatizers | 1.1 GB  | 1,000  
+[r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanitizers | 1.1 GB  | 1,000  
 [r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
@@
 release at time (but will continue to reconstruct the same environment in
-the future after newer R versions are released) should explicilty include
+the future after newer R versions are released) should explicitly include
 the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
@@
 
-**Version tags**:  The version tags are propegated throughout this
+**Version tags**:  The version tags are propagated throughout this
 stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
@@
 be built from a separate Dockerfile, enabling minor differences in the
-build instructions to accomodate changing dependencies.  Dockerfiles for
+build instructions to accommodate changing dependencies.  Dockerfiles for
 past versions (e.g. < `3.4.0` currently) are intended to remain static
 over the long term, while the tag for the current version, `latest`,
-and `devel` may be tweaked to accomodate new features or dependencies.
+and `devel` may be tweaked to accommodate new features or dependencies.
 
@@
 stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
-codenamed `jessie`, while `3.4.0`, `devel`, and `latest` are based on
+code-named `jessie`, while `3.4.0`, `devel`, and `latest` are based on
 Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
@@
 While most system dependencies required by common R packages can still
-be installed from the `apt` repositories, occassionally a more recent
+be installed from the `apt` repositories, occasionally a more recent
 version must be compiled from source (e.g. the Gibbs Sampling library,
@@
 software environment every time, the monthly rebuilding of these images
-on Docker Hub ensures they continue to recieve Debian security updates
+on Docker Hub ensures they continue to receive Debian security updates
 from upstream, and proves the build recipe still executes successfully.
 Note that rebuilding images with software from external repositories
-never produces a bitwise identical image, and thus the image identifier
+never produces a bit-wise identical image, and thus the image identifier
 hash will change at each build.
@@
 design limits flexibility (no option for `tidyverse` without `rstudio`)
-but simplifies use and maintance.  While no single environment will be
+but simplifies use and maintenance.  While no single environment will be
 optimal for everyone, both the packages selected in this stack and the
@@
 The `rstudio` image includes a lightweight, easy-to-use and
-docker-friendly `init` system, s6 \citep{s6} for running persistant services,
+docker-friendly `init` system, s6 \citep{s6} for running persistent services,
 including the RStudio server.  This system provides a convenient way for
@@
 --------|-----------|--------------|------------------|-------------------------------           
-devel   | `stretch` | current date | nigthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+devel   | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
 latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
@@
 One common use case for Rocker containers is to provide a fast and
-reliable mechanism to deploy a custom R envionment to a remote server, such
+reliable mechanism to deploy a custom R environment to a remote server, such
 as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
@@
 Lawrence Berkeley National Labs (LBNL) has made `Singularity`: software which can import
-and run Docker containers without root privileges, as most cluster administrators require [@singularity].  
+and run Docker containers without root privileges, as most cluster administrators require \citep{singularity}.  
 Another LBNL project, *Shifter*, allows Docker containers to be deployed on leadership-scale
-high performance computing platforms [@shifter].
+high performance computing platforms \citep{shifter}.
 
 Importantly, Rocker containers are just as easy to run on locally on most modern laptops
-by first installing the appropraite Docker distribution for Windows, MacOS, and Linux-based
+by first installing the appropriate Docker distribution for Windows, MacOS, and Linux-based
 operating systems. By sharing volumes with the local host, users can still manipulate
 files with familiar, native tools while performing computation through a reproducible,
-containerized environment [@Boettiger2015]. Being
+containerized environment \citep{Boettiger2015}. Being
 able to test code in a predictable, pre-configured R environment on a
@@
 
-Many academic and commerical cloud providers make it
+Many academic and commercial cloud providers make it
 possible to execute such code snippets when an instance is launched, without
@@
 
-This portablility is also valuable in an instructional context. Requiring students
+This portability is also valuable in an instructional context. Requiring students
 to install all necessary software on personal laptops can be particularly challenging
@@
 use case is to access an instance of R with support for particular
-options such as using gcc or clang compiler sanatizers \citep{edd_sanitizers}.
+options such as using gcc or clang compiler sanitizers \citep{edd_sanitizers}.
 These requires R itself be built with specialized settings that may not be
@@
 remote computing context, where it can allow system administrators to
-provide users with administrative (root) privilages to alter and install
+provide users with administrative (root) privileges to alter and install
 software within a container without altering the computational environment
@@
 Users can easily determine the software stack installed on any Rocker image
-by examining the associated Dockerfile recipe, which provides a consise, human-readable
+by examining the associated Dockerfile recipe, which provides a concise, human-readable
 record of the installation.
@@
 than performance.  The Rocker recipes reflect significant community input on these choices,
-as well as the considerable experience and expertise of the R Debian mantainer of 20 years
+as well as the considerable experience and expertise of the R Debian maintainer of 20 years
 in what configuration options to use.  This helps create a more finely tuned, optimized reference
@@
 
-Users building on the version-tagged images will by default use the MRAN snapshot mirror [@MRAN]
+Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
 associated with the most recent date for which that image was current.  This ensures that
@@
 installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
-and thus not change versions (though it will recieve security patches).  Packages installed from BioConductor
-using `bioclite()` utlitity will also install the version appropriate to the version of R found on the system
+and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
+using `bioclite()` utility will also install the version appropriate to the version of R found on the system
 (the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
-from GitHub or other sources can request a specific git release tag or hash for a more reproducibile build,
-or adopt a more comprehensive solution such as packrat [@packrat]; however, @packrat should not be necessary
+from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
+or adopt a more comprehensive solution such as `packrat` \citep{packrat}; however, `packrat` should not be necessary
 for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
-A more general discussion of the use of Docker for computational reproducibility can be found in [@Boettiger2015].
+A more general discussion of the use of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
 
@@
 the "kitchen sink problem" at one extreme, and the "discovery problem" on
-the other.  A kitchen sink image seeks to accomidate too many use cases
+the other.  A kitchen sink image seeks to accommodate too many use cases
 in a single image.  Such images are inevitably very large and thus slow
@@
 
-To make extensiosn transparent and persistent, Rocker images can be extended by
+To make extensions transparent and persistent, Rocker images can be extended by
 any user by writing their own Dockerfiles based on an appropriate Rocker image.  
@@
 of this.  A user begins by selecting an appropriate base image for their needs: 
-if the RStudio interace is desired, a user might start with `FROM rocker/rstudio`;
+if the RStudio interface is desired, a user might start with `FROM rocker/rstudio`;
 an image for testing a particular C code in an R package might use `FROM rocker/r-devel-san`,
@@
 such as the `popgen` image of population-genetics-oriented software developed 
-by the National Evolutionary Synthesis Center (NESCent).
+by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
+being used as base Docker images in the NSF sponsored Whole Tale project for
+reproducible computing \citep{wholetale}.
 
@@
 
-Lessons for related efforts
-
-Future directions, Open Container Initiative, Singularity & Shifter (support for Docker on HPC)
+Over the past several years, Docker has seen immense adoption across industry and academia. 
+The Open Container initiative \citep{oci} now provides an open standard that has further
+extended this container approach to research environments through projects such as 
+Singularity \citep{singularity}, used on clusters without root access, and Shifter \citep{shifter}
+for use on leadership high performance computing facilities. Containerization promises
+to solve numerous challenges such as portability and replicability in research computing,
+which often relies on complex and heterogeneous software stacks \citep{Boettiger2015}. Yet
+implementing such environments in containers is not a trivial task, and not all implementations
+provide the same usability, portability or reproducibility.  Here we have detailed the 
+approach taken by the Rocker project in creating and maintaining these environments through
+an open and community-driven process.  This structure of the Rocker project has evolved
+over three years of operation while drawing in an ever widening base of academic researchers, 
+university instructors and industry users.  We believe this overview will be instructive
+not only to users and developers interested in the Rocker project, but as a model for
+similar efforts around other environments or domains.
 

### 2017-09-23 manuscript.Rmd
a few quick edits mostly to the .bib

@@
 not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
-and in the linked GitHub repository of the Rocker project (<https://github.com/rocker-org), 
+and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
 which provides a transparent versioned history of all changes made to these recipes,

### 2017-09-24 manuscript.Rmd
WIP making fixes proposed by Dirk

--- 2017-09-22 manuscript.Rmd
@@
   with customized R environments for particular tasks. We discuss how this suite is
-  organized and how these tools can increase portability, scaling, reproducibility,
+  organized, and how these tools can increase portability, scaling, reproducibility,
   and convenience of R users and developers.
@@
 and run software applications using *containers*.  Containers provide a
-virtual environment requiring all operating-system components an application needs to run:
+virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
 code, runtime, system tools, system runtime.  Docker containers are

### 2017-09-24 manuscript.Rmd
more consistent use of 'Rocker project'
only Rocker is our name, so opted for consistent lower-cased project

@@
 ---
-title: "An Introduction to Rocker: Docker Containers for R"
+title: | 
+       | An Introduction to Rocker: 
+       | Docker Containers for R
 author:
@@
 
-The Rocker Project consists of a suite of images built automatically by
+The Rocker project consists of a suite of images built automatically by
 and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
@@
 
-The Rocker Project aims to provide a small core of Docker images that
+The Rocker project aims to provide a small core of Docker images that
 serve as convenient 'base' images on which other users can build custom
@@
 the pre-release distribution, `debian:testing`, while the very latest
-binary builds can be found on `debian:unstable`.   The Rocker Project
+binary builds can be found on `debian:unstable`.   The Rocker project
 can be largely divided into two stacks which address different needs,
@@
 for a wide range of commonly used data formats and libraries.  Other community
-images are developed and maintained independently of the Rocker Project,
+images are developed and maintained independently of the Rocker project,
 such as the `popgen` image of population-genetics-oriented software developed 

### 2017-09-24 manuscript.Rmd
a few more quick edits

@@
 are system dependencies rather than R packages installed from within R.
-We should note, however, that only about 500 of the currently almost
+We should note, however, that only about 500 of the over
 11000 CRAN packages are available as Debian packages.
@@
 3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
-4. `testing` and `unstable` do not receive the attention to security updates from the Debian security team \citep{debian_security}.
+<!--4. `testing` and `unstable` do not receive the attention to security updates from the Debian security team \citep{debian_security}.-->
 <!-- DE: Are you sure about that point 4 ? -->
 <!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->
+<!-- DE: Which says nothing about _unstable_ not getting security updates, which is implied here-->
 

### 2017-09-25 manuscript.Rmd
finish addressing Dirk edit list.

@@
 code, runtime, system tools, system runtime.  Docker containers are
-lightweight as they share the operating system kernel, starting instantly,
+lightweight as they share the operating system kernel, starting instantly
 using a layered filesystem which minimizes disk footprint and download time,
@@
 is based on `debian:testing`, the second, more recently introduced stack
-is based only on `debian` stable releases.  These stacks have different
+is based only on Debian stable releases.  These stacks have different
 aims and thus provide different images, as shown in Tables 1 & 2.
@@
 3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
-<!--4. `testing` and `unstable` do not receive the attention to security updates from the Debian security team \citep{debian_security}.-->
-<!-- DE: Are you sure about that point 4 ? -->
-<!-- CB: I dunno, just quoting from https://www.debian.org/releases/testing/ -->
-<!-- DE: Which says nothing about _unstable_ not getting security updates, which is implied here-->
 
@@
 feature of this stack is the ability to run older versions of R along
-with the contemporaneous versions of R packages.  A user specifies the
+with the then-contemporaneous versions of R packages.  A user specifies the
 version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
@@
 JAGS, and the geospatial toolkit GDAL, must both be compiled from source
-on `debian:jessie` images).  In this stack, users should never install R
-packages, using `apt`, since this will install a second, older version
+on `debian:jessie` images).  In this stack, users should avoid installing R
+packages using `apt` without careful consideration, as this will install a second, 
+probably different version
 of R from the Debian repositories and a dated version of the R package
@@
     wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -d -p 80:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
 
@@
 ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name into a browser and entering the
+by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
+into a browser and entering the
 appropriate password.  This provides the user with a familiar,
@@
 from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
-or adopt a more comprehensive solution such as `packrat` \citep{packrat}; however, `packrat` should not be necessary
+or adopt an alternative approach such as `packrat` \citep{packrat}.
+A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
+<!--
+however, `packrat` should not be necessary
 for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
-A more general discussion of the use of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
+-->
 
@@
 being used as base Docker images in the NSF sponsored Whole Tale project for
-reproducible computing \citep{wholetale}.
+reproducible computing \citep{wholetale}, and are heavily used by the `rhub`
+project in automated package testing \citep{rhub}.
 

### 2017-10-06 manuscript.Rmd
Working through Daniel's suggestions in #2. (still going).
Also, flips the order to present the big picture stuff first and details on the project setup / organization after, as Patrick recommended.

@@
 using a layered filesystem which minimizes disk footprint and download time,
-and are built on open standards that run on all major platforms (Linux, Mac,
+are built on open standards that run on all major platforms (Linux, Mac,
 Windows), and provide an added layer of security by running an application
-in an isolated environment \citep{what-docker}.
+in an isolated environment \citep{what-docker}.  Familiarity with a few key
+terms is helpful in understanding this paper. The term "container" refers to 
+an isolated software environment on a computer. R users can think of running
+a container as analgous to loading an R package; it makes a set of software
+functions available.  A Docker "image" is a binary archive of that software,
+analgous to an R binary package -- a given version need be downloaded only
+once and then can be "run" to create a container whenever it is needed.  
+A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
+and contributions to the Rocker project focuses on the construction, organization
+and maintenance of these Dockerfiles.
+
+## Design principles & use cases
+
+<!--  The main point here is that these are generic features that could be goals 
+      of any effort to provide researchers with virtual computing environments.
+
+      While most virtual computing environments (AMI images, scientific cloud
+      platforms like cyverse, etc) provide some subset of these, they are most
+      effective in concert.
+
+- **Portable**: from laptop to cloud
+- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
+- **Sandboxed**: experiment without touching local system
+- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
+- **Optimized**: community-tweaked configurations
+- **Extensible**: interactively and persistent community extensions
+- **Versioned**: reproducible environments
+
+-->
+
+### Portability: From laptop to cloud
+
+One common use case for Rocker containers is to provide a fast and
+reliable mechanism to deploy a custom R environment to a remote server, such
+as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
+Jetstream servers \citep{jetstream}, or private or institutional server hardware. 
+Lawrence Berkeley National Labs (LBNL) has made `Singularity`: software which can import
+and run Docker containers without root privileges, as most cluster administrators require \citep{singularity}.  
+Another LBNL project, *Shifter*, allows Docker containers to be deployed on supercomputing
+architectures such as CRAY \citep{shifter}.
+
+Importantly, Rocker containers are just as easy to run on locally on most modern laptops
+by first installing the appropriate Docker distribution for Windows, MacOS, and Linux-based
+operating systems. By sharing volumes with the local host, users can still manipulate
+files with familiar, native tools while performing computation through a reproducible,
+containerized environment \citep{Boettiger2015}. Being
+able to test code in a predictable, pre-configured R environment on a
+local machine and to then run the same code in an identical environment on a
+remote server (e.g. for access to greater RAM, more processors, or merely
+to free up the local machine from a long-running computation) is essential
+for low-friction scaling of analysis.  Without such containerization,
+getting code to run appropriately in a remote environment can be a major
+undertaking, requiring both time and knowledge many would-be users may
+not have.
+
+For instance, the following code installs Docker on almost any Linux-based server and
+launches a Rocker container providing the RStudio-server environment
+over a web interface.
+
+    wget -qO- https://get.docker.com/ | sh 
+    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+
+Many academic and commercial cloud providers make it
+possible to execute such code snippets when a container is launched, without
+ever needing to `ssh` into the machine. The user may log into the server merely
+by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
+into a browser and entering the
+appropriate password.  This provides the user with a familiar,
+interactive environment running on a remote machine while requiring
+a minimum of expertise.
+
+This portability is also valuable in an instructional context. Requiring students
+to install all necessary software on personal laptops can be particularly challenging
+for short workshops, where download and installation time and troubleshooting across
+heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
+By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
+an instructor can easily provide all students access to the pre-configured software environment
+using only the browser on their laptops. This strategy has proven
+effective in our own experience in both workshops and semester-length courses. Similar Docker-based
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+UC Berkeley \citep{data8}.
+
+### Interfaces
+
+An important aspect of the Rocker project design is the ability
+for users to interact with the software on the container through either
+an interactive shell session (such as the R shell or a bash shell), or
+through a web browser accessing the RStudio\textsuperscript{\textregistered} Server
+integrated development environment (IDE).  Traditional remote and high-performance
+computing workflows for R users have usually required the use of `ssh` and 
+a terminal only interface, posing a challenge for interactive graphics 
+and a barrier to users unfamiliar with these tools and environments. Accessing
+an RStudio\textsuperscript{\textregistered} container through the browser removes these barriers.  Rocker
+images include the RStudio-server software pre-installed and configured with
+the explicit permission of RStudio\textsuperscript{\textregistered} Inc.
+
+Users can access a root `bash` shell in a running Rocker container using
+
+```
+docker exec -ti <container-id> bash
+```
+
+which can be useful for administrative tasks such as installing system dependencies.
+All Rocker images can also be run as an interactive R, RScript or bash shell without running
+RStudio, which can be useful for batch jobs or for anyone who prefers that environment.
+
+<!--  Reserve these details for documentation, reference here? --> 
+As with any interactive Docker container, users
+should specify the interactive (`-i`) and terminal (`-t`, here combined
+with interactive as `-ti`) flags, and specify the desired executable
+environment (e.g. `R`, though other common options may be `Rscript` or
+`bash`):
+
+    docker run --rm -ti rocker/tidyverse R
+
+This example shows the use of the `--rm` flag to indicate that the
+container should be removed when the interactive session is finished.
+
+Details on sharing volumes, managing user permissions, and more can be found
+on the Rocker website, <https://rocker-project.org>.
+
+### Sandboxed
+
+Another feature of Rocker containers is the ability to provide a sandboxed 
+environment, isolated from software and potentially from other data on
+the machine.  Many users are reluctant to upgrade their suite of installed packages,
+which may break their existing code or even their R environment if the
+installation goes poorly.  However, upgrading packages and/or the R
+environment is often necessary to run analysis of a colleague or access
+more recent methods.  Rocker offers an easy solution. For 
+instance, a user can run R code requiring the most recent versions of R
+and related packages inside a Rocker container without having to upgrade their
+local installations first.  Conversely, one could use Rocker to run code
+on an older R release with prior versions of R packages, again without
+having to make any alteration to one's local R install.  Another common
+use case is to access a container with support for particular
+options such as using gcc or clang compiler sanitizers \citep{edd_sanitizers}.
+These requires R itself be built with specialized settings that may not be
+not available or familiar to many R users on their native system, but can
+be easily deployed by pulling the Rocker images `rocker/r-devel-san` or
+`rocker/r-devel-ubsan-clang`.
+
+This sandboxing feature is also valuable in the 
+remote computing context, allowing system administrators to grant
+users freedom to install software which requires root privileges inside a container,
+while not granting them root access on the host machine. Root access
+is required to launch Docker containers, though not to access containers already
+running and providing some service such as RStudio. Users logging into a container
+through the RStudio\textsuperscript{\textregistered} interface do not by default have root privileges, though are able
+to install R packages. Granting these users root privileges in the container still leaves
+them sandboxed from the host container.  Further, Docker images can
+be deployed from a machine without root privileges through the alternate container
+runtime environment, Singularity\citep{singularity}.  Unlike traditional virtual machines,
+these containers do not impose a large footprint of reserved resources --
+a typical host can easily support 100s of containers \cite{what-docker}.
+
+### Transparent
+
+Users can easily determine the software stack installed on any Rocker image
+by examining the associated Dockerfile recipe, which provides a concise, human-readable
+record of the installation.
+All Rocker images use automated builds through Docker Hub, which also acts as the central,
+default repository distributing the images.  Using automated builds rather than uploading
+pre-built image binaries to the Docker Hub avoids the potential for the build
+not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
+and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
+which provides a transparent versioned history of all changes made to these recipes,
+as well as documentation, a community wiki, and issue trackers for discussing proposed
+changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
+
+### Community Optimized
+
+Having a shared, transparent computational environment created by a publicly hosted, reproducible
+recipe facilitates community input into configuration details. R and many of its packages
+and related software can be configured with a wide range of options, compilers, different
+linear-algebra libraries and so forth.  While this flexibility reflects varying needs,
+many users rely on default settings which may be optimized more for simplicity of installation
+than performance.  The Rocker recipes reflect significant community input on these choices,
+as well as the considerable experience and expertise of the R Debian maintainer of 20 years
+in what configuration options to use.  This helps create a more finely tuned, optimized reference
+implementation of the R environment as well as a platform for comparing and discussing these 
+concerns which are often overlooked elsewhere.  Issues and Pull Requests on the Rocker 
+repositories on GitHub attest to these discussions and improvements.
+Widespread use of the Rocker image helps promote both testing of these choices and 
+contributions further tweaking the configuration from many members of the R community.
+
+### Versioned
+
+The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
+and `geospatial`), provides images which are intended to build an identical
+software stack every time, regardless of the release of new libraries and packages.  This 
+can be important for users who need computational reproducibility more than having the latest
+release of any piece of software, since subsequent releases can alter the behavior of code,
+introducing errors or altering previous results.  Users should specify an R version tag in
+the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
+is explicitly requested, Docker will provide the image with the tag `:latest`, which will
+always have the latest available versions of the software (built nightly).
+
+Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
+associated with the most recent date for which that image was current.  This ensures that
+a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
+available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
+can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
+command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
+in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
+writing) will continue to advance on the Docker-hub image until the next R release.  Software
+installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
+and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
+using `bioclite()` utility will also install the version appropriate to the version of R found on the system
+(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
+from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
+or adopt an alternative approach such as `packrat` \citep{packrat}.
+A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
+<!--
+however, `packrat` should not be necessary
+for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
+-->
+
+### Extensible
+
+Any portable computational environment faces an inevitable tension between
+the "kitchen sink problem" at one extreme, and the "discovery problem" on
+the other.  A kitchen sink image seeks to accommodate too many use cases
+in a single image.  Such images are inevitably very large and thus slow
+or difficult to deploy, maintain and optimize.  At the other extreme, providing
+too many specialized images makes it more difficult for a user to discover 
+the one they need.  The Rocker project seeks to avoid both of these problems
+by providing a carefully curated suite of images that an be easily extended by
+individuals and communities.
+
+To make extensions transparent and persistent, Rocker images can be extended by
+any user by writing their own Dockerfiles based on an appropriate Rocker image.  
+The Dockerfiles in the Rocker stack should themselves provide a simple of example
+of this.  A user begins by selecting an appropriate base image for their needs: 
+if the RStudio\textsuperscript{\textregistered} interface is desired, a user might start with `FROM rocker/rstudio`;
+an image for testing a particular C code in an R package might use `FROM rocker/r-devel-san`,
+an image for reproducing a data analysis will probably select a stable version tag
+in addition to an appropriate base library, e.g.: `FROM rocker/tidyverse:3.4.1`
+Users can easily add additional software to any running Rocker image using
+the standard R and Debian mechanisms. Details on how to extend Rocker images can
+be found at <https://rocker-project.org>.
+
+Sharing these Dockerfiles can also facilitate the emergence 
+of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
+image emerged from the input of a number of Rocker users all adding common
+geospatial libraries and packages on top of the existing Rocker images. 
+This coalescence helped create a more fine-tuned image with broad support 
+for a wide range of commonly used data formats and libraries.  Other community
+images are developed and maintained independently of the Rocker project,
+such as the `popgen` image of population-genetics-oriented software developed 
+by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
+being used as base Docker images in the NSF sponsored Whole Tale project for
+reproducible computing \citep{wholetale}, and are heavily used by the R-Hub
+project in automated package testing \citep{rhub}.
 
@@
 under the organization `rocker-org`, <https://github.com/rocker-org>.
-The issues tracker and pull requests are used for community input,
+The issue tracker and pull requests are used for community input,
 discussions and contributions to these images. The rocker project wiki
@@
 The relatively long period between stable Debian releases
-(roughly two years recently) means that software in the `debian:stable`
+(roughly two years recently) means that software in the Debian stable 
+(e.g. `debian:jessie`, `debian:stretch`)
 releases can lag significantly behind current releases of popular
@@
 can be largely divided into two stacks which address different needs,
-defined by which Debian distribution they are based on.  The first stack
+reflected in which Debian distribution they are based on.  The first stack
 is based on `debian:testing`, the second, more recently introduced stack
-is based only on Debian stable releases.  These stacks have different
+is based only on Debian stable releases.  Rocker images always point to specific
+stable releases (`jessie`, `stretch`), and do not use the tag `debian:stable`,
+which is a rolling tag that always points to the most
+recent stable version.   The different Rocker stacks have different
 aims and thus provide different images, as shown in Tables 1 & 2.
@@
 in a Dockerfile until the maintainer redirects the package manager to
-install a the package from the `unstable` sources that could previously
+install a package from the `unstable` sources that could previously
 be installed from `testing`, or vice versa (using the `-t` option in
-`apt`, see Box 1 on apt-pinning.)  That said, packages only migrate from 
+`apt`)  That said, packages only migrate from 
 `unstable` to `testing` after a period of several days---and if the 
@@
 the Rocker team.  Consequently, users should refer to this image in
-Docker commands without an organization namespace, e.g. `docker pull
+Docker commands without an organization namespace, e.g. `docker run -t
 r-base` to access the official image.  All other images in the Rocker
@@
 be referenced using the `rocker` namespace,
-e.g. `docker pull rocker/r-devel`.
+e.g. `docker run -t rocker/r-devel`.
 
@@
 As these images focus on developers and/or as base images for custom uses,
-this stack does include many specific R packages. Additional dependencies
+this stack does not include many specific R packages. Additional dependencies
 and packages can easily be installed from `apt`.  R packages not available
@@
 This stack also include the images `shiny` and `rstudio:testing` that
-provide Shiny server and RStudio server IDE from RStudio Inc, built on the
-`r-base` image.  RStudio and Shiny are registered trademarks of RStudio
+provide Shiny server and RStudio\textsuperscript{\textregistered} server IDE from RStudio\textsuperscript{\textregistered} Inc, built on the
+`r-base` image.  RStudio\textsuperscript{\textregistered} and Shiny are registered trademarks of RStudio
 Inc, and their use and the distribution of their software in binary form on
@@
 further distribution or other questions to <permissions@rstudio.com>.
-The Rocker project also provides images with RStudio server and Shiny
+The Rocker project also provides images with RStudio\textsuperscript{\textregistered} server and Shiny
 server in the stable versioned stack.
@@
 docker-friendly `init` system, s6 \citep{s6} for running persistent services,
-including the RStudio server.  This system provides a convenient way for
+including the RStudio\textsuperscript{\textregistered} server.  This system provides a convenient way for
 downstream Dockerfile developers to add additional persistent services
 (such as an `ssh` server) to a single container, or additional start-up
-or shutdown scripts that should be run when an instance starts up or
+or shutdown scripts that should be run when a container starts up or
 shuts down.  The `rstudio` image uses such a start-up script to configure
@@
 
-## Design principles & use cases
-
-<!--  The main point here is that these are generic features that could be goals 
-      of any effort to provide researchers with virtual computing environments.
-
-      While most virtual computing environments (AMI images, scientific cloud
-      platforms like cyverse, etc) provide some subset of these, they are most
-      effective in concert.
-
-- **Portable**: from laptop to cloud
-- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
-- **Sandboxed**: experiment without touching local system
-- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
-- **Optimized**: community-tweaked configurations
-- **Extensible**: interactively and persistent community extensions
-- **Versioned**: reproducible environments
-
--->
-
-### Portability: From laptop to cloud
-
-One common use case for Rocker containers is to provide a fast and
-reliable mechanism to deploy a custom R environment to a remote server, such
-as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers \citep{jetstream}, or private or institutional server hardware. 
-Lawrence Berkeley National Labs (LBNL) has made `Singularity`: software which can import
-and run Docker containers without root privileges, as most cluster administrators require \citep{singularity}.  
-Another LBNL project, *Shifter*, allows Docker containers to be deployed on leadership-scale
-high performance computing platforms \citep{shifter}.
-
-Importantly, Rocker containers are just as easy to run on locally on most modern laptops
-by first installing the appropriate Docker distribution for Windows, MacOS, and Linux-based
-operating systems. By sharing volumes with the local host, users can still manipulate
-files with familiar, native tools while performing computation through a reproducible,
-containerized environment \citep{Boettiger2015}. Being
-able to test code in a predictable, pre-configured R environment on a
-local machine and to then run the same code in an identical environment on a
-remote server (e.g. for access to greater RAM, more processors, or merely
-to free up the local machine from a long-running computation) is essential
-for low-friction scaling of analysis.  Without such containerization,
-getting code to run appropriately in a remote environment can be a major
-undertaking, requiring both time and knowledge many would-be users may
-not have.
-
-For instance, the following code installs Docker on almost any Linux-based server and
-launches a Rocker container providing the RStudio-server environment
-over a web interface.
-
-    wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
-
-Many academic and commercial cloud providers make it
-possible to execute such code snippets when an instance is launched, without
-ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
-into a browser and entering the
-appropriate password.  This provides the user with a familiar,
-interactive environment running on a remote machine while requiring
-a minimum of expertise.
-
-This portability is also valuable in an instructional context. Requiring students
-to install all necessary software on personal laptops can be particularly challenging
-for short workshops, where download and installation time and troubleshooting across
-heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
-By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
-an instructor can easily provide all students access to the pre-configured software environment
-using only the browser on their laptops. This strategy has proven
-effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
-UC Berkeley \citep{data8}.
-
-### Interfaces
-
-An important aspect of the Rocker project design is the ability
-for users to interact with the software on the container through either
-an interactive shell session (such as the R shell or a bash shell), or
-through a web browser accessing the RStudio\textsuperscript{\textregistered} Server
-integrated development environment (IDE).  Traditional remote and high-performance
-computing workflows for R users have usually required the use of `ssh` and 
-a terminal only interface, posing a challenge for interactive graphics 
-and a barrier to users unfamiliar with these tools and environments. Accessing
-an RStudio instance through the browser removes these barriers.  Rocker
-images include the RStudio-server software pre-installed and configured with
-the explicit permission of RStudio Inc.
-
-Users of the `rstudio` instances can access a root `bash` shell in an 
-instance running the container using
-
-```
-docker exec -ti <container-id> bash
-```
-
-which can be useful for administrative tasks such as installing system dependencies.
-All Rocker images can also be run as an interactive R, RScript or bash shell without running
-RStudio, which can be useful for batch jobs or for anyone who prefers that environment.
-
-<!--  Reserve these details for documentation, reference here? --> 
-As with any interactive docker instance, users
-should specify the interactive (`-i`) and terminal (`-t`, here combined
-with interactive as `-ti`) flags, and specify the desired executable
-environment (e.g. `R`, though other common options may be `Rscript` or
-`bash`):
-
-    docker run --rm -ti rocker/tidyverse R
-
-This example shows the use of the `--rm` flag to indicate that the
-container should be removed when the interactive session is finished.
-
-Details on sharing volumes, managing user permissions, and more can be found
-on the Rocker website, <https://rocker-project.org>.
-
-### Sandboxed
-
-Another feature of Rocker instances is the ability to provide a sandboxed 
-environment, isolated from software and potentially from other data on
-the machine.  Many users are reluctant to upgrade their suite of installed packages,
-which may break their existing code or even their R environment if the
-installation goes poorly.  However, upgrading packages and/or the R
-environment is often necessary to run analysis of a colleague or access
-more recent methods.  Rocker instances offer an easy solution. For 
-instance, a user can run R code requiring the most recent versions of R
-and related packages inside a Rocker container without having to upgrade their
-local installations first.  Conversely, one could use Rocker to run code
-on an older R release with prior versions of R packages, again without
-having to make any alteration to one's local R install.  Another common
-use case is to access an instance of R with support for particular
-options such as using gcc or clang compiler sanitizers \citep{edd_sanitizers}.
-These requires R itself be built with specialized settings that may not be
-not available or familiar to many R users on their native system, but can
-be easily deployed by pulling the Rocker images `rocker/r-devel-san` or
-`rocker/r-devel-ubsan-clang`.
-
-This sandboxing feature of Docker containers is also valuable in the 
-remote computing context, where it can allow system administrators to
-provide users with administrative (root) privileges to alter and install
-software within a container without altering the computational environment
-of other users on the remote machine.  Unlike traditional virtual machines,
-these containers do not impose a large footprint of reserved resources --
-a typical host can easily support 100s of containers \cite{what-docker}.
-
-### Transparent
-
-Users can easily determine the software stack installed on any Rocker image
-by examining the associated Dockerfile recipe, which provides a concise, human-readable
-record of the installation.
-All Rocker images use automated builds through DockerHub, which also acts as the central,
-default repository distributing the images.  Using automated builds rather than uploading
-pre-built image binaries to the Docker Hub avoids the potential for the build
-not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
-and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
-which provides a transparent versioned history of all changes made to these recipes,
-as well as documentation, a community wiki, and issues trackers for discussing proposed
-changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
-
-### Community Optimized
-
-Having a shared, transparent computational environment created by a publicly hosted, reproducible
-recipe facilitates community input into configuration details. R and many of its packages
-and related software can be configured with a wide range of options, compilers, different
-linear-algebra libraries and so forth.  While this flexibility reflects varying needs,
-many users rely on default settings which may be optimized more for simplicity of installation
-than performance.  The Rocker recipes reflect significant community input on these choices,
-as well as the considerable experience and expertise of the R Debian maintainer of 20 years
-in what configuration options to use.  This helps create a more finely tuned, optimized reference
-implementation of the R environment as well as a platform for comparing and discussing these 
-issues which are often overlooked elsewhere.  Issues and Pull Requests on the Rocker 
-repositories on GitHub attest to these discussions and improvements.
-Widespread use of the Rocker image helps promote both testing of these choices and 
-contributions further tweaking the configuration from many members of the R community.
-
-### Versioned
-
-The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
-and `geospatial`), provides images which are intended to build an identical
-software stack every time, regardless of the release of new libraries and packages.  This 
-can be important for users who need computational reproducibility more than having the latest
-release of any piece of software, since subsequent releases can alter the behavior of code,
-introducing errors or altering previous results.  Users should specify an R version tag in
-the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
-is explicitly requested, Docker will provide the image with the tag `:latest`, which will
-always have the latest available versions of the software (built nightly).
-
-Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
-associated with the most recent date for which that image was current.  This ensures that
-a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
-available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
-can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
-command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
-in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
-writing) will continue to advance on the Docker-hub image until the next R release.  Software
-installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
-and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
-using `bioclite()` utility will also install the version appropriate to the version of R found on the system
-(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
-from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
-or adopt an alternative approach such as `packrat` \citep{packrat}.
-A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
-<!--
-however, `packrat` should not be necessary
-for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
--->
-
-### Extensible
-
-Any portable computational environment faces an inevitable tension between
-the "kitchen sink problem" at one extreme, and the "discovery problem" on
-the other.  A kitchen sink image seeks to accommodate too many use cases
-in a single image.  Such images are inevitably very large and thus slow
-or difficult to deploy, maintain and optimize.  At the other extreme, providing
-too many specialized images makes it more difficult for a user to discover 
-the one they need.  The Rocker project seeks to avoid both of these problems
-by providing a carefully curated suite of images that an be easily extended by
-individuals and communities.
-
-To make extensions transparent and persistent, Rocker images can be extended by
-any user by writing their own Dockerfiles based on an appropriate Rocker image.  
-The Dockerfiles in the Rocker stack should themselves provide a simple of example
-of this.  A user begins by selecting an appropriate base image for their needs: 
-if the RStudio interface is desired, a user might start with `FROM rocker/rstudio`;
-an image for testing a particular C code in an R package might use `FROM rocker/r-devel-san`,
-an image for reproducing a data analysis will probably select a stable version tag
-in addition to an appropriate base library, e.g.: `FROM rocker/tidyverse:3.4.1`
-Users can easily add additional software to any running Rocker image using
-the standard R and Debian mechanisms. Details on how to extend Rocker images can
-be found at <https://rocker-project.org>.
-
-Sharing these Dockerfiles can also facilitate the emergence 
-of extensions tuned to particular communities.  For instance, the `rocker/geospatial`
-image emerged from the input of a number of Rocker users all adding common
-geospatial libraries and packages on top of the existing Rocker images. 
-This coalescence helped create a more fine-tuned image with broad support 
-for a wide range of commonly used data formats and libraries.  Other community
-images are developed and maintained independently of the Rocker project,
-such as the `popgen` image of population-genetics-oriented software developed 
-by the National Evolutionary Synthesis Center (NESCent). Rocker images are also 
-being used as base Docker images in the NSF sponsored Whole Tale project for
-reproducible computing \citep{wholetale}, and are heavily used by the `rhub`
-project in automated package testing \citep{rhub}.
-
 ## Conclusions

### 2017-10-06 manuscript.Rmd
Finish Daniel's edits (#2) and adds Patrick's suggested changes.
@eddelbuettel: I think this is ready to go on my end, just give me a thumbs up if it's okay to submit!

@@
 
-## Introduction
+# Introduction
 
@@
 an isolated software environment on a computer. R users can think of running
-a container as analgous to loading an R package; it makes a set of software
+a container as analogous to loading an R package; it makes a set of software
 functions available.  A Docker "image" is a binary archive of that software,
-analgous to an R binary package -- a given version need be downloaded only
+analogous to an R binary package -- a given version need be downloaded only
 once and then can be "run" to create a container whenever it is needed.
@@
 
-## Design principles & use cases
+# Design principles & use cases
 
@@
 
-      While most virtual computing environments (AMI images, scientific cloud
-      platforms like cyverse, etc) provide some subset of these, they are most
-      effective in concert.
-
 - **Portable**: from laptop to cloud
-- **Interfaces**: Console-based (R, bash, etc) or RStudio\textsuperscript{\textregistered} Server, Shiny\textsuperscript{\textregistered} server
+- **Interfaces**: Console-based (R, bash, etc) or RStudio
 - **Sandboxed**: experiment without touching local system
@@
 
-For instance, the following code installs Docker on almost any Linux-based server and
-launches a Rocker container providing the RStudio-server environment
-over a web interface.
+For instance, the following bash commands install Docker on almost any 
+Linux-based server and then launch a Rocker container providing the 
+RStudio-server environment over a web interface.
 
     wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -d -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
 
-Many academic and commercial cloud providers make it
+The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
+the RStudio default. Many academic and commercial cloud providers make it
 possible to execute such code snippets when a container is launched, without
 ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name (followed by the chosen port, RStudio's default `:8787` in the above example)
+by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
 into a browser and entering the
@@
 changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
+Having these public source files built automatically by a trusted provider (Docker Hub),
+rather than built locally and uploaded only in binary form provides is also useful from a
+security perspective in avoiding malware.
 
@@
 
+Access to specific versions of software  
+can be important for users who need computational reproducibility more than having the latest
+release of any piece of software, since subsequent releases can alter the behavior of code,
+introduce errors or otherwise alter previous results.  
 The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
 and `geospatial`), provides images which are intended to build an identical
-software stack every time, regardless of the release of new libraries and packages.  This 
-can be important for users who need computational reproducibility more than having the latest
-release of any piece of software, since subsequent releases can alter the behavior of code,
-introducing errors or altering previous results.  Users should specify an R version tag in
+software stack every time, regardless of the release of new libraries and packages.  
+Users should specify an R version tag in
 the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
@@
 
-## Rocker organization and workflow
+# Rocker organization and workflow
 
@@
 The challenges of balancing diverse needs driven by very different use
-cases (discussed in more detail below) against the overarching goals
+cases against the overarching goals
 of creating images that are still sufficiently light-weight, easy to use
@@
 
-## The `debian:testing`-based images
+### The `debian:testing`-based images
 
@@
 be referenced using the `rocker` namespace,
-e.g. `docker run -t rocker/r-devel`.
+e.g. `docker run -ti rocker/r-devel`.
 
@@
 in the `apt` repositories can be installed directly from CRAN using the
-`littler` scripts, see *Extending Rocker images* section below.
+`littler` scripts, as described in <https://rocker-project.org/use>.
 
@@
 ---------------- | ----------------------------------------- | ------ | ------- 
-[r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 510,000   
+[r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 632,000   
 [r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
-[drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built almost daily  | 571 MB | 4,000
+[drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built weekly  | 571 MB | 4,000
 [r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanitizers | 1.1 GB  | 1,000  
@@
 [rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
-[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 63,000
+[shiny](https://hub.docker.com/r/rocker/shiny)     |  shiny-server on r-base               | 409 MB | 123,000
 
 Table: The `debian:testing` image stack
+
+image            | description                               | size   | downloads 
+---------------- | ----------------------------------------- | ------ | ------- 
+[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 2,000
+[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000 (+46,000 as `hadleyverse`)
+[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
+[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 1,000
+
+Table: The `rocker-versioned` stack of images
 
@@
 the future after newer R versions are released) should explicitly include
-the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time
+the corresponding version tag, e.g. `rocker/r-ver:3.4.2` at the time
 of writing, and not the `latest` tag.  Users can also run the current
@@
 and `devel` may be tweaked to accommodate new features or dependencies.
+Version tags also obey semantics, so that omitting the second or third
+position of the tag is identical to asking for the most recent version:
+i.e. `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
+`rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.
+This is accomplished using post-build hooks in Docker Hub, see examples at
+<https://github.com/rocker-org/rocker-versioned/> for details.
 
@@
 stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
-code-named `jessie`, while `3.4.0`, `devel`, and `latest` are based on
-Debian 9.0, `stretch`, released 2017-06-17, and thus have access to much
+code-named `jessie`, while `3.4.0` - `3.4.2`, `devel`, and `latest` are based on
+Debian 9.0, `stretch`, (released 2017-06-17, while R was at `3.4.0`), 
+and thus have access to much
 newer versions of common system dependencies and compilers.  Dependencies
@@
 
-image            | description                               | size   | downloads 
----------------- | ----------------------------------------- | ------ | ------- 
-[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 2,000
-[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
-[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000 (+46,000 as `hadleyverse`)
-[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
-[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 1,000
-
-Table: The `rocker-versioned` stack of images
-
 tag     | apt repos | MRAN date    | Build frequency  | images with tag 
@@
 latest  | `stretch` | current date | nightly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
-3.4.1   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
-3.4.0   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.4.2   | `stretch` | current date | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.4.1   | `stretch` | 2017-09-28 | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
+3.4.0   | `stretch` | 2017-06-30 | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial`
 3.3.3   | `jessie`  | 2017-04-21   | monthly          | `r-ver`, `rstudio`, `tidyverse`, `verse`, `geospatial` 
@@
 
-## Maintenance and Updates
+### Maintenance and Updates
 
@@
 
-## License ##
+### License
 
@@
 
-<!-- 
-Use cases: instruction, development, 
--->
-
-## Conclusions
+# Conclusions
 
@@
 extended this container approach to research environments through projects such as 
-Singularity \citep{singularity}, used on clusters without root access, and Shifter \citep{shifter}
-for use on leadership high performance computing facilities. Containerization promises
+Singularity \citep{singularity}, allowing users to deploy containerized environments such as Rocker on machines where they do not have root access, such as clusters or private servers.
+Containerization promises
 to solve numerous challenges such as portability and replicability in research computing,

### 2017-10-07 manuscript.Rmd

@@
 
-Docker is a popular open source tool to create, distribute, deploy,
+Docker is a popular open-source tool to create, distribute, deploy,
 and run software applications using *containers*.  Containers provide a
@@
 
+<!-- DE: It is a little weird to have this first as it is a somewhat
+advanced, and I'd conjecture "proportionally speaking, "rare" use.
+Isn't the main use to just "deploy an application" _locally_? I how we
+hand users R, or R-devel, or RStudio, or Tibbliesverse ?
+
+By the same token, it is a bit weird to intro Singularity _here_ and I
+still don't know if Shifter really needs to be in the paper (cf
+Patrick's comments)
+
+So should we maybe insert a really short para like this
+
+   Deployment: Ready-to-run binaries
+
+   Docker permits convenient access to pre-configured and create
+   _binary_ images that "just work".  This allows R users to access a
+   wider-variety of ready-to-use environments than the R Project
+   itself provides.  For example, R users on Windows may run RStudio
+   Server or Shiny Server locally just by launching a single command
+   (once Docker itself is installed), access R-devel without affecting
+   their local system and more.
+
+nd then start the para with 'Another' instead on one?  There is a
+little overlap with 'Sandboxed' below but I think it is worth it to
+start with "first things first" here
+
+-->
+
 ### Portability: From laptop to cloud
@@
 
-Importantly, Rocker containers are just as easy to run on locally on most modern laptops
-by first installing the appropriate Docker distribution for Windows, MacOS, and Linux-based
+Importantly, Rocker containers are just as easy to run locally on most modern laptops
+by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
 operating systems. By sharing volumes with the local host, users can still manipulate
@@
 is required to launch Docker containers, though not to access containers already
+<!-- DE False. On Linux the dockerd has the correct rights, and I (as member of group Docker) 
+can access docker without it.  That is important: minimal sudo use!  -->
 running and providing some service such as RStudio. Users logging into a container
@@
 be deployed from a machine without root privileges through the alternate container
-runtime environment, Singularity\citep{singularity}.  Unlike traditional virtual machines,
+runtime environment, Singularity \citep{singularity}.  Unlike traditional virtual machines,
 these containers do not impose a large footprint of reserved resources --
-a typical host can easily support 100s of containers \cite{what-docker}.
+a typical host can easily support 100s of containers \citep{what-docker}.
 
@@
 
-Access to specific versions of software  
-can be important for users who need computational reproducibility more than having the latest
-release of any piece of software, since subsequent releases can alter the behavior of code,
-introduce errors or otherwise alter previous results.  
-The versioned stack, (`r-ver`, `rstudio`, `tidyverse`, `verse`,
-and `geospatial`), provides images which are intended to build an identical
-software stack every time, regardless of the release of new libraries and packages.  
-Users should specify an R version tag in
-the Docker image name to request a version stable image, e.g. `rocker/verse:3.4.0`.  If no tag
-is explicitly requested, Docker will provide the image with the tag `:latest`, which will
+Access to specific versions of software can be important for users who
+need computational reproducibility more than having the latest release
+of any piece of software, since subsequent releases can alter the
+behavior of code, introduce errors or otherwise alter previous
+results.  The versioned stack, (`r-ver`, `rstudio`, `tidyverse`,
+`verse`, and `geospatial`), provides images which are intended to
+build an identical software stack every time, regardless of the
+release of new libraries and packages.  Users should specify an R
+version tag in the Docker image name to request a version stable
+image, e.g. `rocker/verse:3.4.0`.  If no tag is explicitly requested,
+Docker will provide the image with the tag `:latest`, which will
 always have the latest available versions of the software (built nightly).
@@
 or adopt an alternative approach such as `packrat` \citep{packrat}.
-A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}.
+A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \cite{Boettiger2015}.
 <!--
@@
 the one they need.  The Rocker project seeks to avoid both of these problems
-by providing a carefully curated suite of images that an be easily extended by
+by providing a carefully-curated suite of images that an be easily extended by
 individuals and communities.
@@
 The issue tracker and pull requests are used for community input,
-discussions and contributions to these images. The rocker project wiki
+discussions and contributions to these images. The Rocker project wiki
 provides a place to synthesize community-contributed documentation,
@@
 reflected in which Debian distribution they are based on.  The first stack
-is based on `debian:testing`, the second, more recently introduced stack
+is based on `debian:testing`. The second, more recently introduced stack
 is based only on Debian stable releases.  Rocker images always point to specific
@@
 recent stable version.   The different Rocker stacks have different
-aims and thus provide different images, as shown in Tables 1 & 2.
+aims and thus provide different images, as shown in Tables 1 & 2 below.
 
@@
 and packages can easily be installed from `apt`.  R packages not available
-in the `apt` repositories can be installed directly from CRAN using the
-`littler` scripts, as described in <https://rocker-project.org/use>.
+in the `apt` repositories can be installed directly from CRAN using either `R` or 
+the `littler` scripts, as described in <https://rocker-project.org/use>.
 
-This stack also include the images `shiny` and `rstudio:testing` that
+This stack also includes the images `shiny` and `rstudio:testing` that
 provide Shiny server and RStudio\textsuperscript{\textregistered} server IDE from RStudio\textsuperscript{\textregistered} Inc, built on the
@@
 Table: The `rocker-versioned` stack of images
+
+<!-- DE The (+46,000 as hadleyverse) is visually very awkward. Can the "+46k" be a footnote? I know nothing about Markdown tables but would be happy to redo 1 + 2 in latex ...-->
 
@@
 be installed from the `apt` repositories, occasionally a more recent
-version must be compiled from source (e.g. the Gibbs Sampling library,
-JAGS, and the geospatial toolkit GDAL, must both be compiled from source
+version must be compiled from source (e.g. the Gibbs Sampling program
+JAGS \citep{jags}, and the geospatial toolkit GDAL, must both be compiled from source
 on `debian:jessie` images).  In this stack, users should avoid installing R
@@
 an open and community-driven process.  This structure of the Rocker project has evolved
-over three years of operation while drawing in an ever widening base of academic researchers, 
+over three years of operation while drawing in an ever-widening base of academic researchers, 
 university instructors and industry users.  We believe this overview will be instructive

### 2017-10-07 manuscript.Rmd
Adds Dirk's suggestions:
1. Moves "portability" section below "interfaces" & "sandboxed"

2. Moves comment about singularity to later in that section and clarifies relevance. Drops shifter reference.

3. use footnote on download stats for tidyverse (& updates stats again)

@@
 
-### Portability: From laptop to cloud
-
-One common use case for Rocker containers is to provide a fast and
-reliable mechanism to deploy a custom R environment to a remote server, such
-as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers \citep{jetstream}, or private or institutional server hardware. 
-Lawrence Berkeley National Labs (LBNL) has made `Singularity`: software which can import
-and run Docker containers without root privileges, as most cluster administrators require \citep{singularity}.
-Another LBNL project, *Shifter*, allows Docker containers to be deployed on supercomputing
-architectures such as CRAY \citep{shifter}.
-
-Importantly, Rocker containers are just as easy to run locally on most modern laptops
-by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
-operating systems. By sharing volumes with the local host, users can still manipulate
-files with familiar, native tools while performing computation through a reproducible,
-containerized environment \citep{Boettiger2015}. Being
-able to test code in a predictable, pre-configured R environment on a
-local machine and to then run the same code in an identical environment on a
-remote server (e.g. for access to greater RAM, more processors, or merely
-to free up the local machine from a long-running computation) is essential
-for low-friction scaling of analysis.  Without such containerization,
-getting code to run appropriately in a remote environment can be a major
-undertaking, requiring both time and knowledge many would-be users may
-not have.
-
-For instance, the following bash commands install Docker on almost any 
-Linux-based server and then launch a Rocker container providing the 
-RStudio-server environment over a web interface.
-
-    wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
-
-The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
-the RStudio default. Many academic and commercial cloud providers make it
-possible to execute such code snippets when a container is launched, without
-ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
-into a browser and entering the
-appropriate password.  This provides the user with a familiar,
-interactive environment running on a remote machine while requiring
-a minimum of expertise.
-
-This portability is also valuable in an instructional context. Requiring students
-to install all necessary software on personal laptops can be particularly challenging
-for short workshops, where download and installation time and troubleshooting across
-heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
-By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
-an instructor can easily provide all students access to the pre-configured software environment
-using only the browser on their laptops. This strategy has proven
-effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
-UC Berkeley \citep{data8}.
-
 ### Interfaces
@@
 a typical host can easily support 100s of containers \citep{what-docker}.
+
+### Portability: From laptop to cloud
+
+One common use case for Rocker containers is to provide a fast and
+reliable mechanism to deploy a custom R environment to a remote server, such
+as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
+Jetstream servers \citep{jetstream}, or private or institutional server hardware.
+Rocker containers are  also easy to run locally on most modern laptops
+by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
+operating systems. By sharing volumes with the local host, users can still manipulate
+files with familiar, native tools while performing computation through a reproducible,
+containerized environment \citep{Boettiger2015}. Being
+able to test code in a predictable, pre-configured R environment on a
+local machine and to then run the same code in an identical environment on a
+remote server (e.g. for access to greater RAM, more processors, or merely
+to free up the local machine from a long-running computation) is essential
+for low-friction scaling of analysis.  Without such containerization,
+getting code to run appropriately in a remote environment can be a major
+undertaking, requiring both time and knowledge many would-be users may
+not have.
+
+For instance, the following bash commands install Docker on almost any 
+Linux-based server and then launch a Rocker container providing the 
+RStudio-server environment over a web interface.
+
+    wget -qO- https://get.docker.com/ | sh 
+    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+
+The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
+the RStudio default. Many academic and commercial cloud providers make it
+possible to execute such code snippets when a container is launched, without
+ever needing to `ssh` into the machine. The user may log into the server merely
+by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
+into a browser and entering the
+appropriate password.  This provides the user with a familiar,
+interactive environment running on a remote machine while requiring
+a minimum of expertise.
+
+Deploying Docker containers on centrally administrated multi-user machines, such
+as department or university clusters, has previously been more difficult, since
+system adminstrators do not want to allow the elevated user permissions the Docker
+runtime enviornment requires.  To work around this problem, Lawrence Berkeley National
+Labs (LBNL) has made `Singularity`: a container runtime environment that users can
+both install and use to run most Docker containers without requiring root privileges\citep{singularity},
+making it easier to deploy Rocker containers in this context as well.
+
+This portability is also valuable in an instructional context. Requiring students
+to install all necessary software on personal laptops can be particularly challenging
+for short workshops, where download and installation time and troubleshooting across
+heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
+By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
+an instructor can easily provide all students access to the pre-configured software environment
+using only the browser on their laptops. This strategy has proven
+effective in our own experience in both workshops and semester-length courses. Similar Docker-based
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+UC Berkeley \citep{data8}.
 
@@
 ---------------- | ----------------------------------------- | ------ | ------- 
-[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 2,000
-[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 285,000
-[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 14,000 (+46,000 as `hadleyverse`)
-[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 4,000
-[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 1,000
+[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 6,000
+[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 314,000
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 83,000 [^1]
+[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 9,000
+[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 4,000
 
@@
 
-<!-- DE The (+46,000 as hadleyverse) is visually very awkward. Can the "+46k" be a footnote? I know nothing about Markdown tables but would be happy to redo 1 + 2 in latex ...-->
+[^1]: Includes 49,000 under eariler name, `hadleyverse`.
 

### 2017-10-07 manuscript.Rmd
removed a few old (HTML) comments

@@
 ---
-
-<!-- Notes to Dirk: 
-     Some of this material may be more appropriate for documentation elsewhere, appearing here in only abbreviated form.  
-     To get started, I've just tried to err on the side of comprehensive, particularly since our documentation in places is sparse
-     or dated.  Feel free to flag any areas that should be moved in part or whole to documentation.  -->
 
@@
 is required to launch Docker containers, though not to access containers already
-<!-- DE False. On Linux the dockerd has the correct rights, and I (as member of group Docker) 
-can access docker without it.  That is important: minimal sudo use!  -->
 running and providing some service such as RStudio. Users logging into a container
@@
 A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \cite{Boettiger2015}.
-<!--
-however, `packrat` should not be necessary
-for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
--->
 
@@
 1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
-<!-- DE 1.) would be true of stable too, no?   -->
-<!-- CB  those sources are sometimes too old; e.g. jags sources on jessie are incompatible with rjags, etc -->
 2. These Dockerfiles always install the most recent available software

### 2017-10-07 manuscript.Rmd
one typo leading to one sentence changed
Keep original order, but add a just works intro paragraph.

--- 2017-10-07 manuscript.Rmd
@@
 
-   Docker permits convenient access to pre-configured and create
-   _binary_ images that "just work".  This allows R users to access a
-   wider-variety of ready-to-use environments than the R Project
-   itself provides.  For example, R users on Windows may run RStudio
-   Server or Shiny Server locally just by launching a single command
-   (once Docker itself is installed), access R-devel without affecting
-   their local system and more.
-
 nd then start the para with 'Another' instead on one?  There is a
@@
 -->
+
+Docker permits convenient access to pre-configured and create
+_binary_ images that "just work."  This allows R users to access a
+wider-variety of ready-to-use environments than the R Project
+itself provides.  For example, R users on Windows may run RStudio
+Server or Shiny Server locally just by launching a single command
+(once Docker itself is installed), access R-devel without affecting
+their local system and more.  Here, we detail the some of the principle
+use cases motivate these containerized versions of R environments
+and the design principles that help make them work.
+
+### Portability: From laptop to cloud
+
+One common use case for Rocker containers is to provide a fast and
+reliable mechanism to deploy a custom R environment to a remote server, such
+as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
+Jetstream servers \citep{jetstream}, or private or institutional server hardware.
+Rocker containers are  also easy to run locally on most modern laptops
+by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
+operating systems. By sharing volumes with the local host, users can still manipulate
+files with familiar, native tools while performing computation through a reproducible,
+containerized environment \citep{Boettiger2015}. Being
+able to test code in a predictable, pre-configured R environment on a
+local machine and to then run the same code in an identical environment on a
+remote server (e.g. for access to greater RAM, more processors, or merely
+to free up the local machine from a long-running computation) is essential
+for low-friction scaling of analysis.  Without such containerization,
+getting code to run appropriately in a remote environment can be a major
+undertaking, requiring both time and knowledge many would-be users may
+not have.
+
+For instance, the following bash commands install Docker on almost any 
+Linux-based server and then launch a Rocker container providing the 
+RStudio-server environment over a web interface.
+
+    wget -qO- https://get.docker.com/ | sh 
+    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
+
+The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
+the RStudio default (adding your user to the docker group to avoids the
+nedd for a `sudo` command to call `docker`: `sudo usermod -g docker $USER`). 
+Many academic and commercial cloud providers make it
+possible to execute such code snippets when a container is launched, without
+ever needing to `ssh` into the machine. The user may log into the server merely
+by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
+into a browser and entering the appropriate password.  
+This provides the user with a familiar,
+interactive environment running on a remote machine while requiring
+a minimum of expertise.
+
+Deploying Docker containers on centrally administrated multi-user machines, such
+as department or university clusters, has previously been more difficult, since
+system adminstrators do not want to allow the elevated user permissions the Docker
+runtime enviornment requires.  To work around this problem, Lawrence Berkeley National
+Labs (LBNL) has made `Singularity`: a container runtime environment that users can
+both install and use to run most Docker containers without requiring root privileges\citep{singularity},
+making it easier to deploy Rocker containers in this context as well.
+
+This portability is also valuable in an instructional context. Requiring students
+to install all necessary software on personal laptops can be particularly challenging
+for short workshops, where download and installation time and troubleshooting across
+heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
+By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
+an instructor can easily provide all students access to the pre-configured software environment
+using only the browser on their laptops. This strategy has proven
+effective in our own experience in both workshops and semester-length courses. Similar Docker-based
+cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+UC Berkeley \citep{data8}.
 
@@
 a typical host can easily support 100s of containers \citep{what-docker}.
-
-### Portability: From laptop to cloud
-
-One common use case for Rocker containers is to provide a fast and
-reliable mechanism to deploy a custom R environment to a remote server, such
-as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean, NSF's
-Jetstream servers \citep{jetstream}, or private or institutional server hardware.
-Rocker containers are  also easy to run locally on most modern laptops
-by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
-operating systems. By sharing volumes with the local host, users can still manipulate
-files with familiar, native tools while performing computation through a reproducible,
-containerized environment \citep{Boettiger2015}. Being
-able to test code in a predictable, pre-configured R environment on a
-local machine and to then run the same code in an identical environment on a
-remote server (e.g. for access to greater RAM, more processors, or merely
-to free up the local machine from a long-running computation) is essential
-for low-friction scaling of analysis.  Without such containerization,
-getting code to run appropriately in a remote environment can be a major
-undertaking, requiring both time and knowledge many would-be users may
-not have.
-
-For instance, the following bash commands install Docker on almost any 
-Linux-based server and then launch a Rocker container providing the 
-RStudio-server environment over a web interface.
-
-    wget -qO- https://get.docker.com/ | sh 
-    sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
-
-The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
-the RStudio default. Many academic and commercial cloud providers make it
-possible to execute such code snippets when a container is launched, without
-ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
-into a browser and entering the
-appropriate password.  This provides the user with a familiar,
-interactive environment running on a remote machine while requiring
-a minimum of expertise.
-
-Deploying Docker containers on centrally administrated multi-user machines, such
-as department or university clusters, has previously been more difficult, since
-system adminstrators do not want to allow the elevated user permissions the Docker
-runtime enviornment requires.  To work around this problem, Lawrence Berkeley National
-Labs (LBNL) has made `Singularity`: a container runtime environment that users can
-both install and use to run most Docker containers without requiring root privileges\citep{singularity},
-making it easier to deploy Rocker containers in this context as well.
-
-This portability is also valuable in an instructional context. Requiring students
-to install all necessary software on personal laptops can be particularly challenging
-for short workshops, where download and installation time and troubleshooting across
-heterogeneous machines can prove time consuming and frustrating for students and instructors alike.
-By deploying a Rocker image or Rocker-derived image (see *Extensibility*) on a cloud machine,
-an instructor can easily provide all students access to the pre-configured software environment
-using only the browser on their laptops. This strategy has proven
-effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
-UC Berkeley \citep{data8}.
 

### 2017-10-07 manuscript.Rmd

--- 2017-10-07 manuscript.Rmd
@@
 
-[^1]: Includes 49,000 under eariler name, `hadleyverse`.
+[^1]: This figure includes 49,000 downloads under the earlier name `hadleyverse`.
 

### 2017-10-07 manuscript.Rmd
drop stale comment block

@@
 # Design principles & use cases
-
-<!--  The main point here is that these are generic features that could be goals 
-      of any effort to provide researchers with virtual computing environments.
-
-- **Portable**: from laptop to cloud
-- **Interfaces**: Console-based (R, bash, etc) or RStudio
-- **Sandboxed**: experiment without touching local system
-- **Transparent**: Dockerfiles & automated builds provide a transparent and verifiable record of precisely what is installed.
-- **Optimized**: community-tweaked configurations
-- **Extensible**: interactively and persistent community extensions
-- **Versioned**: reproducible environments
-
--->
-
-<!-- DE: It is a little weird to have this first as it is a somewhat
-advanced, and I'd conjecture "proportionally speaking, "rare" use.
-Isn't the main use to just "deploy an application" _locally_? I how we
-hand users R, or R-devel, or RStudio, or Tibbliesverse ?
-
-By the same token, it is a bit weird to intro Singularity _here_ and I
-still don't know if Shifter really needs to be in the paper (cf
-Patrick's comments)
-
-So should we maybe insert a really short para like this
-
-   Deployment: Ready-to-run binaries
-
-nd then start the para with 'Another' instead on one?  There is a
-little overlap with 'Sandboxed' below but I think it is worth it to
-start with "first things first" here
-
--->
 

### 2017-10-07 manuscript.Rmd
paras merged

@@
 
-Docker permits convenient access to pre-configured and create
-_binary_ images that "just work."  This allows R users to access a
-wider-variety of ready-to-use environments than the R Project
-itself provides.  For example, R users on Windows may run RStudio
-Server or Shiny Server locally just by launching a single command
-(once Docker itself is installed), access R-devel without affecting
-their local system and more.  Here, we detail the some of the principle
-use cases motivate these containerized versions of R environments
-and the design principles that help make them work.
+Docker gives users very convenient access to pre-configured and
+pre-built _binary_ images that "just work".  This allows R users to
+access a wider-variety of ready-to-use environments than provided by
+either the R Project itself or, say, their distribution which will
+generally focus on one (current) release.  For example, R users on
+Windows may run RStudio Server or Shiny Server locally just by launching
+a single command (once Docker itself is installed). Another common
+use-case is access to R-devel without affecting the local system and
+more. Here, we detail the some of the principle use cases motivate these
+containerized versions of R environments and the design principles that
+help make them work.
 

### 2017-10-09 boettiger-eddelbuettel.Rmd
rename source files to match R-Journal reqs

--- 2017-10-07 manuscript.Rmd
@@
 functions available.  A Docker "image" is a binary archive of that software,
-analogous to an R binary package -- a given version need be downloaded only
-once and then can be "run" to create a container whenever it is needed.
+analogous to an R binary package: a given version is downloaded only
+once, and can then be "run" to create a container whenever it is needed.
 A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
@@
 a single command (once Docker itself is installed). Another common
-use-case is access to R-devel without affecting the local system and
-more. Here, we detail the some of the principle use cases motivate these
-containerized versions of R environments and the design principles that
+use-case is access to R-devel without affecting the local system.
+Here, we detail some of the principle use cases motivating these
+containerized versions of R environments, and the design principles that
 help make them work.
@@
 local machine and to then run the same code in an identical environment on a
-remote server (e.g. for access to greater RAM, more processors, or merely
+remote server (_e.g._, for access to greater RAM, more processors, or merely
 to free up the local machine from a long-running computation) is essential
@@
 the RStudio default (adding your user to the docker group to avoids the
-nedd for a `sudo` command to call `docker`: `sudo usermod -g docker $USER`). 
+need for a `sudo` command to call `docker`: `sudo usermod -g docker $USER`). 
 Many academic and commercial cloud providers make it
@@
 ever needing to `ssh` into the machine. The user may log into the server merely
-by pasting its IP address or DNS name (followed by the chosen port, e.g. `:8787`)
-into a browser and entering the appropriate password.  
-This provides the user with a familiar,
+by pasting its IP address or DNS name (followed by the chosen port, _e.g._, `:8787`)
+into a browser and entering the appropriate password. This provides the user with a familiar,
 interactive environment running on a remote machine while requiring
@@
 system adminstrators do not want to allow the elevated user permissions the Docker
-runtime enviornment requires.  To work around this problem, Lawrence Berkeley National
-Labs (LBNL) has made `Singularity`: a container runtime environment that users can
-both install and use to run most Docker containers without requiring root privileges\citep{singularity},
+runtime environment requires.  To work around this problem, Lawrence Berkeley National
+Labs (LBNL) has made 'Singularity' \citep{singularity}: a container runtime environment that users can
+both install and use to run most Docker containers without requiring root privileges,
 making it easier to deploy Rocker containers in this context as well.
@@
 effective in our own experience in both workshops and semester-length courses. Similar Docker-based
-cloud deployments have been scaled to courses of 100s of students, e.g. at Duke \citep{Mine} and
+cloud deployments have been scaled to courses of 100s of students, _e.g._, at Duke \citep{Mine} and
 UC Berkeley \citep{data8}.
@@
 computing workflows for R users have usually required the use of `ssh` and 
-a terminal only interface, posing a challenge for interactive graphics 
+a terminal-only interface, posing a challenge for interactive graphics 
 and a barrier to users unfamiliar with these tools and environments. Accessing
@@
 
-<!--  Reserve these details for documentation, reference here? --> 
 As with any interactive Docker container, users
@@
 with interactive as `-ti`) flags, and specify the desired executable
-environment (e.g. `R`, though other common options may be `Rscript` or
+environment (_e.g._, `R`, though other common options may be `Rscript` or
 `bash`):
@@
 container should be removed when the interactive session is finished.
-
 Details on sharing volumes, managing user permissions, and more can be found
@@
 installation goes poorly.  However, upgrading packages and/or the R
-environment is often necessary to run analysis of a colleague or access
+environment is often necessary to run analysis of a colleague, or access
 more recent methods.  Rocker offers an easy solution. For 
@@
 runtime environment, Singularity \citep{singularity}.  Unlike traditional virtual machines,
-these containers do not impose a large footprint of reserved resources --
+these containers do not impose a large footprint of reserved resources as
 a typical host can easily support 100s of containers \citep{what-docker}.
@@
 behavior of code, introduce errors or otherwise alter previous
-results.  The versioned stack, (`r-ver`, `rstudio`, `tidyverse`,
-`verse`, and `geospatial`), provides images which are intended to
+results.  The versioned stack (`r-ver`, `rstudio`, `tidyverse`,
+`verse`, and `geospatial`) provides images which are intended to
 build an identical software stack every time, regardless of the
@@
 version tag in the Docker image name to request a version stable
-image, e.g. `rocker/verse:3.4.0`.  If no tag is explicitly requested,
+image, _e.g._, `rocker/verse:3.4.0`.  If no tag is explicitly requested,
 Docker will provide the image with the tag `:latest`, which will
@@
 associated with the most recent date for which that image was current.  This ensures that
-a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
-available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
-can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
+a Dockerfile building `FROM rocker/verse:3.4.1` will only install R package versions that were
+available on CRAN on 2017-06-30, that is, the day R 3.4.1 was released.  This default
+can of course be overwritten in the standard R manner, _e.g._, by specifying a different CRAN mirror explicitly in any 
 command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
-in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
+in an `.Rprofile`.  Note that the MRAN date associated with the current release (_e.g._, `3.4.2` at the time of
 writing) will continue to advance on the Docker-hub image until the next R release.  Software
@@
 and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
-using `bioclite()` utility will also install the version appropriate to the version of R found on the system
+using the `bioclite()` utility will also install the version appropriate to the version of R found on the system
 (the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
@@
 an image for reproducing a data analysis will probably select a stable version tag
-in addition to an appropriate base library, e.g.: `FROM rocker/tidyverse:3.4.1`
+in addition to an appropriate base library, _e.g._,: `FROM rocker/tidyverse:3.4.1`
 Users can easily add additional software to any running Rocker image using
@@
 This coalescence helped create a more fine-tuned image with broad support 
-for a wide range of commonly used data formats and libraries.  Other community
+for a wide range of commonly-used data formats and libraries.  Other community
 images are developed and maintained independently of the Rocker project,
@@
 for everyone, but today reflects the considerable community input and
-testing over past few years.
+testing over the past few years.
 
@@
 of the "official" images maintained by Docker's own development team.
-The Debian platform is also perhaps the best supported Linux platform
+The Debian platform is also perhaps the best-supported Linux platform
 within the R community, including an active `r-sig-debian` listserve.
@@
 (roughly two years recently) means that software in the Debian stable 
-(e.g. `debian:jessie`, `debian:stretch`)
+(_e.g._, `debian:jessie`, `debian:stretch`)
 releases can lag significantly behind current releases of popular
@@
 reflected in which Debian distribution they are based on.  The first stack
-is based on `debian:testing`. The second, more recently introduced stack
+is based on `debian:testing`. The second, more recently-introduced stack,
 is based only on Debian stable releases.  Rocker images always point to specific
@@
 
-The `debian:testing` stack aims to makes the most efficient use of
-upstream builds -- the precompiled `.deb` binaries provided by the
+The `debian:testing` stack aims to make the most efficient use of
+upstream builds: the pre-compiled `.deb` binaries provided by the
 Debian repositories.  It is both quicker and easier to install software
@@
 popular R packages can also be installed through the package manager,
-e.g. `apt-get install r-cran-xml`.  This can be particularly helpful
+_e.g._, `apt-get install r-cran-xml`.  This can be particularly helpful
 for packages with external system dependencies (such as `libxml2-dev`
@@
 We should note, however, that only about 500 of the over
-11000 CRAN packages are available as Debian packages.
+11,000 CRAN packages are available as Debian packages.
 
@@
 are sent to `unstable` during the normal course of Debian development.
-This can occasionally break an previously working installation command
+This can occasionally break an previously-working installation command
 in a Dockerfile until the maintainer redirects the package manager to
@@
 be installed from `testing`, or vice versa (using the `-t` option in
-`apt`)  That said, packages only migrate from 
+`apt`).  That said, packages only migrate from 
 `unstable` to `testing` after a period of several days---and if the 
@@
 
-1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
-2. These Dockerfiles always install the most recent available software
-3. These Dockerfiles can almost always build relatively quickly
+1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager.
+2. These Dockerfiles always install the most recent available software.
+3. These Dockerfiles can almost always build relatively quickly.
 
@@
 
-1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (e.g. changing an installation directive from `unstable` to `testing` or vice versa)
+1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (_e.g._, changing an installation directive from `unstable` to `testing` or vice versa).
 2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
@@
 the Rocker team.  Consequently, users should refer to this image in
-Docker commands without an organization namespace, e.g. `docker run -t
+Docker commands without an organization namespace, _e.g._, `docker run -t
 r-base` to access the official image.  All other images in the Rocker
@@
 be referenced using the `rocker` namespace,
-e.g. `docker run -ti rocker/r-devel`.
+_e.g._, `docker run -ti rocker/r-devel`.
 
 Several of the images in this stack are oriented towards the R
-development community. `r-devel`, `drd`, `r-devel-san` &
+development community: `r-devel`, `drd`, `r-devel-san` &
 `r-devel-ubsan-clang` all add a copy of the development version of R
@@
 with the then-contemporaneous versions of R packages.  A user specifies the
-version desired using an image tag, e.g.  `rocker/r-ver:3.3.1` will
+version desired using an image tag, _e.g._,  `rocker/r-ver:3.3.1` will
 refer to an image with R version 3.3.1 installed.  Omitting the tag
@@
 the future after newer R versions are released) should explicitly include
-the corresponding version tag, e.g. `rocker/r-ver:3.4.2` at the time
+the corresponding version tag, _e.g._, `rocker/r-ver:3.4.2` at the time
 of writing, and not the `latest` tag.  Users can also run the current
@@
 **Version tags**:  The version tags are propagated throughout this
-stack: _e.g._, `rocker/tidyverse:devel` will provide the currently
-released versions of the R packages in the `tidyverse` \citep{tidyverse}
+stack: _e.g._, `rocker/tidyverse:devel` will provide the currently-released
+versions of the R packages in the `tidyverse` \citep{tidyverse}
 installed on the nightly build of R-devel.  Developers building packages
@@
 the stack at a later date, we note that the MRAN snapshots began in
-starting 2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
+2014-09-17 and thus go back only to the R `3.1` era.  Each tag must
 be built from a separate Dockerfile, enabling minor differences in the
 build instructions to accommodate changing dependencies.  Dockerfiles for
-past versions (e.g. < `3.4.0` currently) are intended to remain static
+past versions (_e.g._, prior to `3.4.2` currently) are intended to remain static
 over the long term, while the tag for the current version, `latest`,
 and `devel` may be tweaked to accommodate new features or dependencies.
-Version tags also obey semantics, so that omitting the second or third
+Version tags also obey semantics so that omitting the second or third
 position of the tag is identical to asking for the most recent version:
@@
 be installed from the `apt` repositories, occasionally a more recent
-version must be compiled from source (e.g. the Gibbs Sampling program
+version must be compiled from source (_e.g._, the Gibbs Sampling program
 JAGS \citep{jags}, and the geospatial toolkit GDAL, must both be compiled from source
 on `debian:jessie` images).  In this stack, users should avoid installing R
-packages using `apt` without careful consideration, as this will install a second, 
-probably different version
-of R from the Debian repositories and a dated version of the R package
-(since any `r-cran-pkgname` package in the Debian repositories will
-depend on `r-base` in `apt` as well).
+packages using `apt` without careful consideration as this will install a second 
+(probably different) version
+of R from the Debian repositories, and a dated version of the R package
+since any `r-cran-pkgname` package in the Debian repositories will
+depend on `r-base` in `apt` as well.
 
@@
 user settings such as login password and permissions through environmental
-variables at run time.  <!-- link wiki howto page? -->
+variables at run time.
 
@@
 of the commonly-used `tidyverse` and `devtools` R packages, including
-external database libraries (e.g. MariaDB and PostgreSQL).  Users should
+external database libraries (_e.g._, MariaDB and PostgreSQL).  Users should
 consult the package Dockerfiles or `installed.packages()` list directly
@@
 commonly-used dependencies, notably a large but not comprehensive LaTeX
-environment and Java development libraries.  Previously the Rocker
+environment and Java development libraries.  Previously, the Rocker
 project provided the image `hadleyverse` which was since divided into
@@
 by using the `--build-arg` option of `docker build`.   In the `r-ver`
-image, users can set `R_VERSION`, `BUILD_DATE` (MRAN default snapshot).
+image, users can set `R_VERSION` and `BUILD_DATE` (MRAN default snapshot).
 In the `rstudio` image users can set `RSTUDIO_VERSION` (otherwise defaults
@@
 indicating image `license` (GPL-2.0), `vcs-url` (GitHub repository), and `vendor` (Rocker Project).  These metadata can be altered or extended in downstream images.
-
-<!--
-
-### Maintenance and Updates
-
-These images are actively maintained.
-This means that while an effort is made to preserve the general function of these images over time, both these Dockerfiles and the resulting images are subject to some change over time.
-In particular:
-
-- Images are regularly re-built on Docker Hub whenever their base image changes, starting with changes to `debian` Docker image.
-This is the rough equivalent of running `apt-get upgrade` on `debian`, since all `apt-get` commands are re-run and will pull in the most current sources.
-This allows the images to receive security updates to any packages installed from the `debian` repositories, but will not in general change the versions of any software and is very unlikely to break anything.
-
-- The Dockerfiles themselves are subject to change, to improve performance, ease of use, readability, or other concerns raised in the issues.
-These changes should also not alter the general behavior of R or R packages on the image.
-These changes can be seen in the git history.
-The [rocker-versioned](https://github.com/rocker-org/rocker-versioned) repo will use its own semantic version tagging to indicate changes to this repository, with snapshots from these tags archived on Zenodo.
-
-### License
-
-The Dockerfiles in this repository are licensed under the GPL 2 or later.
-
--->
 

### 2017-10-09 boettiger-eddelbuettel.tex
rename source files to match R-Journal reqs

% !TeX root = RJwrapper.tex
\title{An Introduction to Rocker:\\
Docker Containers for R}
\author{by Carl Boettiger, Dirk Eddelbuettel}

\maketitle

\abstract{%
We describe the Rocker project, which provides a widely-used suite of
Docker images with customized R environments for particular tasks. We
discuss how this suite is organized, and how these tools can increase
portability, scaling, reproducibility, and convenience of R users and
developers.
}

\section{Introduction}\label{introduction}

The Rocker project was launched in October 2014 as a collaboration
between CB \& DE to provide high-quality Docker images containing the R
environment \citep{edd2014}. Since that time, the project has seen both
considerable uptake in the community and substantial development and
evolution. Here we seek to document the project's objectives and uses.

\subsubsection{What is Docker?}\label{what-is-docker}

Docker is a popular open-source tool to create, distribute, deploy, and
run software applications using \emph{containers}. Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common
virtual environments) requiring all operating-system components an
application needs to run: code, runtime, system tools, system runtime.
Docker containers are lightweight as they share the operating system
kernel, starting instantly using a layered filesystem which minimizes
disk footprint and download time, are built on open standards that run
on all major platforms (Linux, Mac, Windows), and provide an added layer
of security by running an application in an isolated environment
\citep{what-docker}. Familiarity with a few key terms is helpful in
understanding this paper. The term ``container'' refers to an isolated
software environment on a computer. R users can think of running a
container as analogous to loading an R package; it makes a set of
software functions available. A Docker ``image'' is a binary archive of
that software, analogous to an R binary package: a given version is
downloaded only once, and can then be ``run'' to create a container
whenever it is needed. A ``Dockerfile'' is a recipe, the source-code, to
create a Docker image. Development and contributions to the Rocker
project focuses on the construction, organization and maintenance of
these Dockerfiles.

\section{Design principles \& use
cases}\label{design-principles-use-cases}

Docker gives users very convenient access to pre-configured and
pre-built \emph{binary} images that ``just work''. This allows R users
to access a wider-variety of ready-to-use environments than provided by
either the R Project itself or, say, their distribution which will
generally focus on one (current) release. For example, R users on
Windows may run RStudio Server or Shiny Server locally just by launching
a single command (once Docker itself is installed). Another common
use-case is access to R-devel without affecting the local system. Here,
we detail some of the principle use cases motivating these containerized
versions of R environments, and the design principles that help make
them work.

\subsubsection{Portability: From laptop to
cloud}\label{portability-from-laptop-to-cloud}

One common use case for Rocker containers is to provide a fast and
reliable mechanism to deploy a custom R environment to a remote server,
such as Amazon Web Services Elastic Compute (AWS EC2), DigitalOcean,
NSF's Jetstream servers \citep{jetstream}, or private or institutional
server hardware. Rocker containers are also easy to run locally on most
modern laptops by first installing the appropriate Docker command (or
distribution) for Windows, MacOS, and Linux-based operating systems. By
sharing volumes with the local host, users can still manipulate files
with familiar, native tools while performing computation through a
reproducible, containerized environment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment
on a remote server (\emph{e.g.}, for access to greater RAM, more
processors, or merely to free up the local machine from a long-running
computation) is essential for low-friction scaling of analysis. Without
such containerization, getting code to run appropriately in a remote
environment can be a major undertaking, requiring both time and
knowledge many would-be users may not have.

For instance, the following bash commands install Docker on almost any
Linux-based server and then launch a Rocker container providing the
RStudio-server environment over a web interface.

\begin{verbatim}
wget -qO- https://get.docker.com/ | sh 
sudo docker run -p 8787:8787 -e PASSWORD=<PICK-A-PASSWORD> rocker/rstudio
\end{verbatim}

The \texttt{docker\ run} option \texttt{-p} sets the port on which
RStudio will appear, 8787 is the RStudio default (adding your user to
the docker group to avoids the need for a \texttt{sudo} command to call
\texttt{docker}: \texttt{sudo\ usermod\ -g\ docker\ \$USER}). Many
academic and commercial cloud providers make it possible to execute such
code snippets when a container is launched, without ever needing to
\texttt{ssh} into the machine. The user may log into the server merely
by pasting its IP address or DNS name (followed by the chosen port,
\emph{e.g.}, \texttt{:8787}) into a browser and entering the appropriate
password. This provides the user with a familiar, interactive
environment running on a remote machine while requiring a minimum of
expertise.

Deploying Docker containers on centrally administrated multi-user
machines, such as department or university clusters, has previously been
more difficult, since system adminstrators do not want to allow the
elevated user permissions the Docker runtime environment requires. To
work around this problem, Lawrence Berkeley National Labs (LBNL) has
made `Singularity' \citep{singularity}: a container runtime environment
that users can both install and use to run most Docker containers
without requiring root privileges, making it easier to deploy Rocker
containers in this context as well.

This portability is also valuable in an instructional context. Requiring
students to install all necessary software on personal laptops can be
particularly challenging for short workshops, where download and
installation time and troubleshooting across heterogeneous machines can
prove time consuming and frustrating for students and instructors alike.
By deploying a Rocker image or Rocker-derived image (see
\emph{Extensibility}) on a cloud machine, an instructor can easily
provide all students access to the pre-configured software environment
using only the browser on their laptops. This strategy has proven
effective in our own experience in both workshops and semester-length
courses. Similar Docker-based cloud deployments have been scaled to
courses of 100s of students, \emph{e.g.}, at Duke \citep{Mine} and UC
Berkeley \citep{data8}.

\subsubsection{Interfaces}\label{interfaces}

An important aspect of the Rocker project design is the ability for
users to interact with the software on the container through either an
interactive shell session (such as the R shell or a bash shell), or
through a web browser accessing the
RStudio\textsuperscript{\textregistered} Server integrated development
environment (IDE). Traditional remote and high-performance computing
workflows for R users have usually required the use of \texttt{ssh} and
a terminal-only interface, posing a challenge for interactive graphics
and a barrier to users unfamiliar with these tools and environments.
Accessing an RStudio\textsuperscript{\textregistered} container through
the browser removes these barriers. Rocker images include the
RStudio-server software pre-installed and configured with the explicit
permission of RStudio\textsuperscript{\textregistered} Inc.

Users can access a root \texttt{bash} shell in a running Rocker
container using

\begin{verbatim}
docker exec -ti <container-id> bash
\end{verbatim}

which can be useful for administrative tasks such as installing system
dependencies. All Rocker images can also be run as an interactive R,
RScript or bash shell without running RStudio, which can be useful for
batch jobs or for anyone who prefers that environment.

As with any interactive Docker container, users should specify the
interactive (\texttt{-i}) and terminal (\texttt{-t}, here combined with
interactive as \texttt{-ti}) flags, and specify the desired executable
environment (\emph{e.g.}, \texttt{R}, though other common options may be
\texttt{Rscript} or \texttt{bash}):

\begin{verbatim}
docker run --rm -ti rocker/tidyverse R
\end{verbatim}

This example shows the use of the \texttt{-\/-rm} flag to indicate that
the container should be removed when the interactive session is
finished. Details on sharing volumes, managing user permissions, and
more can be found on the Rocker website,
\url{https://rocker-project.org}.

\subsubsection{Sandboxed}\label{sandboxed}

Another feature of Rocker containers is the ability to provide a
sandboxed environment, isolated from software and potentially from other
data on the machine. Many users are reluctant to upgrade their suite of
installed packages, which may break their existing code or even their R
environment if the installation goes poorly. However, upgrading packages
and/or the R environment is often necessary to run analysis of a
colleague, or access more recent methods. Rocker offers an easy
solution. For instance, a user can run R code requiring the most recent
versions of R and related packages inside a Rocker container without
having to upgrade their local installations first. Conversely, one could
use Rocker to run code on an older R release with prior versions of R
packages, again without having to make any alteration to one's local R
install. Another common use case is to access a container with support
for particular options such as using gcc or clang compiler sanitizers
\citep{edd_sanitizers}. These requires R itself be built with
specialized settings that may not be not available or familiar to many R
users on their native system, but can be easily deployed by pulling the
Rocker images \texttt{rocker/r-devel-san} or
\texttt{rocker/r-devel-ubsan-clang}.

This sandboxing feature is also valuable in the remote computing
context, allowing system administrators to grant users freedom to
install software which requires root privileges inside a container,
while not granting them root access on the host machine. Root access is
required to launch Docker containers, though not to access containers
already running and providing some service such as RStudio. Users
logging into a container through the
RStudio\textsuperscript{\textregistered} interface do not by default
have root privileges, though are able to install R packages. Granting
these users root privileges in the container still leaves them sandboxed
from the host container. Further, Docker images can be deployed from a
machine without root privileges through the alternate container runtime
environment, Singularity \citep{singularity}. Unlike traditional virtual
machines, these containers do not impose a large footprint of reserved
resources as a typical host can easily support 100s of containers
\citep{what-docker}.

\subsubsection{Transparent}\label{transparent}

Users can easily determine the software stack installed on any Rocker
image by examining the associated Dockerfile recipe, which provides a
concise, human-readable record of the installation. All Rocker images
use automated builds through Docker Hub, which also acts as the central,
default repository distributing the images. Using automated builds
rather than uploading pre-built image binaries to the Docker Hub avoids
the potential for the build not to match the recipe. The corresponding
Dockerfile is visible both on the Docker Hub and in the linked GitHub
repository of the Rocker project (\url{https://github.com/rocker-org}),
which provides a transparent versioned history of all changes made to
these recipes, as well as documentation, a community wiki, and issue
trackers for discussing proposed changes, bugs, improvements to the
Dockerfiles and troubleshoot any issues users may encounter. Having
these public source files built automatically by a trusted provider
(Docker Hub), rather than built locally and uploaded only in binary form
provides is also useful from a security perspective in avoiding malware.

\subsubsection{Community Optimized}\label{community-optimized}

Having a shared, transparent computational environment created by a
publicly hosted, reproducible recipe facilitates community input into
configuration details. R and many of its packages and related software
can be configured with a wide range of options, compilers, different
linear-algebra libraries and so forth. While this flexibility reflects
varying needs, many users rely on default settings which may be
optimized more for simplicity of installation than performance. The
Rocker recipes reflect significant community input on these choices, as
well as the considerable experience and expertise of the R Debian
maintainer of 20 years in what configuration options to use. This helps
create a more finely tuned, optimized reference implementation of the R
environment as well as a platform for comparing and discussing these
concerns which are often overlooked elsewhere. Issues and Pull Requests
on the Rocker repositories on GitHub attest to these discussions and
improvements. Widespread use of the Rocker image helps promote both
testing of these choices and contributions further tweaking the
configuration from many members of the R community.

\subsubsection{Versioned}\label{versioned}

Access to specific versions of software can be important for users who
need computational reproducibility more than having the latest release
of any piece of software, since subsequent releases can alter the
behavior of code, introduce errors or otherwise alter previous results.
The versioned stack (\texttt{r-ver}, \texttt{rstudio},
\texttt{tidyverse}, \texttt{verse}, and \texttt{geospatial}) provides
images which are intended to build an identical software stack every
time, regardless of the release of new libraries and packages. Users
should specify an R version tag in the Docker image name to request a
version stable image, \emph{e.g.}, \texttt{rocker/verse:3.4.0}. If no
tag is explicitly requested, Docker will provide the image with the tag
\texttt{:latest}, which will always have the latest available versions
of the software (built nightly).

Users building on the version-tagged images will by default use the MRAN
snapshot mirror \citep{MRAN} associated with the most recent date for
which that image was current. This ensures that a Dockerfile building
\texttt{FROM\ rocker/verse:3.4.1} will only install R package versions
that were available on CRAN on 2017-06-30, that is, the day R 3.4.1 was
released. This default can of course be overwritten in the standard R
manner, \emph{e.g.}, by specifying a different CRAN mirror explicitly in
any command to \texttt{install.packages()}, or by adjusting the default
CRAN mirror in
\texttt{options(repo=\textless{}CRAN-MIRROR\textgreater{})} in an
\texttt{.Rprofile}. Note that the MRAN date associated with the current
release (\emph{e.g.}, \texttt{3.4.2} at the time of writing) will
continue to advance on the Docker-hub image until the next R release.
Software installed from \texttt{apt-get} in these images will come from
the the stable Debian release (\texttt{stretch} or \texttt{jessie}) and
thus not change versions (though it will receive security patches).
Packages installed from BioConductor using the \texttt{bioclite()}
utility will also install the version appropriate to the version of R
found on the system (the Bioconductor semi-annual release model avoids
the need for an MRAN mirror). Users installing packages from GitHub or
other sources can request a specific git release tag or hash for a more
reproducible build, or adopt an alternative approach such as
\texttt{packrat} \citep{packrat}. A more general discussion of the use
and limitations of Docker for computational reproducibility can be found
in \cite{Boettiger2015}.

\subsubsection{Extensible}\label{extensible}

Any portable computational environment faces an inevitable tension
between the ``kitchen sink problem'' at one extreme, and the ``discovery
problem'' on the other. A kitchen sink image seeks to accommodate too
many use cases in a single image. Such images are inevitably very large
and thus slow or difficult to deploy, maintain and optimize. At the
other extreme, providing too many specialized images makes it more
difficult for a user to discover the one they need. The Rocker project
seeks to avoid both of these problems by providing a carefully-curated
suite of images that an be easily extended by individuals and
communities.

To make extensions transparent and persistent, Rocker images can be
extended by any user by writing their own Dockerfiles based on an
appropriate Rocker image. The Dockerfiles in the Rocker stack should
themselves provide a simple of example of this. A user begins by
selecting an appropriate base image for their needs: if the
RStudio\textsuperscript{\textregistered} interface is desired, a user
might start with \texttt{FROM\ rocker/rstudio}; an image for testing a
particular C code in an R package might use
\texttt{FROM\ rocker/r-devel-san}, an image for reproducing a data
analysis will probably select a stable version tag in addition to an
appropriate base library, \emph{e.g.},:
\texttt{FROM\ rocker/tidyverse:3.4.1} Users can easily add additional
software to any running Rocker image using the standard R and Debian
mechanisms. Details on how to extend Rocker images can be found at
\url{https://rocker-project.org}.

Sharing these Dockerfiles can also facilitate the emergence of
extensions tuned to particular communities. For instance, the
\texttt{rocker/geospatial} image emerged from the input of a number of
Rocker users all adding common geospatial libraries and packages on top
of the existing Rocker images. This coalescence helped create a more
fine-tuned image with broad support for a wide range of commonly-used
data formats and libraries. Other community images are developed and
maintained independently of the Rocker project, such as the
\texttt{popgen} image of population-genetics-oriented software developed
by the National Evolutionary Synthesis Center (NESCent). Rocker images
are also being used as base Docker images in the NSF sponsored Whole
Tale project for reproducible computing \citep{wholetale}, and are
heavily used by the R-Hub project in automated package testing
\citep{rhub}.

\section{Rocker organization and
workflow}\label{rocker-organization-and-workflow}

The Rocker project consists of a suite of images built automatically by
and hosted on the Docker Hub, \url{https://hub.docker.com/r/rocker}.
Source Dockerfiles, supporting scripts and documentation are hosted on
GitHub under the organization \texttt{rocker-org},
\url{https://github.com/rocker-org}. The issue tracker and pull requests
are used for community input, discussions and contributions to these
images. The Rocker project wiki provides a place to synthesize
community-contributed documentation, use-cases, and other knowledge
about using the Rocker images.

\subsubsection{Images in the Rocker
Project}\label{images-in-the-rocker-project}

The Rocker project aims to provide a small core of Docker images that
serve as convenient `base' images on which other users can build custom
R environments by writing their own Dockerfiles, while also providing a
`batteries included' approach of images that can be used out of the box.
The challenges of balancing diverse needs driven by very different use
cases against the overarching goals of creating images that are still
sufficiently light-weight, easy to use and easy to maintain is a
difficult art. The implementation in both individual Rocker images and
image stacks can never perfect that balance for everyone, but today
reflects the considerable community input and testing over the past few
years.

All Rocker images are based on the Debian Linux distribution. The Debian
platform provides a small base image, the well-known \texttt{apt}
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many of
the ``official'' images maintained by Docker's own development team. The
Debian platform is also perhaps the best-supported Linux platform within
the R community, including an active \texttt{r-sig-debian} listserve.
The relatively long period between stable Debian releases (roughly two
years recently) means that software in the Debian stable (\emph{e.g.},
\texttt{debian:jessie}, \texttt{debian:stretch}) releases can lag
significantly behind current releases of popular software, including R.
More recent versions of packages can be found in the pre-release
distribution, \texttt{debian:testing}, while the very latest binary
builds can be found on \texttt{debian:unstable}. The Rocker project can
be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on. The first
stack is based on \texttt{debian:testing}. The second, more
recently-introduced stack, is based only on Debian stable releases.
Rocker images always point to specific stable releases (\texttt{jessie},
\texttt{stretch}), and do not use the tag \texttt{debian:stable}, which
is a rolling tag that always points to the most recent stable version.
The different Rocker stacks have different aims and thus provide
different images, as shown in Tables 1 \& 2 below.

\subsubsection{\texorpdfstring{The \texttt{debian:testing}-based
images}{The debian:testing-based images}}\label{the-debiantesting-based-images}

The \texttt{debian:testing} stack aims to make the most efficient use of
upstream builds: the pre-compiled \texttt{.deb} binaries provided by the
Debian repositories. It is both quicker and easier to install software
from binaries, since the package manager (\texttt{apt}) manages the
necessary (binary) dependencies and bypasses the time-consuming process
of compiling from source. Basing this stack on \texttt{debian:testing}
means that much more recent versions of commonly-used libraries and
compilers are available as binaries than would be found in a Debian
stable release. In order to provide access to the most recent available
binaries, this stack uses apt-pinning \citep{apt_pinning} to allow the
\texttt{apt} package manager to also install binaries from
\texttt{debian:unstable}, which represents the most recent, bleeding
edge of packages built for Debian when necessary. For instance, the
\texttt{r-base} image provided by Rocker installs the most recent
version of R as a binary from Debian \texttt{unstable}. Similarly,
recent versions of many popular R packages can also be installed through
the package manager, \emph{e.g.}, \texttt{apt-get\ install\ r-cran-xml}.
This can be particularly helpful for packages with external system
dependencies (such as \texttt{libxml2-dev} in this example) which cannot
be installed from the R console as they are system dependencies rather
than R packages installed from within R. We should note, however, that
only about 500 of the over 11,000 CRAN packages are available as Debian
packages.

As the names \texttt{testing} and \texttt{unstable} imply, this approach
is not without challenges. The particular version of any given package
can change as packages move from \texttt{unstable} into
\texttt{testing}. New versions are sent to \texttt{unstable} during the
normal course of Debian development. This can occasionally break an
previously-working installation command in a Dockerfile until the
maintainer redirects the package manager to install a package from the
\texttt{unstable} sources that could previously be installed from
\texttt{testing}, or vice versa (using the \texttt{-t} option in
\texttt{apt}). That said, packages only migrate from \texttt{unstable}
to \texttt{testing} after a period of several days---and if the
migration and installation of the particular version is free of
interactions with other packages in their dependency graph. That way,
\texttt{unstable} serves as validation lab which leaves \texttt{testing}
reasonably stable yet current.

Relative to \texttt{stable}, the \texttt{testing} stack thus offers the
following advantages:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  These Dockerfiles are easy to develop and extend because almost all
  software can be installed through the package manager.
\item
  These Dockerfiles always install the most recent available software.
\item
  These Dockerfiles can almost always build relatively quickly.
\end{enumerate}

and these down-sides:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  These Dockerfiles require occasional minor maintenance to ensure
  successful builds. (\emph{e.g.}, changing an installation directive
  from \texttt{unstable} to \texttt{testing} or vice versa).
\item
  The resulting images are inherently dynamic: rebuilding the same
  Dockerfile months or years apart will generate images with
  significantly different versions of software installed.
\item
  The use of apt pinning may be unfamiliar to some users, where the user
  must ultimately be responsible to ensure compatibility
  \citep{apt_pinning}.
\end{enumerate}

\subsubsection{Images overview}\label{images-overview}

The \texttt{debian:testing}-based stack currently includes seven images
actively maintained by the Rocker development team (Table 1).
\texttt{r-base} builds on \texttt{debian:testing}, and the other six in
the stack each build directly from \texttt{r-base}. The \texttt{r-base}
image is unique in that it is designated as the official image for the R
language by the Docker organization itself. This official image is
reviewed and then built by employees of Docker Inc based on a Dockerfile
maintained by the Rocker team. Consequently, users should refer to this
image in Docker commands without an organization namespace, \emph{e.g.},
\texttt{docker\ run\ -t\ r-base} to access the official image. All other
images in the Rocker project are not individually reviewed and built by
Docker Inc and must be referenced using the \texttt{rocker} namespace,
\emph{e.g.}, \texttt{docker\ run\ -ti\ rocker/r-devel}.

Several of the images in this stack are oriented towards the R
development community: \texttt{r-devel}, \texttt{drd},
\texttt{r-devel-san} \& \texttt{r-devel-ubsan-clang} all add a copy of
the development version of R side-by-side the current release of R
provided by \texttt{r-base}. On these images, the development version is
aliased to \texttt{RD} to distinguish from the current release,
\texttt{R}. As the names suggest, each provide slightly different
configurations. Of particular interest are the images providing
development R built with support for C/C++ address and
undefined-behavior sanitizers, which are somewhat more difficult to
configure \citep{edd_sanitizers}.

As these images focus on developers and/or as base images for custom
uses, this stack does not include many specific R packages. Additional
dependencies and packages can easily be installed from \texttt{apt}. R
packages not available in the \texttt{apt} repositories can be installed
directly from CRAN using either \texttt{R} or the \texttt{littler}
scripts, as described in \url{https://rocker-project.org/use}.

This stack also includes the images \texttt{shiny} and
\texttt{rstudio:testing} that provide Shiny server and
RStudio\textsuperscript{\textregistered} server IDE from
RStudio\textsuperscript{\textregistered} Inc, built on the
\texttt{r-base} image. RStudio\textsuperscript{\textregistered} and
Shiny are registered trademarks of RStudio Inc, and their use and the
distribution of their software in binary form on Docker Hub has been
granted to the Rocker project by explicit permission from RStudio. Users
should review RStudio's trademark use policy
(\url{http://www.rstudio.com/about/trademark/}) and address inquiries
about further distribution or other questions to
\href{mailto:permissions@rstudio.com}{\nolinkurl{permissions@rstudio.com}}.
The Rocker project also provides images with
RStudio\textsuperscript{\textregistered} server and Shiny server in the
stable versioned stack.

\textbf{Build schedule}: The official \texttt{r-base} image is rebuilt
by Docker following any updates to the official \texttt{debian} images
(roughly every few weeks). The rest of the stack uses build triggers
that rebuild the images either whenever \texttt{r-base} is updated, or
the Dockerfile sources are updated on the corresponding GitHub
repository. The only exception in this stack is the \texttt{drd} image,
which is rebuilt each week by a \texttt{cron} trigger.

\begin{longtable}[]{@{}llll@{}}
\caption{The \texttt{debian:testing} image stack}\tabularnewline
\toprule
\begin{minipage}[b]{0.20\columnwidth}\raggedright\strut
image\strut
\end{minipage} & \begin{minipage}[b]{0.50\columnwidth}\raggedright\strut
description\strut
\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
size\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\raggedright\strut
downloads\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.20\columnwidth}\raggedright\strut
image\strut
\end{minipage} & \begin{minipage}[b]{0.50\columnwidth}\raggedright\strut
description\strut
\end{minipage} & \begin{minipage}[b]{0.08\columnwidth}\raggedright\strut
size\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\raggedright\strut
downloads\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/_/r-base}{r-base}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
official image with current version of R\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
254 MB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
632,000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/r-devel}{r-devel}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
R-devel added side-by-side r-base (using alias \texttt{RD})\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1 GB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
4,000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/drd}{drd}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
lightweight r-devel, built weekly\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
571 MB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
4,000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/r-devel-san}{r-devel-san}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
as r-devel, but built with compiler sanitizers\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1.1 GB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
1,000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/r-devel-ubsan-clang}{r-devel-ubsan-clang}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
Sanitizers, clang c compiler (instead of gcc)\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1.1 GB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
525\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/r-devel-san}{rstudio:testing}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
rstudio on debian:testing\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
1.1 GB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
1,000\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\href{https://hub.docker.com/r/rocker/shiny}{shiny}\strut
\end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
shiny-server on r-base\strut
\end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
409 MB\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright\strut
123,000\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\begin{longtable}[]{@{}llll@{}}
\caption{The \texttt{rocker-versioned} stack of images}\tabularnewline
\toprule
image & description & size & downloads\tabularnewline
\midrule
\endfirsthead
\toprule
image & description & size & downloads\tabularnewline
\midrule
\endhead
\href{https://hub.docker.com/r/rocker/r-ver}{r-ver} & Version-stable
base R \& src build tools & 219 MB & 6,000\tabularnewline
\href{https://hub.docker.com/r/rocker/rstudio}{rstudio} & Adds rstudio &
334 MB & 314,000\tabularnewline
\href{https://hub.docker.com/r/rocker/tidyverse}{tidyverse} & Adds
tidyverse \& devtools & 656 MB & 83,000 \footnote{This figure includes
  49,000 downloads under the earlier name \texttt{hadleyverse}.}\tabularnewline
\href{https://hub.docker.com/r/rocker/verse}{verse} & Adds java, tex \&
publishing-related packages & 947 MB & 9,000\tabularnewline
\href{https://hub.docker.com/r/rocker/geospatial}{geospatial} & Adds
geospatial libraries & 1.3 GB & 4,000\tabularnewline
\bottomrule
\end{longtable}

\subsubsection{\texorpdfstring{The \texttt{debian:stable}-based
stack}{The debian:stable-based stack}}\label{the-debianstable-based-stack}

This stack emphasizes stability and reproducibility of the Docker build.
This stack was introduced much more recently (November 2016) in response
to considerable user input and requests. The key feature of this stack
is the ability to run older versions of R along with the
then-contemporaneous versions of R packages. A user specifies the
version desired using an image tag, \emph{e.g.},
\texttt{rocker/r-ver:3.3.1} will refer to an image with R version 3.3.1
installed. Omitting the tag is equivalent to using the tag
\texttt{latest}, which, as the name implies, will always point to an
image using the current R release. Thus, users wanting to create
downstream Dockerfiles which are based on the current release at time
(but will continue to reconstruct the same environment in the future
after newer R versions are released) should explicitly include the
corresponding version tag, \emph{e.g.}, \texttt{rocker/r-ver:3.4.2} at
the time of writing, and not the \texttt{latest} tag. Users can also run
the current development version of R using the tag \texttt{devel}, which
is built nightly from R-devel sources from \texttt{subversion}.

\textbf{MRAN archives}: To facilitate installation of only
contemporaneous versions of R packages on these images, the default CRAN
mirror from which to install R packages is fixed to a snapshot of CRAN
corresponding to the last date for which that version of R was the
latest release. These snapshots are provided by the MRAN archive created
by Revolution Analytics (now part of Microsoft). It archives daily
snapshots of all of CRAN from which a user can install packages with the
usual \texttt{install.packages()} function \citep{MRAN}. Users can
always override this default by passing any current CRAN repository
explicitly. Unlike CRAN, Bioconductor only updates its repositories
through bi-annual releases aligned to R's spring release schedule. Thus,
Bioconductor packages can be installed in the usual way using
\texttt{bioclite}, which automatically selects the Bioconductor release
corresponding to the version of R in use.

\textbf{Version tags}: The version tags are propagated throughout this
stack: \emph{e.g.}, \texttt{rocker/tidyverse:devel} will provide the
currently-released versions of the R packages in the \texttt{tidyverse}
\citep{tidyverse} installed on the nightly build of R-devel. Developers
building packages on this stack are encouraged to tag their images
accordingly as well. Table 3 indicates which versions of R are currently
available in the stack, going back to \texttt{3.1.0}. While older
versions may be added to the stack at a later date, we note that the
MRAN snapshots began in 2014-09-17 and thus go back only to the R
\texttt{3.1} era. Each tag must be built from a separate Dockerfile,
enabling minor differences in the build instructions to accommodate
changing dependencies. Dockerfiles for past versions (\emph{e.g.}, prior
to \texttt{3.4.2} currently) are intended to remain static over the long
term, while the tag for the current version, \texttt{latest}, and
\texttt{devel} may be tweaked to accommodate new features or
dependencies. Version tags also obey semantics so that omitting the
second or third position of the tag is identical to asking for the most
recent version: i.e. \texttt{rocker/verse:3.3} is the same as
\texttt{rocker/verse:3.3.3}, and \texttt{rocker/verse:3} is (at the time
of writing), \texttt{rocker/verse:3.4.2}. This is accomplished using
post-build hooks in Docker Hub, see examples at
\url{https://github.com/rocker-org/rocker-versioned/} for details.

\textbf{Installation}: In this stack, the desired version of R is always
built directly from source rather than the \texttt{apt} repositories.
Compilers and dependencies are still installed from the stable
\texttt{apt} repositories, and thus lag behind the more recent versions
found in the \texttt{testing} stack. Version tags \texttt{3.3.3} and
older are based on the Debian 8.0 release, code-named \texttt{jessie},
while \texttt{3.4.0} - \texttt{3.4.2}, \texttt{devel}, and
\texttt{latest} are based on Debian 9.0, \texttt{stretch}, (released
2017-06-17, while R was at \texttt{3.4.0}), and thus have access to much
newer versions of common system dependencies and compilers. Dependencies
needed to compile R that are not required at runtime are removed once R
is installed, keeping the base images light-weight for faster download
times. While most system dependencies required by common R packages can
still be installed from the \texttt{apt} repositories, occasionally a
more recent version must be compiled from source (\emph{e.g.}, the Gibbs
Sampling program JAGS \citep{jags}, and the geospatial toolkit GDAL,
must both be compiled from source on \texttt{debian:jessie} images). In
this stack, users should avoid installing R packages using \texttt{apt}
without careful consideration as this will install a second (probably
different) version of R from the Debian repositories, and a dated
version of the R package since any \texttt{r-cran-pkgname} package in
the Debian repositories will depend on \texttt{r-base} in \texttt{apt}
as well.

\textbf{Build schedule}: All images are built automatically from their
corresponding Dockerfiles (found in the GitHub repositories
\texttt{rocker-org/rocker-versioned} and
\texttt{rocker-org/geospatial}). A \texttt{cron} job sends nightly build
triggers to Docker Hub to rebuild the \texttt{latest} and \texttt{devel}
tagged images throughout the stack. To decrease load on the hub, build
triggers for the numeric version tags are sent monthly. Although the
Dockerfiles for older R versions installs an almost-identical software
environment every time, the monthly rebuilding of these images on Docker
Hub ensures they continue to receive Debian security updates from
upstream, and proves the build recipe still executes successfully. Note
that rebuilding images with software from external repositories never
produces a bit-wise identical image, and thus the image identifier hash
will change at each build.

\subsubsection{Images overview}\label{images-overview-1}

In this stack, each image builds on the previous image, rather than all
other images building directly on the base image, as in the
\texttt{testing} stack. Table 2 lists the names and descriptions of the
five images in this stack, along with image size and approximate
download counts from Docker Hub. Sizes reflect (compressed) cumulative
size: a user who has already downloaded the most recent version of
\texttt{r-ver} and then pulls a copy of \texttt{rstudio} image will only
need to download the additional 115 MB in the \texttt{rstudio} layers
and not the full 334 MB listed. This linear design limits flexibility
(no option for \texttt{tidyverse} without \texttt{rstudio}) but
simplifies use and maintenance. While no single environment will be
optimal for everyone, both the packages selected in this stack and the
stack ordering reflect considerable community input and tuning.

The \texttt{rstudio} image includes a lightweight, easy-to-use and
docker-friendly \texttt{init} system, s6 \citep{s6} for running
persistent services, including the
RStudio\textsuperscript{\textregistered} server. This system provides a
convenient way for downstream Dockerfile developers to add additional
persistent services (such as an \texttt{ssh} server) to a single
container, or additional start-up or shutdown scripts that should be run
when a container starts up or shuts down. The \texttt{rstudio} image
uses such a start-up script to configure user settings such as login
password and permissions through environmental variables at run time.

The \texttt{tidyverse} image contains all required and suggested
dependencies of the commonly-used \texttt{tidyverse} and
\texttt{devtools} R packages, including external database libraries
(\emph{e.g.}, MariaDB and PostgreSQL). Users should consult the package
Dockerfiles or \texttt{installed.packages()} list directly for a
complete list of installed packages. The \texttt{verse} library adds
commonly-used dependencies, notably a large but not comprehensive LaTeX
environment and Java development libraries. Previously, the Rocker
project provided the image \texttt{hadleyverse} which was since divided
into \texttt{tidyverse} and \texttt{verse} through community input.

\begin{longtable}[]{@{}lllll@{}}
\caption{Available tags in the \texttt{rocker-versioned}
stack.}\tabularnewline
\toprule
\begin{minipage}[b]{0.09\columnwidth}\raggedright\strut
tag\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\raggedright\strut
apt repos\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\raggedright\strut
MRAN date\strut
\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\raggedright\strut
Build frequency\strut
\end{minipage} & \begin{minipage}[b]{0.32\columnwidth}\raggedright\strut
images with tag\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{0.09\columnwidth}\raggedright\strut
tag\strut
\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\raggedright\strut
apt repos\strut
\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\raggedright\strut
MRAN date\strut
\end{minipage} & \begin{minipage}[b]{0.19\columnwidth}\raggedright\strut
Build frequency\strut
\end{minipage} & \begin{minipage}[b]{0.32\columnwidth}\raggedright\strut
images with tag\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
devel\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{stretch}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
current date\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
nightly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
latest\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{stretch}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
current date\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
nightly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.4.2\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{stretch}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
current date\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.4.1\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{stretch}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2017-09-28\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.4.0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{stretch}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2017-06-30\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.3.3\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2017-04-21\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.3.2\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2017-03-06\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.3.1\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2016-10-31\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}, \texttt{rstudio}, \texttt{tidyverse}, \texttt{verse},
\texttt{geospatial}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.3.0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2016-06-21\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.2.0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2015-06-18\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut
3.1.0\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut
\texttt{jessie}\strut
\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright\strut
2014-09-17\strut
\end{minipage} & \begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
monthly\strut
\end{minipage} & \begin{minipage}[t]{0.32\columnwidth}\raggedright\strut
\texttt{r-ver}\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

Several images in the \texttt{rocker-versioned} stack can be customized
on build when built locally (rather than pulling prebuilt images from
Docker Hub) by using the \texttt{-\/-build-arg} option of
\texttt{docker\ build}. In the \texttt{r-ver} image, users can set
\texttt{R\_VERSION} and \texttt{BUILD\_DATE} (MRAN default snapshot). In
the \texttt{rstudio} image users can set \texttt{RSTUDIO\_VERSION}
(otherwise defaults to the most recent version), and the
\texttt{PANDOC\_TEMPLATES\_VERSION} .

This stack also makes use of Docker metadata labels defined by
\url{http://schema-label.org}, indicating image \texttt{license}
(GPL-2.0), \texttt{vcs-url} (GitHub repository), and \texttt{vendor}
(Rocker Project). These metadata can be altered or extended in
downstream images.

\section{Conclusions}\label{conclusions}

Over the past several years, Docker has seen immense adoption across
industry and academia. The Open Container initiative \citep{oci} now
provides an open standard that has further extended this container
approach to research environments through projects such as Singularity
\citep{singularity}, allowing users to deploy containerized environments
such as Rocker on machines where they do not have root access, such as
clusters or private servers. Containerization promises to solve numerous
challenges such as portability and replicability in research computing,
which often relies on complex and heterogeneous software stacks
\citep{Boettiger2015}. Yet implementing such environments in containers
is not a trivial task, and not all implementations provide the same
usability, portability or reproducibility. Here we have detailed the
approach taken by the Rocker project in creating and maintaining these
environments through an open and community-driven process. This
structure of the Rocker project has evolved over three years of
operation while drawing in an ever-widening base of academic
researchers, university instructors and industry users. We believe this
overview will be instructive not only to users and developers interested
in the Rocker project, but as a model for similar efforts around other
environments or domains.

\bibliography{RJreferences}

\address{%
Carl Boettiger\\
UC Berkeley\\
ESPM Department, University of California,\\ 130 Mulford Hall Berkeley, CA 94720-3114, USA\\
}
\href{mailto:cboettig@berkeley.edu}{\nolinkurl{cboettig@berkeley.edu}}

\address{%
Dirk Eddelbuettel\\
Debian and R Projects; Ketchum Trading\\
Chicago, IL, USA\\
}
\href{mailto:edd@debian.org}{\nolinkurl{edd@debian.org}}

### 2017-11-22 boettiger-eddelbuettel.Rmd
A few edits re major comments of reviewer 3_2

@@
 code, runtime, system tools, system runtime.  Docker containers are
-lightweight as they share the operating system kernel, starting instantly
+lightweight as they share the operating system kernel, start instantly
 using a layered filesystem which minimizes disk footprint and download time,
@@
 once, and can then be "run" to create a container whenever it is needed.
-A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
-and contributions to the Rocker project focuses on the construction, organization
+A "Dockerfile" is a recipe, the source-code, to create a Docker image.  
+Pre-built Docker images are publicly available through Docker Hub, which
+plays a role for central distribution similar to CRAN in our analogy.  Development
+and contributions to the Rocker project focus on the construction, organization
 and maintenance of these Dockerfiles.
 
-# Design principles & use cases
+# Design principles and use cases
 
@@
 use-case is access to R-devel without affecting the local system.
-Here, we detail some of the principle use cases motivating these
+Here, we detail some of the principal use cases motivating these
 containerized versions of R environments, and the design principles that
@@
 Rocker containers are  also easy to run locally on most modern laptops
-by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
+by first issuing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
 operating systems. By sharing volumes with the local host, users can still manipulate
@@
 
-For instance, the following bash commands install Docker on almost any 
-Linux-based server and then launch a Rocker container providing the 
+For instance, on any platform wiht Docker installed, the following Docker
+command will launch a Rocker container providing the 
 RStudio-server environment over a web interface.
@@
 
-The `docker run` option `-p` sets the port on which RStudio will appear, 8787 is
+The `docker run` option `-p` sets the port on which RStudio will appear, for which 8787 is
 the RStudio default (adding your user to the docker group to avoids the
@@
 options such as using gcc or clang compiler sanitizers \citep{edd_sanitizers}.
-These requires R itself be built with specialized settings that may not be
+These require R itself be built with specialized settings that may not be
 not available or familiar to many R users on their native system, but can
@@
 be deployed from a machine without root privileges through the alternate container
-runtime environment, Singularity \citep{singularity}.  Unlike traditional virtual machines,
+runtime environment, Singularity \citep{singularity}.  
+Sandboxing also serves an important function
+in reproducible research by making it easier to test a specified environment
+in isolation from the host machine.
+Unlike traditional virtual machines,
 these containers do not impose a large footprint of reserved resources as
@@
 to install R packages is fixed to a snapshot of CRAN corresponding to
-the last date for which that version of R was the latest release. These
+the last date for which that version of R was current (e.g. `3.4.2` was released
+on 2017-09-28, thus `3.4.1` is pinned to the MRAN snapshot for that date). These
 snapshots are provided by the MRAN archive created by Revolution

### 2017-11-23 boettiger-eddelbuettel.Rmd
WIP

@@
 and run software applications using *containers*.  Containers provide a
-virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
-code, runtime, system tools, system runtime.  Docker containers are
+virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run.  Docker containers are
 lightweight as they share the operating system kernel, start instantly
@@
 an isolated software environment on a computer. R users can think of running
-a container as analogous to loading an R package; it makes a set of software
-functions available.  A Docker "image" is a binary archive of that software,
+a container as analogous to loading an R package; a container is an active
+instance of a static Docker image.
+A Docker "image" is a binary archive of that software,
 analogous to an R binary package: a given version is downloaded only
@@
 Rocker containers are also easy to run locally on most modern laptops
-by first issuing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
-operating systems. By sharing volumes with the local host, users can still manipulate
+using Windows, MacOS, or Linux-based operating systems. 
+By sharing volumes with the local host, users can still manipulate
 files with familiar, native tools while performing computation through a reproducible,
@@
 
-Users can access a root `bash` shell in a running Rocker container using
+Users can access a `bash` shell running as root within a Rocker container using
 
@@
 As with any interactive Docker container, users
-should specify the interactive (`-i`) and terminal (`-t`, here combined
-with interactive as `-ti`) flags, and specify the desired executable
+should specify the terminal (`-t`) and interactive (`-i`) flags, (here combined
+with interactive as `-ti`), and specify the desired executable
 environment (_e.g._, `R`, though other common options may be `Rscript` or
@@
 to install R packages. Granting these users root privileges in the container still leaves
-them sandboxed from the host container.  Further, Docker images can
-be deployed from a machine without root privileges through the alternate container
-runtime environment, Singularity \citep{singularity}.  
+them sandboxed from the host container.   
 Sandboxing also serves an important function
@@
 default repository distributing the images.  Using automated builds rather than uploading
-pre-built image binaries to the Docker Hub avoids the potential for the build
+pre-built image binaries to Docker Hub avoids the potential for the build
 not to match the recipe.  The corresponding Dockerfile is visible both on the Docker Hub
-and in the linked GitHub repository of the Rocker project (https://github.com/rocker-org), 
+and in the linked GitHub repository, 
 which provides a transparent versioned history of all changes made to these recipes,
@@
 changes, bugs, improvements to the Dockerfiles and troubleshoot any issues users may encounter.
-Having these public source files built automatically by a trusted provider (Docker Hub),
-rather than built locally and uploaded only in binary form provides is also useful from a
-security perspective in avoiding malware.
+Having these public source files built automatically by a trusted provider (Docker Hub), rather than built locally and uploaded as binaries, is also useful from a security perspective in avoiding malware.
 

### 2017-11-23 boettiger-eddelbuettel.Rmd
ok, most other minor issues from 3_2 now addressed.

@@
 generally focus on one (current) release.  For example, R users on
-Windows may run RStudio Server or Shiny Server locally just by launching
+Windows may run RStudio\textsuperscript{\textregistered} Server or Shiny\textsuperscript{\textregistered} Server locally just by launching
 a single command (once Docker itself is installed). Another common
@@
 command will launch a Rocker container providing the 
-RStudio-server environment over a web interface.
+rstudio-server environment over a web interface.
 
@@
 
-The `docker run` option `-p` sets the port on which RStudio will appear, for which 8787 is
-the RStudio default (adding your user to the docker group to avoids the
+The `docker run` option `-p` sets the port on which RStudio\textsuperscript{\textregistered} will appear, for which 8787 is
+the default (adding your user to the docker group to avoids the
 need for a `sudo` command to call `docker`: `sudo usermod -g docker $USER`). 
@@
 linear-algebra libraries and so forth.  While this flexibility reflects varying needs,
-many users rely on default settings which may be optimized more for simplicity of installation
-than performance.  The Rocker recipes reflect significant community input on these choices,
-as well as the considerable experience and expertise of the R Debian maintainer of 20 years
-in what configuration options to use.  This helps create a more finely tuned, optimized reference
+many users rely on default settings which are most often are optimized more for 
+simplicity of installation rather than than performance.
+The Rocker recipes reflect significant community input on these choices.
+This helps create a more finely tuned, optimized reference
 implementation of the R environment as well as a platform for comparing and discussing these 
 concerns which are often overlooked elsewhere.  Issues and Pull Requests on the Rocker 
-repositories on GitHub attest to these discussions and improvements.
+repositories on GitHub attest to some of these discussions and improvements.
+In particular, input from the Docker Inc employees through the official approval process
+for the `r-base` image, expertise from the Debian R maintainer and other Debian developers,
+and both direct and indirect feedback from the experience and user-generated documentation
+from many early adopters in the R community has helped shape and strengthen the project
+over the past few years.
 Widespread use of the Rocker image helps promote both testing of these choices and 
@@
 The Dockerfiles in the Rocker stack should themselves provide a simple of example
-of this.  A user begins by selecting an appropriate base image for their needs: 
+of this, (as described in the following section).  
+A user begins by selecting an appropriate base image for their needs: 
 if the RStudio\textsuperscript{\textregistered} interface is desired, a user might start with `FROM rocker/rstudio`;
-an image for testing a particular C code in an R package might use `FROM rocker/r-devel-san`,
+an image for testing an R package with compiled code might use `FROM rocker/r-devel-san`,
 an image for reproducing a data analysis will probably select a stable version tag
@@
 The issue tracker and pull requests are used for community input,
-discussions and contributions to these images. The Rocker project wiki
+discussions and contributions to these images. The Rocker project wiki,
+<https://github.com/rocker-org/rocker/wiki>,
 provides a place to synthesize community-contributed documentation,
@@
 the Rocker team.  Consequently, users should refer to this image in
-Docker commands without an organization namespace, _e.g._, `docker run -t
-r-base` to access the official image.  All other images in the Rocker
+Docker commands without an organization namespace, _e.g._, `docker run -ti r-base` 
+to access the official image.  All other images in the Rocker
 project are not individually reviewed and built by Docker Inc and must
@@
 Docker Hub has been granted to the Rocker project by explicit permission
-from RStudio. Users should review RStudio's trademark use policy
+from RStudio. Users should review RStudio\textsuperscript{\textregistered}'s trademark use policy
 (<http://www.rstudio.com/about/trademark/>) and address inquiries about
@@
 project provided the image `hadleyverse` which was since divided into
-`tidyverse` and `verse` through community input.
+`tidyverse` and `verse` based on community input.
 

### 2017-11-23 boettiger-eddelbuettel.Rmd
Flush out a bit more about singularity
include example of importing and running rocker container on singularity, as reviewer 1 suggests.

@@
 
-Deploying Docker containers on centrally administrated multi-user machines, such
-as department or university clusters, has previously been more difficult, since
-system adminstrators do not want to allow the elevated user permissions the Docker
-runtime environment requires.  To work around this problem, Lawrence Berkeley National
-Labs (LBNL) has made 'Singularity' \citep{singularity}: a container runtime environment that users can
-both install and use to run most Docker containers without requiring root privileges,
-making it easier to deploy Rocker containers in this context as well.
-
 This portability is also valuable in an instructional context. Requiring students
@@
 UC Berkeley \citep{data8}.
+
+## HPC application
+
+The portability of Rocker images can be particularly valuable in High Performance Computing
+contexts Setting up a specific R environment on High Performance Computing platforms and other 
+centrally administrated multi-user machines or clusters has traditionally been challenging
+due to restrictions on root access that may be needed to install certain libraries. 
+Versions of R and packages installed by the system administrator may also lag behind
+the most recent releases. Deploying Docker containers on HPC systems has previously
+been more very problematic since most
+system adminstrators do not want to allow the elevated user permissions the Docker
+runtime environment requires.  To work around this problem, Lawrence Berkeley National
+Labs (LBNL) has made 'Singularity' \citep{singularity}: a container runtime environment that users can
+both install and use to run most Docker containers without requiring root privileges. 
+Singularity has seen rapid adoption in the HPC community (<http://singularity.lbl.gov/install-request>).
+Rocker containers can be run through Singularity with a single command much like
+the native Docker commands, _e.g._
+
+```
+singularity exec docker://rocker/tidyverse:latest R
+```
+
+More details can be found in the Singularity documentation.
 

### 2017-11-23 boettiger-eddelbuettel.tex
'pinning' rewrite
(the diff shows changes in .tex too which may be editor-whitespace
stuff; I only change the 'debian testing' section as the diff to Rmd shows)

@@
 virtual environments) requiring all operating-system components an
-application needs to run: code, runtime, system tools, system runtime.
-Docker containers are lightweight as they share the operating system
-kernel, starting instantly using a layered filesystem which minimizes
-disk footprint and download time, are built on open standards that run
-on all major platforms (Linux, Mac, Windows), and provide an added layer
-of security by running an application in an isolated environment
-\citep{what-docker}. Familiarity with a few key terms is helpful in
-understanding this paper. The term ``container'' refers to an isolated
-software environment on a computer. R users can think of running a
-container as analogous to loading an R package; it makes a set of
-software functions available. A Docker ``image'' is a binary archive of
-that software, analogous to an R binary package: a given version is
-downloaded only once, and can then be ``run'' to create a container
-whenever it is needed. A ``Dockerfile'' is a recipe, the source-code, to
-create a Docker image. Development and contributions to the Rocker
-project focuses on the construction, organization and maintenance of
-these Dockerfiles.
+application needs to run. Docker containers are lightweight as they
+share the operating system kernel, start instantly using a layered
+filesystem which minimizes disk footprint and download time, are built
+on open standards that run on all major platforms (Linux, Mac, Windows),
+and provide an added layer of security by running an application in an
+isolated environment \citep{what-docker}. Familiarity with a few key
+terms is helpful in understanding this paper. The term ``container''
+refers to an isolated software environment on a computer. R users can
+think of running a container as analogous to loading an R package; a
+container is an active instance of a static Docker image. A Docker
+``image'' is a binary archive of that software, analogous to an R binary
+package: a given version is downloaded only once, and can then be
+``run'' to create a container whenever it is needed. A ``Dockerfile'' is
+a recipe, the source-code, to create a Docker image.\\
+Pre-built Docker images are publicly available through Docker Hub, which
+plays a role for central distribution similar to CRAN in our analogy.
+Development and contributions to the Rocker project focus on the
+construction, organization and maintenance of these Dockerfiles.
 
-\section{Design principles \& use
-cases}\label{design-principles-use-cases}
+\section{Design principles and use
+cases}\label{design-principles-and-use-cases}
 
@@
 generally focus on one (current) release. For example, R users on
-Windows may run RStudio Server or Shiny Server locally just by launching
+Windows may run RStudio\textsuperscript{\textregistered} Server or
+Shiny\textsuperscript{\textregistered} Server locally just by launching
 a single command (once Docker itself is installed). Another common
 use-case is access to R-devel without affecting the local system. Here,
-we detail some of the principle use cases motivating these containerized
+we detail some of the principal use cases motivating these containerized
 versions of R environments, and the design principles that help make
@@
 server hardware. Rocker containers are also easy to run locally on most
-modern laptops by first installing the appropriate Docker command (or
-distribution) for Windows, MacOS, and Linux-based operating systems. By
-sharing volumes with the local host, users can still manipulate files
+modern laptops using Windows, MacOS, or Linux-based operating systems.
+By sharing volumes with the local host, users can still manipulate files
 with familiar, native tools while performing computation through a
@@
 
-For instance, the following bash commands install Docker on almost any
-Linux-based server and then launch a Rocker container providing the
-RStudio-server environment over a web interface.
+For instance, on any platform wiht Docker installed, the following
+Docker command will launch a Rocker container providing the
+rstudio-server environment over a web interface.
 
@@
 The \texttt{docker\ run} option \texttt{-p} sets the port on which
-RStudio will appear, 8787 is the RStudio default (adding your user to
-the docker group to avoids the need for a \texttt{sudo} command to call
-\texttt{docker}: \texttt{sudo\ usermod\ -g\ docker\ \$USER}). Many
-academic and commercial cloud providers make it possible to execute such
-code snippets when a container is launched, without ever needing to
+RStudio\textsuperscript{\textregistered} will appear, for which 8787 is
+the default (adding your user to the docker group to avoids the need for
+a \texttt{sudo} command to call \texttt{docker}:
+\texttt{sudo\ usermod\ -g\ docker\ \$USER}). Many academic and
+commercial cloud providers make it possible to execute such code
+snippets when a container is launched, without ever needing to
 \texttt{ssh} into the machine. The user may log into the server merely
@@
 expertise.
-
-Deploying Docker containers on centrally administrated multi-user
-machines, such as department or university clusters, has previously been
-more difficult, since system adminstrators do not want to allow the
-elevated user permissions the Docker runtime environment requires. To
-work around this problem, Lawrence Berkeley National Labs (LBNL) has
-made `Singularity' \citep{singularity}: a container runtime environment
-that users can both install and use to run most Docker containers
-without requiring root privileges, making it easier to deploy Rocker
-containers in this context as well.
 
@@
 Berkeley \citep{data8}.
+
+\subsection{HPC application}\label{hpc-application}
+
+The portability of Rocker images can be particularly valuable in High
+Performance Computing contexts Setting up a specific R environment on
+High Performance Computing platforms and other centrally administrated
+multi-user machines or clusters has traditionally been challenging due
+to restrictions on root access that may be needed to install certain
+libraries. Versions of R and packages installed by the system
+administrator may also lag behind the most recent releases. Deploying
+Docker containers on HPC systems has previously been more very
+problematic since most system adminstrators do not want to allow the
+elevated user permissions the Docker runtime environment requires. To
+work around this problem, Lawrence Berkeley National Labs (LBNL) has
+made `Singularity' \citep{singularity}: a container runtime environment
+that users can both install and use to run most Docker containers
+without requiring root privileges. Singularity has seen rapid adoption
+in the HPC community (\url{http://singularity.lbl.gov/install-request}).
+Rocker containers can be run through Singularity with a single command
+much like the native Docker commands, \emph{e.g.}
+
+\begin{verbatim}
+singularity exec docker://rocker/tidyverse:latest R
+\end{verbatim}
+
+More details can be found in the Singularity documentation.
 
@@
 
-Users can access a root \texttt{bash} shell in a running Rocker
+Users can access a \texttt{bash} shell running as root within a Rocker
 container using
@@
 As with any interactive Docker container, users should specify the
-interactive (\texttt{-i}) and terminal (\texttt{-t}, here combined with
-interactive as \texttt{-ti}) flags, and specify the desired executable
-environment (\emph{e.g.}, \texttt{R}, though other common options may be
-\texttt{Rscript} or \texttt{bash}):
+terminal (\texttt{-t}) and interactive (\texttt{-i}) flags, (here
+combined with interactive as \texttt{-ti}), and specify the desired
+executable environment (\emph{e.g.}, \texttt{R}, though other common
+options may be \texttt{Rscript} or \texttt{bash}):
 
@@
 for particular options such as using gcc or clang compiler sanitizers
-\citep{edd_sanitizers}. These requires R itself be built with
-specialized settings that may not be not available or familiar to many R
-users on their native system, but can be easily deployed by pulling the
-Rocker images \texttt{rocker/r-devel-san} or
+\citep{edd_sanitizers}. These require R itself be built with specialized
+settings that may not be not available or familiar to many R users on
+their native system, but can be easily deployed by pulling the Rocker
+images \texttt{rocker/r-devel-san} or
 \texttt{rocker/r-devel-ubsan-clang}.
@@
 these users root privileges in the container still leaves them sandboxed
-from the host container. Further, Docker images can be deployed from a
-machine without root privileges through the alternate container runtime
-environment, Singularity \citep{singularity}. Unlike traditional virtual
-machines, these containers do not impose a large footprint of reserved
-resources as a typical host can easily support 100s of containers
-\citep{what-docker}.
+from the host container.\\
+Sandboxing also serves an important function in reproducible research by
+making it easier to test a specified environment in isolation from the
+host machine. Unlike traditional virtual machines, these containers do
+not impose a large footprint of reserved resources as a typical host can
+easily support 100s of containers \citep{what-docker}.
 
@@
 default repository distributing the images. Using automated builds
-rather than uploading pre-built image binaries to the Docker Hub avoids
-the potential for the build not to match the recipe. The corresponding
+rather than uploading pre-built image binaries to Docker Hub avoids the
+potential for the build not to match the recipe. The corresponding
 Dockerfile is visible both on the Docker Hub and in the linked GitHub
-repository of the Rocker project (\url{https://github.com/rocker-org}),
-which provides a transparent versioned history of all changes made to
-these recipes, as well as documentation, a community wiki, and issue
-trackers for discussing proposed changes, bugs, improvements to the
-Dockerfiles and troubleshoot any issues users may encounter. Having
-these public source files built automatically by a trusted provider
-(Docker Hub), rather than built locally and uploaded only in binary form
-provides is also useful from a security perspective in avoiding malware.
+repository, which provides a transparent versioned history of all
+changes made to these recipes, as well as documentation, a community
+wiki, and issue trackers for discussing proposed changes, bugs,
+improvements to the Dockerfiles and troubleshoot any issues users may
+encounter. Having these public source files built automatically by a
+trusted provider (Docker Hub), rather than built locally and uploaded as
+binaries, is also useful from a security perspective in avoiding
+malware.
 
@@
 linear-algebra libraries and so forth. While this flexibility reflects
-varying needs, many users rely on default settings which may be
-optimized more for simplicity of installation than performance. The
-Rocker recipes reflect significant community input on these choices, as
-well as the considerable experience and expertise of the R Debian
-maintainer of 20 years in what configuration options to use. This helps
-create a more finely tuned, optimized reference implementation of the R
-environment as well as a platform for comparing and discussing these
-concerns which are often overlooked elsewhere. Issues and Pull Requests
-on the Rocker repositories on GitHub attest to these discussions and
-improvements. Widespread use of the Rocker image helps promote both
-testing of these choices and contributions further tweaking the
-configuration from many members of the R community.
+varying needs, many users rely on default settings which are most often
+are optimized more for simplicity of installation rather than than
+performance. The Rocker recipes reflect significant community input on
+these choices. This helps create a more finely tuned, optimized
+reference implementation of the R environment as well as a platform for
+comparing and discussing these concerns which are often overlooked
+elsewhere. Issues and Pull Requests on the Rocker repositories on GitHub
+attest to some of these discussions and improvements. In particular,
+input from the Docker Inc employees through the official approval
+process for the \texttt{r-base} image, expertise from the Debian R
+maintainer and other Debian developers, and both direct and indirect
+feedback from the experience and user-generated documentation from many
+early adopters in the R community has helped shape and strengthen the
+project over the past few years. Widespread use of the Rocker image
+helps promote both testing of these choices and contributions further
+tweaking the configuration from many members of the R community.
 
@@
 appropriate Rocker image. The Dockerfiles in the Rocker stack should
-themselves provide a simple of example of this. A user begins by
-selecting an appropriate base image for their needs: if the
-RStudio\textsuperscript{\textregistered} interface is desired, a user
-might start with \texttt{FROM\ rocker/rstudio}; an image for testing a
-particular C code in an R package might use
+themselves provide a simple of example of this, (as described in the
+following section).\\
+A user begins by selecting an appropriate base image for their needs: if
+the RStudio\textsuperscript{\textregistered} interface is desired, a
+user might start with \texttt{FROM\ rocker/rstudio}; an image for
+testing an R package with compiled code might use
 \texttt{FROM\ rocker/r-devel-san}, an image for reproducing a data
@@
 are used for community input, discussions and contributions to these
-images. The Rocker project wiki provides a place to synthesize
-community-contributed documentation, use-cases, and other knowledge
-about using the Rocker images.
+images. The Rocker project wiki,
+\url{https://github.com/rocker-org/rocker/wiki}, provides a place to
+synthesize community-contributed documentation, use-cases, and other
+knowledge about using the Rocker images.
 
@@
 compilers are available as binaries than would be found in a Debian
-stable release. In order to provide access to the most recent available
-binaries, this stack uses apt-pinning \citep{apt_pinning} to allow the
-\texttt{apt} package manager to also install binaries from
-\texttt{debian:unstable}, which represents the most recent, bleeding
-edge of packages built for Debian when necessary. For instance, the
-\texttt{r-base} image provided by Rocker installs the most recent
-version of R as a binary from Debian \texttt{unstable}. Similarly,
-recent versions of many popular R packages can also be installed through
-the package manager, \emph{e.g.}, \texttt{apt-get\ install\ r-cran-xml}.
-This can be particularly helpful for packages with external system
-dependencies (such as \texttt{libxml2-dev} in this example) which cannot
-be installed from the R console as they are system dependencies rather
-than R packages installed from within R. We should note, however, that
-only about 500 of the over 11,000 CRAN packages are available as Debian
-packages.
+stable release. In order to provide \emph{optional} access to the most
+recent available binaries, this stack uses apt-pinning
+\citep{apt_pinning} to allow the \texttt{apt} package manager to
+\emph{selectively} install binaries from \texttt{debian:unstable}, which
+represents the most recent set of packages built for Debian. Similarly,
+recent versions of many popular R packages can also be installed
+pre-built through the package manager, \emph{e.g.},
+\texttt{apt-get\ install\ r-cran-xml}. This can be particularly helpful
+for packages with external system dependencies (such as
+\texttt{libxml2-dev} in this example) which cannot be installed from the
+R console as they are system dependencies rather than R packages
+installed from within R. We should note, however, that only about 500 of
+the over 11,000 CRAN packages are available as Debian packages.
 
-As the names \texttt{testing} and \texttt{unstable} imply, this approach
-is not without challenges. The particular version of any given package
-can change as packages move from \texttt{unstable} into
-\texttt{testing}. New versions are sent to \texttt{unstable} during the
-normal course of Debian development. This can occasionally break an
+As the names \texttt{testing} and \texttt{unstable} imply, particular
+versions of package can change as packages move from \texttt{unstable}
+into \texttt{testing}. New versions are sent to \texttt{unstable} during
+the normal course of Debian development. This can occasionally break an
 previously-working installation command in a Dockerfile until the
@@
 
-Relative to \texttt{stable}, the \texttt{testing} stack thus offers the
-following advantages:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These Dockerfiles are easy to develop and extend because almost all
-  software can be installed through the package manager.
-\item
-  These Dockerfiles always install the most recent available software.
-\item
-  These Dockerfiles can almost always build relatively quickly.
-\end{enumerate}
-
-and these down-sides:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These Dockerfiles require occasional minor maintenance to ensure
-  successful builds. (\emph{e.g.}, changing an installation directive
-  from \texttt{unstable} to \texttt{testing} or vice versa).
-\item
-  The resulting images are inherently dynamic: rebuilding the same
-  Dockerfile months or years apart will generate images with
-  significantly different versions of software installed.
-\item
-  The use of apt pinning may be unfamiliar to some users, where the user
-  must ultimately be responsible to ensure compatibility
-  \citep{apt_pinning}.
-\end{enumerate}
+Relative to \texttt{stable}, the \texttt{testing} stack thus offers some
+advantages as almost all software can be installed through the package
+manager. Installation of binary packages from \texttt{testing} generally
+provides the most recent available software, and installs it quickly as
+a binary. On the other hand, these Dockerfiles may require occasional
+maintenance when packages migrate and/or versions change. The resulting
+images are also inherently dynamic: rebuilding the same Dockerfile
+months or years apart will generate images with significantly different
+versions of software installed as the pool of underlying packages
+changes through time.
 
@@
 image in Docker commands without an organization namespace, \emph{e.g.},
-\texttt{docker\ run\ -t\ r-base} to access the official image. All other
-images in the Rocker project are not individually reviewed and built by
-Docker Inc and must be referenced using the \texttt{rocker} namespace,
-\emph{e.g.}, \texttt{docker\ run\ -ti\ rocker/r-devel}.
+\texttt{docker\ run\ -ti\ r-base} to access the official image. All
+other images in the Rocker project are not individually reviewed and
+built by Docker Inc and must be referenced using the \texttt{rocker}
+namespace, \emph{e.g.}, \texttt{docker\ run\ -ti\ rocker/r-devel}.
 
@@
 granted to the Rocker project by explicit permission from RStudio. Users
-should review RStudio's trademark use policy
-(\url{http://www.rstudio.com/about/trademark/}) and address inquiries
-about further distribution or other questions to
+should review RStudio\textsuperscript{\textregistered}'s trademark use
+policy (\url{http://www.rstudio.com/about/trademark/}) and address
+inquiries about further distribution or other questions to
 \href{mailto:permissions@rstudio.com}{\nolinkurl{permissions@rstudio.com}}.
@@
 mirror from which to install R packages is fixed to a snapshot of CRAN
-corresponding to the last date for which that version of R was the
-latest release. These snapshots are provided by the MRAN archive created
-by Revolution Analytics (now part of Microsoft). It archives daily
-snapshots of all of CRAN from which a user can install packages with the
-usual \texttt{install.packages()} function \citep{MRAN}. Users can
-always override this default by passing any current CRAN repository
-explicitly. Unlike CRAN, Bioconductor only updates its repositories
-through bi-annual releases aligned to R's spring release schedule. Thus,
-Bioconductor packages can be installed in the usual way using
-\texttt{bioclite}, which automatically selects the Bioconductor release
-corresponding to the version of R in use.
+corresponding to the last date for which that version of R was current
+(e.g. \texttt{3.4.2} was released on 2017-09-28, thus \texttt{3.4.1} is
+pinned to the MRAN snapshot for that date). These snapshots are provided
+by the MRAN archive created by Revolution Analytics (now part of
+Microsoft). It archives daily snapshots of all of CRAN from which a user
+can install packages with the usual \texttt{install.packages()} function
+\citep{MRAN}. Users can always override this default by passing any
+current CRAN repository explicitly. Unlike CRAN, Bioconductor only
+updates its repositories through bi-annual releases aligned to R's
+spring release schedule. Thus, Bioconductor packages can be installed in
+the usual way using \texttt{bioclite}, which automatically selects the
+Bioconductor release corresponding to the version of R in use.
 
@@
 project provided the image \texttt{hadleyverse} which was since divided
-into \texttt{tidyverse} and \texttt{verse} through community input.
+into \texttt{tidyverse} and \texttt{verse} based on community input.
 

### 2017-11-23 boettiger-eddelbuettel.Rmd
inline comments from review_3_1.pdf
all open ones added, apart from the comments at the bottom of page 1
for section 'Portability: From laptop to cloud' which I cannot make
sense of

@@
 command will launch a Rocker container providing the 
-rstudio-server environment over a web interface.
+RStudio\textsuperscript{\textregistered} server environment over a web interface.
 
@@
 The `docker run` option `-p` sets the port on which RStudio\textsuperscript{\textregistered} will appear, for which 8787 is
-the default (adding your user to the docker group to avoids the
+the default (adding your user to the docker group avoids the
 need for a `sudo` command to call `docker`: `sudo usermod -g docker $USER`). 
@@
 
-which can be useful for administrative tasks such as installing system dependencies.
+\noindent which can be useful for administrative tasks such as installing system dependencies.
 All Rocker images can also be run as an interactive R, RScript or bash shell without running
@@
 with interactive as `-ti`), and specify the desired executable
-environment (_e.g._, `R`, though other common options may be `Rscript` or
+environment (_e.g._, `R`, though other common options could be `Rscript` or
 `bash`):
@@
 installation goes poorly.  However, upgrading packages and/or the R
-environment is often necessary to run analysis of a colleague, or access
+environment is often necessary to run analyses from a colleague, or access
 more recent methods.  Rocker offers an easy solution. For 
@@
 Unlike traditional virtual machines,
-these containers do not impose a large footprint of reserved resources as
+containers do not impose a large footprint of reserved resources as
 a typical host can easily support 100s of containers \citep{what-docker}.
@@
 Widespread use of the Rocker image helps promote both testing of these choices and 
-contributions further tweaking the configuration from many members of the R community.
+contributions, further tweaking the configuration from many members of the R community.
 
@@
 a Dockerfile building `FROM rocker/verse:3.4.1` will only install R package versions that were
-available on CRAN on 2017-06-30, that is, the day R 3.4.1 was released.  This default
+available on CRAN on 2017-06-30, _i.e._, the day R 3.4.1 was released.  This default
 can of course be overwritten in the standard R manner, _e.g._, by specifying a different CRAN mirror explicitly in any 
-command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
+command to install packages, _e.g._, `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
 in an `.Rprofile`.  Note that the MRAN date associated with the current release (_e.g._, `3.4.2` at the time of
@@
 any user by writing their own Dockerfiles based on an appropriate Rocker image.
-The Dockerfiles in the Rocker stack should themselves provide a simple of example
+The Dockerfiles in the Rocker stack should themselves provide a simple example
 of this, (as described in the following section).  
@@
 an image for testing an R package with compiled code might use `FROM rocker/r-devel-san`,
-an image for reproducing a data analysis will probably select a stable version tag
+and an image for reproducing a data analysis will probably select a stable version tag
 in addition to an appropriate base library, _e.g._,: `FROM rocker/tidyverse:3.4.1`
@@
 The issue tracker and pull requests are used for community input,
-discussions and contributions to these images. The Rocker project wiki,
+discussions, and contributions to these images. The Rocker project wiki,
 <https://github.com/rocker-org/rocker/wiki>,
@@
 R environments by writing their own Dockerfiles, while also providing a
-'batteries included' approach of images that can be used out of the box.
+'batteries included' approach to images that can be used out of the box.
 The challenges of balancing diverse needs driven by very different use
 cases against the overarching goals
-of creating images that are still sufficiently light-weight, easy to use
+of creating images that are still sufficiently light-weight, easy to use,
 and easy to maintain is a difficult art.  The implementation in both
@@
 The Debian platform provides a small base image, the well-known `apt`
-package management system and rich ecosystem of software libraries,
+package management system, and a rich ecosystem of software libraries,
 making it the base image of choice for Docker images, including many
@@
 Docker organization itself.  This official image is reviewed and then
-built by employees of Docker Inc based on a Dockerfile maintained by
+built by employees of Docker Inc. based on a Dockerfile maintained by
 the Rocker team.  Consequently, users should refer to this image in
@@
 to access the official image.  All other images in the Rocker
-project are not individually reviewed and built by Docker Inc and must
+project are not individually reviewed and built by Docker Inc. and must
 be referenced using the `rocker` namespace,
@@
 Several of the images in this stack are oriented towards the R
-development community: `r-devel`, `drd`, `r-devel-san` &
-`r-devel-ubsan-clang` all add a copy of the development version of R
-side-by-side the current release of R provided by `r-base`.  On these
+development community: `r-devel`, `drd`, `r-devel-san`, and 
+`r-devel-ubsan-clang` which all add a copy of the development version of R
+side-by-side to the current release of R provided by `r-base`.  On these
 images, the development version is aliased to `RD` to distinguish from
@@
 providing development R built with support for C/C++ address and
-undefined-behavior sanitizers, which are somewhat more difficult to
+undefined-behavior sanitizers, which are somewhat difficult to
 configure \citep{edd_sanitizers}.
@@
 few weeks).  The rest of the stack uses build triggers that rebuild the
-images either whenever `r-base` is updated, or the Dockerfile sources are
+images whenever `r-base` is updated or the Dockerfile sources are
 updated on the corresponding GitHub repository.  The only exception in
@@
 [r-base](https://hub.docker.com/r/_/r-base)        | official image with current version of R  | 254 MB | 632,000   
-[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side r-base (using alias `RD`)    | 1 GB | 4,000
+[r-devel](https://hub.docker.com/r/rocker/r-devel) |  R-devel added side-by-side to r-base (using alias `RD`)    | 1 GB | 4,000
 [drd](https://hub.docker.com/r/rocker/drd) |  lightweight r-devel, built weekly  | 571 MB | 4,000
 [r-devel-san](https://hub.docker.com/r/rocker/r-devel-san)  |  as r-devel, but built with compiler sanitizers | 1.1 GB  | 1,000  
-[r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  Sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
+[r-devel-ubsan-clang](https://hub.docker.com/r/rocker/r-devel-ubsan-clang)            |  sanitizers, clang c compiler (instead of gcc) | 1.1 GB | 525
 [rstudio:testing](https://hub.docker.com/r/rocker/r-devel-san)        |  rstudio on debian:testing | 1.1 GB | 1,000 
@@
 ---------------- | ----------------------------------------- | ------ | ------- 
-[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  Version-stable base R & src build tools  | 219 MB | 6,000
-[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  Adds rstudio                             | 334 MB | 314,000
-[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  Adds tidyverse & devtools                | 656 MB | 83,000 [^1]
-[verse](https://hub.docker.com/r/rocker/verse)          |  Adds java, tex & publishing-related packages   | 947 MB | 9,000
-[geospatial](https://hub.docker.com/r/rocker/geospatial)|  Adds geospatial libraries   |  1.3 GB | 4,000
+[r-ver](https://hub.docker.com/r/rocker/r-ver)          |  version-stable base R & src build tools  | 219 MB | 6,000
+[rstudio](https://hub.docker.com/r/rocker/rstudio)      |  adds rstudio                             | 334 MB | 314,000
+[tidyverse](https://hub.docker.com/r/rocker/tidyverse)  |  adds tidyverse & devtools                | 656 MB | 83,000 [^1]
+[verse](https://hub.docker.com/r/rocker/verse)          |  adds java, tex & publishing-related packages   | 947 MB | 9,000
+[geospatial](https://hub.docker.com/r/rocker/geospatial)|  adds geospatial libraries   |  1.3 GB | 4,000
 
@@
 will always point to an image using the current R release.  Thus, users
-wanting to create downstream Dockerfiles which are based on the current
-release at time (but will continue to reconstruct the same environment in
-the future after newer R versions are released) should explicitly include
+who want to create downstream Dockerfiles, which are based on the current
+release at the time (but will continue to reconstruct the same environment in
+the future after newer R versions are released), should explicitly include
 the corresponding version tag, _e.g._, `rocker/r-ver:3.4.2` at the time
@@
 `rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.
-This is accomplished using post-build hooks in Docker Hub, see examples at
+This is accomplished using post-build hooks in Docker Hub---see examples at
 <https://github.com/rocker-org/rocker-versioned/> for details.
@@
 build triggers for the numeric version tags are sent monthly. Although
-the Dockerfiles for older R versions installs an almost-identical
+the Dockerfiles for older R versions install an almost-identical
 software environment every time, the monthly rebuilding of these images
@@
 environment and Java development libraries.  Previously, the Rocker
-project provided the image `hadleyverse` which was since divided into
+project provided the image `hadleyverse` which has since been divided into
 `tidyverse` and `verse` based on community input.

### 2017-11-24 boettiger-eddelbuettel.Rmd

--- 2017-11-23 boettiger-eddelbuettel.Rmd
@@
 as binaries than would be found in a Debian stable release.  In order
-to provide access to the most recent available binaries, this stack uses
-apt-pinning \citep{apt_pinning} to allow the `apt` package manager to also
-install binaries from `debian:unstable`, which represents the most recent,
-bleeding edge of packages built for Debian when necessary.  For instance,
-the `r-base` image provided by Rocker installs the most recent version of
-R as a binary from Debian `unstable`.  Similarly, recent versions of many
-popular R packages can also be installed through the package manager,
+to provide _optional_ access to the most recent available binaries, this stack uses
+apt-pinning \citep{apt_pinning} to allow the `apt` package manager to _selectively_
+install binaries from `debian:unstable`, which represents the most recent
+set of packages built for Debian. Similarly, recent versions of many
+popular R packages can also be installed pre-built through the package manager,
 _e.g._, `apt-get install r-cran-xml`.  This can be particularly helpful
@@
 
-As the names `testing` and `unstable` imply, this approach is not
-without challenges.  The particular version of any given package can
+As the names `testing` and `unstable` imply, particular versions of package can
 change as packages move from `unstable` into `testing`. New versions
@@
 
-Relative to `stable`, the `testing` stack thus offers the following advantages:
-
-1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager.
-2. These Dockerfiles always install the most recent available software.
-3. These Dockerfiles can almost always build relatively quickly.
-
-and these down-sides:
-
-1. These Dockerfiles require occasional minor maintenance to ensure successful builds. (_e.g._, changing an installation directive from `unstable` to `testing` or vice versa).
-2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
-3. The use of apt pinning may be unfamiliar to some users, where the user must ultimately be responsible to ensure compatibility \citep{apt_pinning}.
+Relative to `stable`, the `testing` stack thus offers some advantages
+as almost all software can be installed through the package
+manager. Installation of binary packages from `testing` generally provides
+the most recent available software, and installs it quickly as a
+binary. On the other hand, these Dockerfiles may require occasional 
+maintenance when packages migrate and/or versions change. The resulting images 
+are also inherently dynamic: rebuilding the same Dockerfile months or years 
+apart will generate images with significantly different versions of software 
+installed as the pool of underlying packages changes through time.
 

### 2017-11-24 boettiger-eddelbuettel.tex
render and commit

@@
 Docker command will launch a Rocker container providing the
-rstudio-server environment over a web interface.
+RStudio\textsuperscript{\textregistered} server environment over a web
+interface.
 
@@
 RStudio\textsuperscript{\textregistered} will appear, for which 8787 is
-the default (adding your user to the docker group to avoids the need for
-a \texttt{sudo} command to call \texttt{docker}:
+the default (adding your user to the docker group avoids the need for a
+\texttt{sudo} command to call \texttt{docker}:
 \texttt{sudo\ usermod\ -g\ docker\ \$USER}). Many academic and
@@
 
-which can be useful for administrative tasks such as installing system
-dependencies. All Rocker images can also be run as an interactive R,
-RScript or bash shell without running RStudio, which can be useful for
-batch jobs or for anyone who prefers that environment.
+\noindent which can be useful for administrative tasks such as
+installing system dependencies. All Rocker images can also be run as an
+interactive R, RScript or bash shell without running RStudio, which can
+be useful for batch jobs or for anyone who prefers that environment.
 
@@
 executable environment (\emph{e.g.}, \texttt{R}, though other common
-options may be \texttt{Rscript} or \texttt{bash}):
+options could be \texttt{Rscript} or \texttt{bash}):
 
@@
 environment if the installation goes poorly. However, upgrading packages
-and/or the R environment is often necessary to run analysis of a
+and/or the R environment is often necessary to run analyses from a
 colleague, or access more recent methods. Rocker offers an easy
@@
 making it easier to test a specified environment in isolation from the
-host machine. Unlike traditional virtual machines, these containers do
-not impose a large footprint of reserved resources as a typical host can
+host machine. Unlike traditional virtual machines, containers do not
+impose a large footprint of reserved resources as a typical host can
 easily support 100s of containers \citep{what-docker}.
@@
 project over the past few years. Widespread use of the Rocker image
-helps promote both testing of these choices and contributions further
+helps promote both testing of these choices and contributions, further
 tweaking the configuration from many members of the R community.
@@
 \texttt{FROM\ rocker/verse:3.4.1} will only install R package versions
-that were available on CRAN on 2017-06-30, that is, the day R 3.4.1 was
-released. This default can of course be overwritten in the standard R
-manner, \emph{e.g.}, by specifying a different CRAN mirror explicitly in
-any command to \texttt{install.packages()}, or by adjusting the default
-CRAN mirror in
+that were available on CRAN on 2017-06-30, \emph{i.e.}, the day R 3.4.1
+was released. This default can of course be overwritten in the standard
+R manner, \emph{e.g.}, by specifying a different CRAN mirror explicitly
+in any command to install packages, \emph{e.g.},
+\texttt{install.packages()}, or by adjusting the default CRAN mirror in
 \texttt{options(repo=\textless{}CRAN-MIRROR\textgreater{})} in an
@@
 appropriate Rocker image. The Dockerfiles in the Rocker stack should
-themselves provide a simple of example of this, (as described in the
+themselves provide a simple example of this, (as described in the
 following section).\\
@@
 testing an R package with compiled code might use
-\texttt{FROM\ rocker/r-devel-san}, an image for reproducing a data
+\texttt{FROM\ rocker/r-devel-san}, and an image for reproducing a data
 analysis will probably select a stable version tag in addition to an
@@
 \url{https://github.com/rocker-org}. The issue tracker and pull requests
-are used for community input, discussions and contributions to these
+are used for community input, discussions, and contributions to these
 images. The Rocker project wiki,
@@
 R environments by writing their own Dockerfiles, while also providing a
-`batteries included' approach of images that can be used out of the box.
+`batteries included' approach to images that can be used out of the box.
 The challenges of balancing diverse needs driven by very different use
 cases against the overarching goals of creating images that are still
-sufficiently light-weight, easy to use and easy to maintain is a
+sufficiently light-weight, easy to use, and easy to maintain is a
 difficult art. The implementation in both individual Rocker images and
@@
 platform provides a small base image, the well-known \texttt{apt}
-package management system and rich ecosystem of software libraries,
+package management system, and a rich ecosystem of software libraries,
 making it the base image of choice for Docker images, including many of
@@
 language by the Docker organization itself. This official image is
-reviewed and then built by employees of Docker Inc based on a Dockerfile
-maintained by the Rocker team. Consequently, users should refer to this
-image in Docker commands without an organization namespace, \emph{e.g.},
-\texttt{docker\ run\ -ti\ r-base} to access the official image. All
-other images in the Rocker project are not individually reviewed and
-built by Docker Inc and must be referenced using the \texttt{rocker}
-namespace, \emph{e.g.}, \texttt{docker\ run\ -ti\ rocker/r-devel}.
+reviewed and then built by employees of Docker Inc. based on a
+Dockerfile maintained by the Rocker team. Consequently, users should
+refer to this image in Docker commands without an organization
+namespace, \emph{e.g.}, \texttt{docker\ run\ -ti\ r-base} to access the
+official image. All other images in the Rocker project are not
+individually reviewed and built by Docker Inc. and must be referenced
+using the \texttt{rocker} namespace, \emph{e.g.},
+\texttt{docker\ run\ -ti\ rocker/r-devel}.
 
@@
 development community: \texttt{r-devel}, \texttt{drd},
-\texttt{r-devel-san} \& \texttt{r-devel-ubsan-clang} all add a copy of
-the development version of R side-by-side the current release of R
-provided by \texttt{r-base}. On these images, the development version is
-aliased to \texttt{RD} to distinguish from the current release,
-\texttt{R}. As the names suggest, each provide slightly different
-configurations. Of particular interest are the images providing
-development R built with support for C/C++ address and
-undefined-behavior sanitizers, which are somewhat more difficult to
-configure \citep{edd_sanitizers}.
+\texttt{r-devel-san}, and \texttt{r-devel-ubsan-clang} which all add a
+copy of the development version of R side-by-side to the current release
+of R provided by \texttt{r-base}. On these images, the development
+version is aliased to \texttt{RD} to distinguish from the current
+release, \texttt{R}. As the names suggest, each provide slightly
+different configurations. Of particular interest are the images
+providing development R built with support for C/C++ address and
+undefined-behavior sanitizers, which are somewhat difficult to configure
+\citep{edd_sanitizers}.
 
@@
 (roughly every few weeks). The rest of the stack uses build triggers
-that rebuild the images either whenever \texttt{r-base} is updated, or
-the Dockerfile sources are updated on the corresponding GitHub
-repository. The only exception in this stack is the \texttt{drd} image,
-which is rebuilt each week by a \texttt{cron} trigger.
+that rebuild the images whenever \texttt{r-base} is updated or the
+Dockerfile sources are updated on the corresponding GitHub repository.
+The only exception in this stack is the \texttt{drd} image, which is
+rebuilt each week by a \texttt{cron} trigger.
 
@@
 \end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
-R-devel added side-by-side r-base (using alias \texttt{RD})\strut
+R-devel added side-by-side to r-base (using alias \texttt{RD})\strut
 \end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
@@
 \end{minipage} & \begin{minipage}[t]{0.50\columnwidth}\raggedright\strut
-Sanitizers, clang c compiler (instead of gcc)\strut
+sanitizers, clang c compiler (instead of gcc)\strut
 \end{minipage} & \begin{minipage}[t]{0.08\columnwidth}\raggedright\strut
@@
 \endhead
-\href{https://hub.docker.com/r/rocker/r-ver}{r-ver} & Version-stable
+\href{https://hub.docker.com/r/rocker/r-ver}{r-ver} & version-stable
 base R \& src build tools & 219 MB & 6,000\tabularnewline
-\href{https://hub.docker.com/r/rocker/rstudio}{rstudio} & Adds rstudio &
+\href{https://hub.docker.com/r/rocker/rstudio}{rstudio} & adds rstudio &
 334 MB & 314,000\tabularnewline
-\href{https://hub.docker.com/r/rocker/tidyverse}{tidyverse} & Adds
+\href{https://hub.docker.com/r/rocker/tidyverse}{tidyverse} & adds
 tidyverse \& devtools & 656 MB & 83,000 \footnote{This figure includes
   49,000 downloads under the earlier name \texttt{hadleyverse}.}\tabularnewline
-\href{https://hub.docker.com/r/rocker/verse}{verse} & Adds java, tex \&
+\href{https://hub.docker.com/r/rocker/verse}{verse} & adds java, tex \&
 publishing-related packages & 947 MB & 9,000\tabularnewline
-\href{https://hub.docker.com/r/rocker/geospatial}{geospatial} & Adds
+\href{https://hub.docker.com/r/rocker/geospatial}{geospatial} & adds
 geospatial libraries & 1.3 GB & 4,000\tabularnewline
@@
 \texttt{latest}, which, as the name implies, will always point to an
-image using the current R release. Thus, users wanting to create
-downstream Dockerfiles which are based on the current release at time
-(but will continue to reconstruct the same environment in the future
-after newer R versions are released) should explicitly include the
-corresponding version tag, \emph{e.g.}, \texttt{rocker/r-ver:3.4.2} at
-the time of writing, and not the \texttt{latest} tag. Users can also run
-the current development version of R using the tag \texttt{devel}, which
-is built nightly from R-devel sources from \texttt{subversion}.
+image using the current R release. Thus, users who want to create
+downstream Dockerfiles, which are based on the current release at the
+time (but will continue to reconstruct the same environment in the
+future after newer R versions are released), should explicitly include
+the corresponding version tag, \emph{e.g.}, \texttt{rocker/r-ver:3.4.2}
+at the time of writing, and not the \texttt{latest} tag. Users can also
+run the current development version of R using the tag \texttt{devel},
+which is built nightly from R-devel sources from \texttt{subversion}.
 
@@
 of writing), \texttt{rocker/verse:3.4.2}. This is accomplished using
-post-build hooks in Docker Hub, see examples at
+post-build hooks in Docker Hub---see examples at
 \url{https://github.com/rocker-org/rocker-versioned/} for details.
@@
 triggers for the numeric version tags are sent monthly. Although the
-Dockerfiles for older R versions installs an almost-identical software
+Dockerfiles for older R versions install an almost-identical software
 environment every time, the monthly rebuilding of these images on Docker
@@
 environment and Java development libraries. Previously, the Rocker
-project provided the image \texttt{hadleyverse} which was since divided
-into \texttt{tidyverse} and \texttt{verse} based on community input.
+project provided the image \texttt{hadleyverse} which has since been
+divided into \texttt{tidyverse} and \texttt{verse} based on community
+input.
 

### 2017-11-24 boettiger-eddelbuettel.Rmd
three small changes based on proofreading-checklist

@@
 
-### Community Optimized
+### Community optimized
 
@@
 repositories on GitHub attest to some of these discussions and improvements.
-In particular, input from the Docker Inc employees through the official approval process
+In particular, input from the Docker Inc.\ employees through the official approval process
 for the `r-base` image, expertise from the Debian R maintainer and other Debian developers,
@@
 Docker organization itself.  This official image is reviewed and then
-built by employees of Docker Inc. based on a Dockerfile maintained by
+built by employees of Docker Inc.\ based on a Dockerfile maintained by
 the Rocker team.  Consequently, users should refer to this image in
@@
 to access the official image.  All other images in the Rocker
-project are not individually reviewed and built by Docker Inc. and must
+project are not individually reviewed and built by Docker Inc.\ and must
 be referenced using the `rocker` namespace,

### 2017-11-24 boettiger-eddelbuettel.tex
three small changes based on proofreading-checklist

@@
 
-\subsubsection{Community Optimized}\label{community-optimized}
+\subsubsection{Community optimized}\label{community-optimized}
 
@@
 attest to some of these discussions and improvements. In particular,
-input from the Docker Inc employees through the official approval
+input from the Docker Inc.~employees through the official approval
 process for the \texttt{r-base} image, expertise from the Debian R
@@
 language by the Docker organization itself. This official image is
-reviewed and then built by employees of Docker Inc. based on a
+reviewed and then built by employees of Docker Inc.~based on a
 Dockerfile maintained by the Rocker team. Consequently, users should
@@
 official image. All other images in the Rocker project are not
-individually reviewed and built by Docker Inc. and must be referenced
+individually reviewed and built by Docker Inc.~and must be referenced
 using the \texttt{rocker} namespace, \emph{e.g.},

### 2017-11-24 boettiger-eddelbuettel.Rmd
whitespace mostly

@@
 and an image for reproducing a data analysis will probably select a stable version tag
-in addition to an appropriate base library, _e.g._,: `FROM rocker/tidyverse:3.4.1`
+in addition to an appropriate base library, _e.g._,: `FROM rocker/tidyverse:3.4.1`.
 Users can easily add additional software to any running Rocker image using
@@
 are sent to `unstable` during the normal course of Debian development.
-This can occasionally break an previously-working installation command
+This can occasionally break a previously-working installation command
 in a Dockerfile until the maintainer redirects the package manager to

### 2017-11-24 boettiger-eddelbuettel.tex
whitespace mostly

@@
 ``run'' to create a container whenever it is needed. A ``Dockerfile'' is
-a recipe, the source-code, to create a Docker image.\\
-Pre-built Docker images are publicly available through Docker Hub, which
-plays a role for central distribution similar to CRAN in our analogy.
-Development and contributions to the Rocker project focus on the
-construction, organization and maintenance of these Dockerfiles.
+a recipe, the source-code, to create a Docker image. Pre-built Docker
+images are publicly available through Docker Hub, which plays a role for
+central distribution similar to CRAN in our analogy. Development and
+contributions to the Rocker project focus on the construction,
+organization and maintenance of these Dockerfiles.
 
@@
 these users root privileges in the container still leaves them sandboxed
-from the host container.\\
-Sandboxing also serves an important function in reproducible research by
-making it easier to test a specified environment in isolation from the
-host machine. Unlike traditional virtual machines, containers do not
-impose a large footprint of reserved resources as a typical host can
-easily support 100s of containers \citep{what-docker}.
+from the host container. Sandboxing also serves an important function in
+reproducible research by making it easier to test a specified
+environment in isolation from the host machine. Unlike traditional
+virtual machines, containers do not impose a large footprint of reserved
+resources as a typical host can easily support 100s of containers
+\citep{what-docker}.
 

### 2017-11-25 boettiger-eddelbuettel.Rmd
a few more tweaks

@@
 files with familiar, native tools while performing computation through a reproducible,
-containerized environment \citep{Boettiger2015}. Being
+containerized environhment \citep{Boettiger2015}. Being
 able to test code in a predictable, pre-configured R environment on a
@@
 
-More details can be found in the Singularity documentation.
+\noindent More details can be found in the Singularity documentation.
 
@@
 being used as base Docker images in the NSF sponsored Whole Tale project for
-reproducible computing \citep{wholetale}, and are heavily used by the R-Hub
+reproducible computing \citep{wholetale}, and are heavily used by the \CRANpkg{rhub}
 project in automated package testing \citep{rhub}.
@@
 All Rocker images are based on the Debian Linux distribution.
-The Debian platform provides a small base image, the well-known `apt`
+It provides a small base image, the well-known `apt`
 package management system, and a rich ecosystem of software libraries,
@@
 to install R packages is fixed to a snapshot of CRAN corresponding to
-the last date for which that version of R was current (e.g. `3.4.2` was released
+the last date for which that version of R was current (_e.g._, `3.4.2` was released
 on 2017-09-28, thus `3.4.1` is pinned to the MRAN snapshot for that date). These
@@
 stack: _e.g._, `rocker/tidyverse:devel` will provide the currently-released
-versions of the R packages in the `tidyverse` \citep{tidyverse}
+versions of the R packages in the \CRANpkg{tidyverse} \citep{tidyverse}
 installed on the nightly build of R-devel.  Developers building packages
@@
 position of the tag is identical to asking for the most recent version:
-i.e. `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
+_i.e._, `rocker/verse:3.3` is the same as `rocker/verse:3.3.3`, and 
 `rocker/verse:3` is (at the time of writing), `rocker/verse:3.4.2`.

### 2017-11-25 boettiger-eddelbuettel.tex
a few more tweaks

@@
 with familiar, native tools while performing computation through a
-reproducible, containerized environment \citep{Boettiger2015}. Being
+reproducible, containerized environhment \citep{Boettiger2015}. Being
 able to test code in a predictable, pre-configured R environment on a
@@
 themselves provide a simple example of this, (as described in the
-following section).\\
-A user begins by selecting an appropriate base image for their needs: if
-the RStudio\textsuperscript{\textregistered} interface is desired, a
-user might start with \texttt{FROM\ rocker/rstudio}; an image for
-testing an R package with compiled code might use
-\texttt{FROM\ rocker/r-devel-san}, and an image for reproducing a data
-analysis will probably select a stable version tag in addition to an
-appropriate base library, \emph{e.g.},:
-\texttt{FROM\ rocker/tidyverse:3.4.1} Users can easily add additional
+following section). A user begins by selecting an appropriate base image
+for their needs: if the RStudio\textsuperscript{\textregistered}
+interface is desired, a user might start with
+\texttt{FROM\ rocker/rstudio}; an image for testing an R package with
+compiled code might use \texttt{FROM\ rocker/r-devel-san}, and an image
+for reproducing a data analysis will probably select a stable version
+tag in addition to an appropriate base library, \emph{e.g.},:
+\texttt{FROM\ rocker/tidyverse:3.4.1}. Users can easily add additional
 software to any running Rocker image using the standard R and Debian
@@
 Tale project for reproducible computing \citep{wholetale}, and are
-heavily used by the R-Hub project in automated package testing
+heavily used by the \CRANpkg{rhub} project in automated package testing
 \citep{rhub}.
@@
 
-All Rocker images are based on the Debian Linux distribution. The Debian
-platform provides a small base image, the well-known \texttt{apt}
-package management system, and a rich ecosystem of software libraries,
-making it the base image of choice for Docker images, including many of
-the ``official'' images maintained by Docker's own development team. The
+All Rocker images are based on the Debian Linux distribution. It
+provides a small base image, the well-known \texttt{apt} package
+management system, and a rich ecosystem of software libraries, making it
+the base image of choice for Docker images, including many of the
+``official'' images maintained by Docker's own development team. The
 Debian platform is also perhaps the best-supported Linux platform within
@@
 into \texttt{testing}. New versions are sent to \texttt{unstable} during
-the normal course of Debian development. This can occasionally break an
+the normal course of Debian development. This can occasionally break a
 previously-working installation command in a Dockerfile until the
@@
 corresponding to the last date for which that version of R was current
-(e.g. \texttt{3.4.2} was released on 2017-09-28, thus \texttt{3.4.1} is
-pinned to the MRAN snapshot for that date). These snapshots are provided
-by the MRAN archive created by Revolution Analytics (now part of
-Microsoft). It archives daily snapshots of all of CRAN from which a user
-can install packages with the usual \texttt{install.packages()} function
-\citep{MRAN}. Users can always override this default by passing any
-current CRAN repository explicitly. Unlike CRAN, Bioconductor only
-updates its repositories through bi-annual releases aligned to R's
-spring release schedule. Thus, Bioconductor packages can be installed in
-the usual way using \texttt{bioclite}, which automatically selects the
-Bioconductor release corresponding to the version of R in use.
+(\emph{e.g.}, \texttt{3.4.2} was released on 2017-09-28, thus
+\texttt{3.4.1} is pinned to the MRAN snapshot for that date). These
+snapshots are provided by the MRAN archive created by Revolution
+Analytics (now part of Microsoft). It archives daily snapshots of all of
+CRAN from which a user can install packages with the usual
+\texttt{install.packages()} function \citep{MRAN}. Users can always
+override this default by passing any current CRAN repository explicitly.
+Unlike CRAN, Bioconductor only updates its repositories through
+bi-annual releases aligned to R's spring release schedule. Thus,
+Bioconductor packages can be installed in the usual way using
+\texttt{bioclite}, which automatically selects the Bioconductor release
+corresponding to the version of R in use.
 
@@
 stack: \emph{e.g.}, \texttt{rocker/tidyverse:devel} will provide the
-currently-released versions of the R packages in the \texttt{tidyverse}
+currently-released versions of the R packages in the \CRANpkg{tidyverse}
 \citep{tidyverse} installed on the nightly build of R-devel. Developers
@@
 second or third position of the tag is identical to asking for the most
-recent version: i.e. \texttt{rocker/verse:3.3} is the same as
+recent version: \emph{i.e.}, \texttt{rocker/verse:3.3} is the same as
 \texttt{rocker/verse:3.3.3}, and \texttt{rocker/verse:3} is (at the time

### 2017-11-25 boettiger-eddelbuettel.Rmd
added ORCiD as suggested in proofreading-checklist

@@
     - 130 Mulford Hall Berkeley, CA 94720-3114, USA
+    - ORCiD \href{https://orcid.org/0000-0002-1642-628X}{0000-0002-1642-628X}
     email:  cboettig@berkeley.edu
@@
     - Chicago, IL, USA
+    - ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}
     email:  edd@debian.org

### 2017-11-25 boettiger-eddelbuettel.tex
added ORCiD as suggested in proofreading-checklist

@@
 UC Berkeley\\
-ESPM Department, University of California,\\ 130 Mulford Hall Berkeley, CA 94720-3114, USA\\
+ESPM Department, University of California,\\ 130 Mulford Hall Berkeley, CA 94720-3114, USA\\ ORCiD \href{https://orcid.org/0000-0002-1642-628X}{0000-0002-1642-628X}\\
 }
@@
 Debian and R Projects; Ketchum Trading\\
-Chicago, IL, USA\\
+Chicago, IL, USA\\ ORCiD \href{https://orcid.org/0000-0001-6419-907X}{0000-0001-6419-907X}\\
 }

### 2017-11-28 boettiger-eddelbuettel.Rmd
one more \CRANpkg{}

@@
 from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
-or adopt an alternative approach such as `packrat` \citep{packrat}.
+or adopt an alternative approach such as \CRANpkg{packrat} \citep{packrat}.
 A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \cite{Boettiger2015}.
